\chapter{ Optimising machine learners to learn on their own}
\label{ch:subjects}


```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, tidy=TRUE, fig.height=8, 
                      echo=FALSE,  warning=FALSE, message=FALSE, dev='pdf')
library(RSQLite)
con <- dbConnect(RSQLite::SQLite(),'../ml_lit/all_refs.sqlite3')
res = dbGetQuery(con, statement ="select * from basic_refs limit 10;")
library(ggplot2)
library(stringr)
library(xtable)
options(xtable.comment = FALSE)
```

> If a proposition, a sentence, a group of signs can be called 'statement' , it is not therefore because, one day, someone happened to speak them or put them into some concrete form of writing; it is because the position of the subject can be assigned [@Foucault_1972,  95] \index{statement!position of subject}

> 'generalization error is what we care about' [@Ng_2008f]

> *Predict if an online bid is made by a machine or a human*, 'Facebook Recruiting IV: Human or Robot?' [@Kaggle_2015e]

In early 2002, while carrying out an ethnographic study of 'extreme programming,' a software development methodology popular at that time [@Mackenzie_2004], I spent several months visiting a company in Manchester developing software for call centres. The purpose of the software was to manage 'knowledge' in call centres such that any query from a caller would be readily answered by call centre staff who would query a knowledge management system. This system was marketed on the promise of machine learning. It contained  an artificial neural network that learned to match queries and responses over time. \index{machine learner!neural net} A taciturn neural network expert, Vlad, sat in a different part of the room from the developers working on the databases and the web interfaces of the knowledge management system. Vlad's work on the neural network was at the core of the knowledge management system yet outside the orbit of the software development team and its agile software development processes. They generally regarded Vlad and the neural net as an esoteric,  temperamental yet powerful component, a hidden node we might say, of the knowledge management system. As we have already seen with `kittydar` \index{kittydar}, today the situation of machine learners has changed. They are no longer exotic or specialized, but both banal and occasionally spectacular. Hilary Mason, who was Chief Scientist at bitly.com (an online service that shortens URLs), outlined a machine learning subject position at a London conference in 2012 called 'Bacon: Things Developers Love':

>You have all of these things that are different – engineering, infrastructure, mathematics and computer science, curiosity and an understanding of human behaviour – that is something that usually falls under the social sciences. At the intersection of all these things are wonderful people. But we're starting to develop something new, and that is - not that all of these things have not been done for a very long time - but we are only just now building systems where people, individual people, have all of these tools in one package, in one mind. And there are lots of reasons this is happening now. But its a pretty exciting time to be in any of these things. (Hilary Mason, Chief Scientist, bitly.com) [@Bacon_2012] 

These 'things that are different,'  if they amount to a _statement_ in Michel Foucault's sense of that term, together assign subject positions \index{statement} \index{Foucault, Michel}. In what ways? In front of an audience of several hundred software developers, Mason describes shifts in the work of programming associated with the growth of large amounts of data associated with 'human behaviour.' At the centre of this shift stand 'wonderful people' who combine practices and knowledges of communication infrastructure, technology, statistics, and 'human behaviour' through curiosity and technical skills.  Mason was, in effect, telling her audience of software developers who they should become and at the same time pointing to the some of the expansive changes occurring around them. The title of her talk was 'machine learning for hackers', and her audience were those hackers or programmers whose coding and programming attention may have been previously trained on web interfaces or database queries, but was now drawn towards machine learning. A change in programming practice and a shift towards machine learning was, she implied,  the key to programmers becoming the wonderful people, agents of their own time, capable of doing what is only now just possible because it is all together in 'one package, one mind.' Note that concatenation of 'one package, one mind' does not definitively allocate agency to people or things. Here, I would tentatively suggest, Mason adumbrates the outline of an agent of anticipation, someone or something at the intersection of network infrastructure, mathematics and human behaviour.[^7.1] Mason, one of _Fortune_ magazines 'Top 40 under 40' business leaders to watch [@CNN_2011] and also featured in _Glamour_, a teenage fashion magazine [@Mason_2012], might personify such a wonderful person.\index{Mason, Hilary} She is not a lone example.[^7.2] Equally so, she might _a-personify_ the intersection. 

'It is the privileged machine in this context that creates its marginalized human others' writes Lucy Suchman in her account of  the encounters that 'effect "persons" and "machines" as distinct entities' [@Suchman_2006, 269] \index{Suchman, Lucy}. While Mason and other relatively well-known human machine learners are not exactly marginalized (just the opposite, they achieve minor celebrity status in some cases), Suchman recommends 'recognition of the particularities of bodies and artifacts, of the cultural-historical practices through which human-machine differences are (re-)iteratively drawn, and of the possibilities for and politics of redistribution across the human machine boundary' (285). The intersections that machine learners currently occupy are heavily re-distributional. In almost every instance, machine learners claim to do something that humans alone, no matter how expert, could not. (We need only think of the demonstration of the radio-controlled helicopter flying upside down that Andrew Ng shows in his introduction of machine learning lecture [@Ng_2008] \index{Ng, Andrew}.) Could the 'wonderful people' that Mason describes also be seen as marginalized human others? Does the re-distribution of engineering, mathematics, curiosity, infrastructure and 'something that usually falls under the social sciences' (but perhaps no longer does so?) both energise subjects ('its a pretty exciting time to be in any of these things') and assign them a marginal albeit still pivotal position?

Machine learner subject positions are the topic of this chapter. I draw on artificial neural networks, or neural nets, in their various forms ranging from the perceptron, through the multilayer perceptron (MLP), the convolutional neural nets (CNN), recurrent neural nets (RNN) and deep belief networks or deep neural networks of many recent deep learning projects (particularly in machine learning  competitions, as discussed below [@Dahl_2013]). \index{neural net|see{machine learner!neural net}} in exploring the re-drawing of subject-machine positions.  \index{diagram!machine learner as} Neural nets propagate between infrastructures, engineering and human behaviour (as Mason puts it), re-drawing human-machine differences. A straightforward example of this re-drawing appears in neural net publications. Geoffrey Hinton, Simon Osindero and Yee-Why Teh \index{Hinton, Geoffrey} writing in _Neural Computation_ in 2006 described a 'fast learning algorithm for deep belief nets' [@Hinton_2006a]. Their description, whilst mostly couched in terms of conditional distributions, parameters, and generative models \index{model!generative} also contains a section entitled 'Looking into the Mind of a Neural Network' (1545-1546). In that section, they describe how they used their deep belief network to _generate_ rather than classify images.[^7.01] In the process they were able to see what the 'associative memory has in mind' (1545). The term 'mind' it turns out 'is not intended to be metaphorical' (1546) because the neural net in question has a distributed memory of the digits it has seen. Put slightly more formally, 'the network has a full generative model, so it is easy to look into its mind - we simply generate an image from its high-level representations' (1529).   The conjunction or intersection of 'mind' and generative model draws a new diagram of artificial intelligence (which has typically relied on rule-based or symbolic reasoning), but the appearance of 'mind' in the form of a generative model (see chapter \ref{ch:probability}) covers over other layers of positioning and subjectification. These layers might be more interesting to explore archaeologically (for instance, in the work done on specific artifacts such as images) than in terms of the general existential threat of artificial intelligence. Neural nets re-iterate human machine boundaries through a combination of feeding-forward of potentials and propagating backwards of differences specifically concerned with images. Similarly, the practice of machine learning assigns subject positions in a backward and forward movement that propagates potentialising optimism even as it undercuts the very differences that give rise to that optimism. \index{back-propagation!see{algorithm!back-propagation}}

## Propagating neural nets in machine learning 

```{r nn_de, cache=TRUE}

library(stringr)
library(RSQLite)
library(ggplot2)
con <- dbConnect(RSQLite::SQLite(),'../ml_lit/all_refs.sqlite3')
res = dbGetQuery(con, statement = "select DE, WC, SC, TI, PY from basic_refs")
res$WC = tolower(res$WC)
res$DE = tolower(res$DE)
stat =  grep(res$WC, pattern = 'statistic', value=FALSE)
compsci =  grep(res$WC, pattern = 'computer', value=FALSE)
eng =  grep(res$WC, pattern = 'engineer', value=FALSE)
interdisc = intersect(stat, compsci)

norm_terms <- function(x){
    y = str_replace(x, pattern='artificial ', '')
    z = str_replace(y, pattern='svm', 'support vector machine')
    return(z)
}

de_plural <- function(x){
    z = str_replace(x, pattern = '(.+)s$', replacement = '\\1')
    z = norm_terms(z)
    return(z)
}

de_split <- function(x){
    y = unlist(str_split(x, '; '))
    return(y)
}

tablify <- function(x) {
    z  = sort(table(de_plural(de_split(x))), decreasing=TRUE)
    df = data.frame(count = z, discipline = names(z))
    return(df)
} 

stat_df = (head(tablify(res$DE[stat]), 100))
inter_df = (head(tablify(res$DE[interdisc]), 100))
compsci_df = (head(tablify(res$DE[compsci]), 100))
eng_df = (head(tablify(res$DE[eng]), 100))

df = cbind(stat_df, inter_df, compsci_df, eng_df)
dft = xtable(df[1:20,],  label='tab:disc_tech', caption='Techniques and concepts most frequently mentioned in machine learning publication keywords, 1955-2015')
print(dft, type='latex', include.rownames = FALSE)
```

Almost every machine learning class, textbook, demonstration, and in recent years, in machine learning competitions at some point turns to neural nets. They display, however, some instability in the research literature. Table \ref{tab:disc_tech} shows the most frequent keywords for technical publications across the three main disciplinary domains inhabited by machine learners. While neural nets rank very high in computer science and engineering disciplines (appearing just after support vector machines), they do not appear in the statistics literature until `r which(stat_df$discipline == 'neural network')` in the rankings. The prominence of neural nets on the engineering side of machine learning suggests they perform specific enunciative functions. \index{function!enunciative!of neural net}   Neural nets are often described from a deeply split perspective that in some points turns towards human subjects, or at least, the brains of human subjects, and in other ways towards the ongoing expansion of computational platforms. In some ways, they renew long-standing cybernetic hopes of brains and cognition as models of computational intelligence and agency.  Although they stem from a biological inspiration (dating at least back to the work on McCulloch and Pitts in the 1940s [@Halpern_2015; @Wilson_2010]), they gain traction first in the 1980s and then again from mid-2000s onwards, as ways of dealing with changing computational infrastructures, and the difficulties of capitalising on infrastructure that is powerful yet difficult to manage. Neural nets almost constantly oscillate between brain and information infrastructure. In the course of fifty years, their triple re-invention -- from perceptron via neural net to deep belief net -- re-distributes subject positions amidst infrastructural re-configurations and vectorisation. \index{infrastructure!reconfiguration of} \index{vectorisation!of infrastructure} For instance, David Ackley, Geoffrey Hinton (an important figure in the inception of neural nets during the 1980s and in the revival of neural net in the form of deep learning in the last decade), and Terrence Sejnowski wrote in the early 1980s: \index{Hinton, Geoffrey}

> Evidence about the architecture of the brain and the potential of the new VLSI technology have led to a resurgence of interest in “connectionist” systems ...  that store their long-term knowledge as the strengths of the connections between simple neuron-like processing elements. These networks are clearly suited to tasks like vision that can be performed efficiently in parallel networks which have physical connections in just the places where processes need to communicate. ... The more difficult problem is to discover parallel organizations that do not require so much problem-dependent information to be built into the architecture of the network. Ideally, such a system would adapt a given structure of processors and communication paths to whatever problem it was faced with [@Ackley_1985, 147-148].

Alignments and diagrammatic overlaps between brain and 'new VSLI [Very Large Scale Integrated] technology' -- semiconductor chips -- architectures sought to map the plasticity of neuronal networks onto the parallel distributed processing enabled by very densely packed semiconductor circuits. \index{diagram!overlay} The problem here was how to organize these connections without having to hardwire domain specificity into 'the architecture of the network.' How could the architectures adapt to the problem in hand?

We saw in chapter \ref{ch:diagram} that the psychologist Frank Rosenblatt's perceptron [@Rosenblatt_1958] first implemented McCulloch and Pitts' cybernetic vision of neurones as models of computation [@Edwards_1996] \index{machine learner!perceptron}. While the perceptron did not weather the criticism of artificial intelligence experts such as Marvin Minsky (Minsky famously showed that a perceptron cannot learn the logical exclusive OR or `XOR` function; [@Minsky_1969]), cognitive psychologists such as David Rumelhart, Geoffrey Hinton and Ronald Williams persisted with perceptrons, seeking to generalize their operations by connecting them together in networks (also known as multilayer perceptrons). In the mid-1980s, they developed the back-propagation algorithm \index{algorithm!back-propagation} [@Rumelhart_1985; @Hinton_1989], a way of adjusting the connections between nodes (neurones) in the network in response to features in the data (see Figure \ref{fig:rumelhart1985}). The back-propagation algorithm directly addressed the problem of modifying parallel network organizations without reliance on problem-dependent architectures, and in without having to program them in. \index{programmability!problem of!neural net as solution to} Effectively, an architecture of generalization was implemented. While cognition, and the idea that machines would be cognitive (rather than say, mechanical, calculative, or even algorithmic) constantly organised research work in artificial intelligence for several decades, the development of the back-propagation algorithm as a way for a set of connected computational nodes to learn also had strong infrastructural resonances. These resonances became much more visible from around 2006 when 'deep belief nets' appeared as a way of training many-layered neural nets [@Hinton_2006b]. These resonances continue to echo today and indeed attract much attention.[^7.005] Like the advent of VSLI in the early 1980s, the vast concentrations of processing units in contemporary data centres (hundreds of thousands of cores as we saw in the case of Google Compute in the previous chapter \ref{ch:genomic}) and even in the graphics cards developed for high-end gaming  and video rendering (GPUs for PC gaming now typically have a thousand and sometimes several thousand cores) pose the problem of organizing infrastructure so that processes can communicate with each other.Machine learners may well become more important as loose or mutable infrastructural arrangements than as epistemic instruments. 


\begin{figure}
  \centering
      \includegraphics[width=0.9\textwidth]{figure/hinton.pdf}
        \caption{An early publication of the back-propagation algorithm: Rumelhart, Hinton and William's 1985 paper [@Rumelhart\_1985]}
  \label{fig:rumelhart1985}
\end{figure}

The back-propagation or 'backprop' algorithm will return below, but for the moment, this oscillation between cognition and infrastructures, between people and machines, itself suggests another way of thinking about how 'long-term knowledge' takes shape today. At the same time as infrastructural reorganization takes place around learning, and around the production of statements by machine learners, both human and non-human machine learners are assigned new positions. These positions are dispersed, hierarchical and distributed. The subject position in machine learning is a highly relational one rather than a single concentrated form of expertise (as we might find in a clinical oncologist, biostatistician or geologist). Because machine learners construct vectorized epistopics, partial observers traversing vector spaces, inverse probabilisations, pattern dispersions and regularizing hyperobjects, what counts as agency, skill, action, experience and learning shifts constantly. \index{common vector space!epistopics} It is intimately bound and connected to transforms in infrastructure, variations in referentiality (such as we have seen in the construction of the common vector space), and competing forms of authority (as we have seen in the accumulations of different techniques). As Suchman suggests, examining privileged machines is a way to pay attention to variously marginalized human others.

The ranking of keywords in table \ref{tab:tech_disc} suggests that some machine learners attribute a more privileged and constitutive function to neural nets. Neural nets synchronically spread into many difference disciplines: cognitive science, computer science, linguistics, adaptive control engineering, psychology, finance, operations research, etc., and particularly statistics and computer science during the 1980-1990s. This  dendritic growth did not just popularise machine learning. It brought engineering and statistics together more strongly. Ethem Alpaydin, a computer scientist, writes:

>Perhaps the most important contribution of research on neural networks is this synergy that bridged various disciplines, especially statistics and engineering. It is thanks to this that the field of machine learning is now well established [@Alpaydin_2010, 274]. 

The forms of this field-making bridging are various.[^7.0010]   The primary space of coexistence of different disciplines around machine learning has perhaps been competitions during the 1990s that pitted neural nets against other machine learners in classifying handwritten numerals such as zipcodes on envelopes, and in particular one set of handwritten digits that remain in constant use, the MNIST (Modified National Institute of Standards) dataset [@LeCun_2012]. \index{dataset!MNIST} The many competitions focused on the MNIST dataset are, I suggest, a form of demonstration and testing of machines and people that propagate machine-human differences in machine learning. It is no accident that Hastie and co-authors in _Elements of Statistical Learning_ devote a lengthy section to the analysis of these handwritten digit recognition competitions that began in the early 1990s and continue today. Like Alpaydin, they affirm the coordinating effect of these competitions on the development of machine learning:

>This problem captured the attention of the machine learning and neural network community for many years, and has remained a benchmark problem in the field [@Hastie_2009, 404].

The handwritten digits used in these competitions, particularly the Neural Information Processing System workshops and KDD Cup (Knowledge Discovery and Data Mining) [@KDD_2013], all come from the MNIST dataset  and during the 1990s, much effort focused on crafting neural nets to recognise these 60,000 or so handwritten digits. As Hastie and co-authors observe, 'at this point the digit recognition datasets became test beds for every new learning procedure, and researchers worked hard to drive down the error rates' [@Hastie_2009, 408-409].

Despite this binding function, neural networks have had a somewhat problematic position in to machine learning. Even in relation to the paradigmatic handwritten digit recognition problem, neural nets struggled to gain purchase.  On the one hand, their analogies and figurations as sophisticated neuronal-style models suggested cognitive capacities surpassing the more geometrical, algebraic and statistically grounded machine learners such as linear discriminant analysis, logistic regression, or decision trees. On the other hand, the density and complexity of their architecture made them difficult to train. Neural nets could easily overfit \index{overfitting} the data. As _Elements of Statistical Learning_ puts it, it required 'pioneering efforts to handcraft the neural network to overcome some these deficiencies..., which ultimately led to the state of the art in neural network performance' [@Hastie_2009, 404]. It is rare to find the word 'handcraft' in machine learning literature. The operational premise of most machine learners is that machine learning works without handcrafting, or that is automates what had previously been programmed by hand. \index{programmability!problem of} Somewhat loopily, the competition to automate recognition of handwritten digits, the traces that epitomise movements of hands,  entailed much handcrafting and recognition of variations in performances of the machine.

The shifting fortunes of neural nets are frequently discussed in contrasting terms by machine learners themselves.[^7.0011] In recent years, they often  point to some kind of transformation:

>Neural networks went out of fashion for a while in the 90s - 2005 because they are hard to train and other techniques like SVMs beat them on some problems. Now people have figured out better methods for training deep neural networks, requiring far fewer problem-specific tweaks. You can use the same pretraining whether you want a neural network to identify whose handwriting it is or if you want to decipher the handwriting, and the same pretraining methods work on very different problems. Neural networks are back in fashion and have been outperforming other methods, and not just in contests [@Zare_2012].

The somewhat vacillating presence of neural nets in the machine learning literature itself finds parallels in the fortunes of individual machine learners. Yann Le Cun's work on optical character recognition during 1980-1990s is said to have discovered the back-propagation algorithm at the same time as Rumelhart, Hinton and Williams [@Rumelhart_1986]. His implementations in `LeNet` won many research machine learning  competitions during the 1990s. In 2007, Andrew Ng could casually observe that neural nets _were_ the best, but in  2014, Le Cun find himself working on machine learning at Facebook [@Gomes_2014]. \index{Le Cun, Yann} Similarly, the cognitive psychologist Geoffrey Hinton's involvement in the early 1980s work on connectionist learning procedures in neural nets and subsequently on 'deep learning nets' [@Hinton_2006] delivers him to Google in 2013. \index{Hinton, Geoffrey} Trajectories between academic research and industry are not unusual for machine learners. Many of the techniques in machine learning have been incorporated into companies later acquired by other larger companies. Even if there is no spin-off company to be acquired, machine learners themselves have been assigned key positions in many industry settings. Corinna Cortes, co-inventor with Vladimir Vapnik of the  support vector machine, heads research at Google New York \index{Cortes, Corinna}. In 2011, Ng led a neural net-based project at Google that had, among other things, detected cats  in millions of hours of Youtube videos.[^7.3] Ng himself in 2014 began work as chief scientist for the Chinese search engine, Baidu leading a team of AI researchers specializing in 'deep learning,' the contemporary incarnation of neural nets [@Hof_2014] \index{deep learning|see{machine learner!deep learning}}. \index{Ng, Andrew} In recent years, (2012-2015), work on neural nets has again intensified, most prominently in association with social media platforms, but also in the increasingly common speech and face recognition systems found in everyday services and devices. Many of these neural nets are like `kittydar` \index{kittydar}, but implemented on a much larger and more distributed scale (for instance, in classifying videos on Youtube). In contemporary machine learning competitions, as we will see, neural nets again surface as intersectional machines, re-distributing differences between humans and machines. 

## A privileged machine and its diagrammatic forms

What accounts for the somewhat uneven fortunes of the neural net amongst machine learners? The unevenness of their performance, from limited curiosity in the late 1960s to best performer in the machine learning image classification competitions of the 1990s, from second best competitor in late 1990s to the spectacular promise of deep belief networks in 2012, suggests that some powerful dynamics or becomings are in play around them. These dynamics are not easily understood in terms of celebrity machine learners (human and non-human) suddenly rising to prominent or privileged positions in the research departments of social media platforms.[^7.7]  Nor does it make sense to attribute the rising fortunes of the neural net to the algorithms themselves, as if some decisive advance occurred in algorithms. The algorithms used in neural net have not, as we will see, been radically transformed in their core operations since the 1980s, and even then, the algorithms themselves (principally gradient descent \index{gradient descent}) were not new. There have been important changes in scale (similar to those described in the previous chapter in the case of the `RF-ACE` algorithm and Google Compute), but as is often the case in machine learning, their re-invention occurs through proliferation, changes in scale, re-distributions of knowledge and infrastructure and specific optimisations. While machine learners in their machine form can be assigned a privileged position in the transformations of knowledge and action today, human machine learners are not exactly marginalized, at least in high profile cases such as Ng, Le Cun, Hinton and others. Rather, the scale of machine learning seems to be changing both for the algorithms and for the human computer scientists, programmers and engineers. 

What accounts for this acceleration and slowing-down, the intensified interactions and abandonments of neural nets over the last three decades? Despite their apparent differences in origin, neural net share much with other machine learners. The language of brain, neurones and cognition associated with neural net covers over their much more familiar vector space and function-finding optimisations they rely on. Diagrammatic groupings and lines of movement operate in neural nets to expand their architecture in alignment with a series of existing machine learners.   'The central idea,' write Hastie and co-authors, 'is to extract linear combinations of the inputs as derived features, and then model the target as a nonlinear function of these features. The result is a powerful learning method, with widespread applications in many fields' [@Hastie_2009, 389]. The 'central idea' can be seen in the algebraic expressions that Hastie and co-authors provide for the basic neural net model:

\begin {equation}
\label {eq:nn}
\begin {split}
Z_m = \sigma(\alpha_0m + \alpha_m^TX) m = 1, ...,M\\
T_k = \beta_0k + \beta_k^TZ, k = 1, ..., K,\\
f_k (X) = g_k (T), k = 1, ..., K,
\end {split}
\end {equation}

>where $Z = (Z_1 , Z_2, ..., Z_M )$, and $T = (T_1 , T_2 ,..., T_K )$. The activation function $\sigma(v)$ is usually chosen to be the sigmoid $\sigma(v) = 1/(1 + e - v )$

>[@Hastie_2009, 392]

Equation \ref{eq:nn} is a diagram with some familiar elements as well as some novelty. Some of the diagrammatic operation of the neural net is already familiar from the linear models. The neural networks traverse data in a vector space denoted by $X$. That is common to nearly all machine learners.  They make use of the non-linear sigmoid function \index{function!sigmoid} that lies at the heart of one of the main linear classifiers used in machine learning, logistic regression \index{machine learner!logistic regression}. Their training and learning processes have come to rely on the same kinds of cost, loss or error functions \index{cost function} we have seen in other  machine learners.

Their apparently increasing power to learn (to see, to find, to classify, to rank, to predict) owes much to the diagrammatic movements that recombine operations of past machine learners in new intersections. These movements appear in the equations.  Equation \ref{eq:nn} has three lines rather than one, and this layering and its diagonal patterns of indexical referencing running between subscripts distinguishes neural nets from the linear  models it assembles. \index{diagram!equation as} 

\begin {equation}
\label {eq:linear_model}
\hat{Y} = \hat{\beta_0}  + \sum^p_{j=1} X_j \hat{\beta_j}
\end {equation}
 
Whereas the standard linear model \index{linear model} shown in Equation \ref{eq:linear_model} indexes a single common vector space $X_j$ and approximates a single function $\hat{Y}$ by searching for the values of the parameters $\beta_j$ that best incline a plane through the given data, the three lines shown in equation \ref{eq:nn} are woven through each other much more consecutively. Much hinges on the unobstrusive sigmoid function operator written as $\sigma$: 'a neural network can be thought of as a nonlinear generalization of the linear model, both for regression and classification. By introducing the nonlinear transformation $\sigma$, it greatly enlarges the class of linear models' [@Hastie_2009, 394]. $\sigma$, it seems, allows neural nets to generalize beyond the linear model.[^7.32]

The common diagrammatic operations of neural nets and other supervised machine learners immediately appears in almost any actual example of a neural net. In the code vignette shown below, the data is a spreadsheet of information about passengers of the Titanic. The `titanic` dataset, like `iris` or `boston` is often used in contemporary machine learning pedagogy. \index{dataset!Titanic}  It is for instance, the main training dataset used by (kaggle.com)[http://kaggle.com], an online machine learning competition site I will discuss below \index{Kaggle.com}. The first few lines of the `R` code load the dataset and transform it into vector space. For instance, variables such as `sex` that take values such as `male` and `female` become vectors of `1` and `0` in a new variable `sexmale`. 

```{r neural_net1, echo=TRUE, message=FALSE, warning=FALSE, cache=TRUE, fig.cap=''}
library(neuralnet)
titanic = read.csv('data/titanic3.csv')
titanic_transformed = as.data.frame(model.matrix(~survived + age+ pclass + fare+ sibsp + sex
+ parch + embarked, titanic))
head(titanic_transformed)
train_index = sample.int(nrow(titanic)/2)
titanic_train = titanic_transformed[train_index,]
titanic_net = neuralnet(survived ~ age +pclass + fare + sexmale + sibsp + parch + embarkedC + embarkedQ + embarkedS, data=titanic_train, err.fct='ce', linear.output=FALSE, rep=5, hidden=3, stepmax=1e6) 
titanic_test = titanic_transformed[-train_index,]
test_error = round(sum( 0.5 < compute(titanic_net, titanic_test[,-c(1,2)])$net.result)/sum(titanic_test$survived), 2)
```

The line of the code that constructs a neural net using the `neuralnet` library [@Fritsch_2012], and the description of the classifier here is a familiar one. Despite its biological inspiration, the `R` formula for the neural net looks very similar to other machine learners such as logistic regression. It models whether someone `survived` the wreck of the Titanic in terms of their age, class of fare (`pclass`), sex, number of siblings/spouse (`sibsp`), number of parents/children (`parch`) and port of departure:

`survived ~ age +pclass + fare + sexmale + sibsp + parch + embarkedC + embarkedQ + embarkedS`

`R` model formula express the response or target variable `survived` as a combination of other variables. In this case, the plus sign `+` indicates that the combination is linear or additive. As Hastie puts, 'the central idea is to extract linear combinations of the inputs' or predictor variables. If this model formula looks so similar to other machine learning techniques we have been discussing, what do neural networks add? Why did and do so many people turn to them?

In the code vignette above, the expression `hidden = 5` points to the distinctive architecture of these models, an architecture that does not appear in the model formula in the `R` code but does, as we have already seen,  operate in the lines of Equation \ref{eq:nn}. The hidden units are indexed in the first line of the model in the variables $Z_m$.  These 'hidden units' are key to neural net since they construct the 'derived features' that the model learns from the input data $X$. As Rumelhart, Winton and Williams announce the algorithm in a letter to _Nature_ in 1986 entitled 'Learning representations by back-propagating errors' \index{algorithm!back-propagation}:

> We describe a new learning procedure, back-propagation, for networks of neurone-like units. The procedure repeatedly adjusts the weights of the connections in the network so as to minimize a measure of the difference between the actual output vector of the net and the desired output vector. As a result of the weight adjustments, internal ‘hidden’ units which are not part of the input or output come to represent important features of the task domain, and the regularities in the task are captured by the interactions of these units. The ability to create useful new features distinguishes back-propagation from earlier, simpler methods such as the perceptron-convergence procedure [@Rumelhart_1986, 533].

Again, despite the persistent reference to biology, the description of the 'new learning procedure' starts to sound more like machine learning. There is talk of minimizing a measure of difference between actual and desire output vectors (optimizing through a cost or loss function), as well as mention of 'features' and 'weights' (usually a synonym for model parameters: 'the neural network model has unknown parameters, often called weights, and we seek values for them that make the model fit the training data well' [@Hastie_2009, 395]). The novelty, however, consists in the 'hidden' units whose interactions 'represent important features.' In other words, the flat additive combination of features expressed in the `R` model formula above does not convey the interactions of these units. As Hastie and co-authors put it, 'the units in the middle of the network, computing the derived features $Z_m$, are called hidden units because the values $Z_m$ are not directly observed' [@Hastie_2009, 393]. These units can only viably interact in the neural nets because the back-propagation algorithm offers a way to create 'useful, new features' from the data. But because they interact through back-propagation, the hidden units 'capture' regularities in the 'task domain' and thereby do what counts as cognition in the connectionist philosophies associated with neural nets (see the PDP group's work [@McClelland_1986]).

The final major form in which neural net appear is the network diagram. Network graphs already appeared in Rosenblatt's perceptron work [@Rosenblatt_1958], but they ramify tremendously in the aftermath of back-propagation. Almost every book and article relating to  neural net presents some version of the diagram shown in Figure \ref{fig:titanic_net}.  

```{r plot_neural_net, cache=TRUE}
source('plot_nn.r')
pdf(file = 'figure/titanic_net.pdf')
print(plot.nn(titanic_net, fontsize=8, show.weights=FALSE))
dev.off()
```

\begin{figure}
  \centering
      \includegraphics[width=0.9\textwidth]{figure/titanic_net.pdf}
        \caption{Neural network topology for 5-hidden unit `titanic` data}
  \label{fig:titanic_net}
\end{figure}


Although the network topology of the model appears in many more complicated forms, it does several things in neural net literature. First, it presents a surface -- the input layer -- that indexes something in the world. \index{diagram!network} \index{graphic!network} The input layer, shown as $X$ in the algebraic diagram of equation \ref{eq:nn}, acts as an organ, an eye or perhaps a camera. Early neural net papers on the handwritten digital recognition problem sometimes describe cameras mounted above tables focused on images [@LeCun_1989].  Second, it presents an output layer that can contain single or multiple nodes, the $k$ of equation \ref{eq:nn}. In the `titanic` examples, a single target node appears (survived or not). In the `MNIST` handwritten digit recognition models, there are usually ten output nodes, one for each of the digits 0 ... 9.  Third, the network diagram orders forms of movement. Data and calculation propagate from bottom to top or vice-versa. (Sometimes the networks are rotated, and the flow is horizontal, but still bi-directional). Bi-directional hierarchical movement is key to the back-propagation algorithm in feed-forward and more complicated recurrent and convolutional neural nets. \index{algorithm!back-propagation} Fourth, it renders visible in principle the vital hidden nodes. Without the hidden nodes, neural nets revert to linear models. With the hidden nodes, the $Z_m$ of the equations \ref{eq:nn},  neural nets, like some other machine learners we have discussed such as support vector machines, effectively expand the common vector space by constructing new dimensions in it. The derived features or 'learned representations' (to use the language of [@Rumelhart_1986]) can expand indefinitely, according to different network topologies. Hidden nodes and hidden layers can multiply, bringing many new interactions and relations into the model (as we see in more recent revivals of neural net  as deep learning [@Hinton_2006]). \index{machine learner!neural net!hidden nodes}

## The subjects of a hidden operation

Given the diagrammatic forms of the basic model equations, the network diagram and the operational code comprising the privileged machine at work recognising handwritten digits or classifying the passengers on the Titanic, how are subject positions assigned? The layered architecture of model generates new textures, densities, pools, sparsities and saturations in both the common vector space, and in the operations of machine learners. How would we describe the figure of human machine learner in this setting? Is the human machine learner like Vlad, the former Eastern European mathematician tending the neural net at the heart of the call centre knowledge management system, or more like Heather Arthur, the programmer who wrote `kittydar`? \index{machine learner!kittydar} In _The Archaeology of Knowledge_, Michel Foucault refers to the 'position of the subject' \index{subject position} as an anchor point for the power-laden, epistemically conditioning enunciative functions called 'statements.'\index{statements!enunciative function}  When a subject position can be assigned, propositions, diagrams, numbers, models, calculations and data structures can come together in statements, as 'the specific forms of an accumulation' [@Foucault_1972, 125]. But this anchor point is not a unifying point grounded in interiority, in intentionality or even in  single speaking position or voice (that of *the*  machine learning expert, for instance). On the contrary, 'various enunciative modalities manifest his [sic] dispersion' [@Foucault_1972, 54]. In the mist of this dispersion, the position of subject derive from operations that determine statements that become a kind of law for the subject.\index{subject position!operational assignment of}

As Foucault puts it, in a formulation that anticipates accounts of performativity \index{performativity} that gain currency elsewhere several decades later,  

>in each case the position of the subject is linked to the existence of an operation that is both determined and present; in each case, the subject of the statement is also the subject of the operation (he who establishes the definition of a straight line is also he who states it; he who posits the existence of a finite series is also, and at the same time, he who states it) ; and in each case, the subject links, by means of this operation and the statement in which it is embodied, his future statements and operations (as an enunciating subject, he accepts this statement as his own law) [@Foucault_1972, 94-95]. 

Foucault's examples here include subjects who say things like 'I call straight any series of points that ...': just such statements  operate in machine learning. The _operation_ is crucial, since it connects many different practices and techniques (function finding, optimisation, transformation of data into the common vector space, mobilisation of probability distributions as a kind of rule of existence for learnable situations, etc.) that accompany, ornament, armour and diagram the statement. Foucault posits a subject-positioning circularity between the operation and accompanying statement: the subject of the statement is also the subject of the operation. The provisional coincidence of operation and statement both creates a subject position, with some agency and subjects machine learners to future operations.  The process might be formalised for any machine learner as follows: the diagrammatic operations of the machine learner support the production of statements; these operations become a way of producing future statements to the extent that the subject of the operation is _also_ the subject of the statement. The assignation of a subject position occurs in this forward and backward, feed-forwarding and back-propagating movement between operation and statement. \index{subject position!back-propagation} \index{machine learner!subject as|see{subject position}} 

Although the gap between what is done operationally and what is stated might seem  small, there are many slippages and divergences in it. A seemingly minor statements such as 'we see that Net-5 does the best, having errors of only 1.6%, compared to 13% for the “vanilla” network Net-2' [@Hastie_2009, 407] bears within it, in its  coupling to all the operations comprising 'Net-5,' a set of determinations and relations for variously positioned subjects. (These might include machine learners, such as Hinton or Le Cun, but also U.S. Postal workers).   In any concrete situation, in relation to any specific machine learner, the diagrammatic operations and statements will position subjects in specific ways. There is no simple referent here, no simple object facing a knowing or controlling subject, since on this account, the operations and statements in their dispersions, accumulations and distributions overflow any simple dyadic relation between a subject-object or human-machine/world. 


## Conclusion

Neural nets figure the vicissitudes of machine learners. They test machine learners for their capacity to understand how models relate to data, and they test them in relation to their knowledge and experience of how of to craft and to regularize the parameters of models in a given situation. Perhaps more profoundly, they figure a much more deeply competitive imperative that frames much of the practice in machine learning, and in many ways constrains thinking around machine learning. This competition is not always explicit or overt but it almost transpires in the form of examination.

In _Doing Data Science_, Cathy O'Neil and Rachel Schutt offer a pragmatic set of steps for working with machine learners:

>The big picture is that given data, a real-world classification problem, and constraints, you need to determine:
    >1. Which classifier to use
    >2. Which optimization method to employ
    >3. Which loss function to minimize
    >4. Which features to take from the data
    >5. Which evaluation metric to use

    [@Schutt_2013, 116]

Perhaps the order of these steps would be contested by some people, but each of these determinations is definitely a matter of competitive contention between machine learners. Competitions encourage a proliferation of determinations. Whether or not this competition deeply structures the field of machine learning is also a matter for further investigation. But in my experience, immersion in the field of machine learning, navigating its social textures, and getting a feeling for how it is to exist in these conditions definitely entails a sense of becoming the subject of examination. This feeling, which is concretised in the numerous machine learning competitions and online instructional materials, is convoluted because it concerns the constitution of machine learners themselves as operations of examination and as a form of documentation or statement of examination, test, and validation focused on errors. As Andrew Ng opines, 'we care about generalization error,' \index{Ng, Andrew} \index{error} \index{examination}, but error here both refers to something that neither the responsibility simply of the machine or its human others. Rather, responses to and care for this error move forwards and backwards between the human and non-human machine learners. 

Throughout this chapter, the proximities between the human and machine machine learners have been discussed in two different ways. First, in terms an intersection  that assigns a subject position through a machine learning operation. Neural nets epitomise this operation. The subject who operates the neural net as a way of fitting a model has been increasingly caught up in a network of neural nets that feed forward into a widening stream of parallel and layered architectures. This feeding-forward, however, is regularized or narrowed down through examination and error, through back-propagation on various scales. The principle of testing through generalization error that has long guided the supervision of machine learners has been generalized to an examination or test of machine learners in competitions. While competitive testing and examination has long been a constitutive practice in machine learning (for instance, in the case of the handwritten digit recognition contests of the 1990s), these competitions have now become a somewhat general mechanism for assigning positions to machine learners. The positions assigned include the subject positions of people whose participation in machine learning competitions are ranked and indeed the machines themselves, whose rankings rise and fall as different infrastructures, algorithms and optimisations come into play. 

Neural nets figure this re-iterative drawing of human-machine learning differences in several respects. Their own ups and downs, their potential to drive down error or learn features \index{feature} from data given enough data  derives less from some exotic mathematical abstraction or encompassing algorithm, and more from an accumulation of layers and connections between units of modelling. The central algorithm -- back-propagation -- is instructive here: it propagates errors throughout all the elements of the network, and every element in the network adjusts its weights in trying to minimise error. The predictive power of the model derives from the networked collective of linear models all seeking to optimise their error rates. So too, the competitive examinations that constitute and today generalize machine learning as a practice predicate the ongoing potential of hidden units -- machine learners  -- to learn from their rankings in tests of error, and to collectively learn from those rankings. 

What does this algorithmic propagation and generalization of errors and their testing mean for subject positions associated with machine learning? The cultural theorist Lauren Berlant describes optimism as an operation:

>The surrender to the return to the scene where the object hovers in its potentialities is the operation of optimism as an affective form [@Berlant_2007, 20]

This is a complicated formulation since it brings together surrender, return, scene, object, potentialities and affective form. All of these movements, places and things might be understood as purely psychic or even semiotic. I would suggest that optimism as an asignifying diagrammatic operation also plays out across the manifold surfaces of algorithms, datasets, models, platforms and ranking systems associated with machine learning as a competitive examination. The 'object hovers in its potentialities' in the case of machine learning because neural nets and their kin assimilate and adjust their weights in response to changes in infrastructures and in the generalization of operations to newly adjacent domains. Machine learners generate optimism through and about optimisation, an optimisation that is deeply predictive, prospective, anticipatory and abductive. But this adjusting of weights carried out through the propagation of errors is also inherently a ranking or examination.

The Kaggle competitions play out the logic of the operation of optimism most directly. The leaderboard is an examinatory ranking of efforts to drive down error rates, the same efforts that perhaps first affected machine learning around handwritten digit recognition in the 1990s. Entrants work hard to optimise the generalization errors of their models, but Kaggle aligns their work in a bootstrap aggregating architecture that assigns subject positions based around those error rates.  Maurizio Lazzarato describes how subject positions are assigned through examination, testing and ranking: 

>The thought and consciousness of the individuated subject come into play when there is an obstacle, a disturbance, or an "event." Then the subject, consciousness, and representation are used in order to modify the feedback relations among the human and non-human components of the automobile "machine," in order to reestablish the automatic mechanisms and machinic operations [@Lazzarato_2014, 89-90]

While the ensembles of machine learners we saw in the Kaggle competitions are not entirely individuated through the competitions, the Kaggle competitions certainly help establish new automatic and machinic operations. The ranking of machine learners (human and non-human) then should be understood as less a privileging of wonderful people at the intersection of infrastructure, human behaviour and mathematics and more as a re-iterative drawing of distinctions between people that at the same time modifies feed-forward and feed-back relations between people and machines. These relations almost always prospectively marginalize humans, even as they also inaugurate ontological mutations. 

[^7.3]: Unlike the cats detected by `kittydar,` the software discussed in the introduction to this book, the Google experiment did not use supervised learning. The deep learning approach was unsupervised [@Markoff_2012]. That is, the neural nets were not trained using labelled images of cats.


[^7.01]: In the case of this paper, and many others related to neural nets, the images are of hand-written digits. These digits have an almost constitutive role, as I discuss in this chapter. 


[^7.1]: In earlier work on machine learning [@Mackenzie_2013], I presented programmers as agents of anticipation, suggesting that the turn to machine learning amongst programmers could be useful in understanding how predictivity was being done amidst broader shift to the regime of anticipation described by Vincenne Adams, Michelle Murphy and Adele Clarke [@Adams_2009]. Subsequently developments in machine learning, even just in the last three years, confirm that view, but in this chapter and in this book more generally, I focus less on transformations in programming practice and software development, and more on the asymmetries of different machine learner  subjects in relation to infrastructures and knowledge.

[^7.2]: Other figures we might follow include Claudia Perlich, Andrew Ng, Geoffrey Hinton, Corinna Cortez,  Daphne Koller, Christopher Bishop, Yann LeCun, or Jeff Hammerbacher.  Although some women's names appear here, in any such list, men's names are much more likely to appear. This is no accident.  \index{machine learner!gender of}


[^7.005]: Although mainstream media accounts of machine learning are not the focus of my interest here, it is hard to ignore the extraordinary level of interest that deep learning projects and techniques have attracted in the last few years. Articles have appeared in all the usual places -- _The New York Times_ [@Markoff_2012], _Wired_[@Garling_2015], or _The Guardian_ [@Arthur_2015]. In many of these accounts, machine learning and neural nets in particular appear both in the guise of the existential threat of artificial intelligence and as a mundane device (for instance, speech recognition on a mobile phone). The spectacular character of deep learning could be analysed in terms like that of the genomes discussed in chapter \ref{ch:genome}. In both cases, the advent and transformation of these machine learners is closely linked to networked platforms (such as Google, Facebook, Yahoo and Microsoft) and their efforts to encompass within their services as many elements of experience, exchange, communication and power as possible. Deep learning  machine learners currently focus mostly on images (photographs and video) and sounds (speech and music), and usually attempt to locate and label objects, words or genres.  \index{machine learner!deep learning!existential threat of}

[^7.0010]: We saw some use of neural nets in genomics in the previous chapter (\ref{ch:genome}). The initial publication of the `SRBCT` microarray dataset in [@Khan_2001] relied on neural nets.  \index{dataset!SRBCT}

[^7.0011]: Neural nets also receive uneven attention in the machine learning literature. In Andrew Ng's Stanford CS229 lectures from 2007, they receive somewhat short shrift: around 30 minutes of discussion in Lecture 6, in between Naive Bayes classifiers and several weeks of lectures on support vector machines [@Ng_2008b]. As he introduces a video of an autonomous vehicle steered by a neural net after a 20 minute training session with a human driver, Ng comments that 'neural nets were the best for many years.' \index{Ng, Andrew} The lectures quickly moves on to the successor, support vector machines.  In _Elements of Statistical Learning_, a whole chapter appears on the topic, but prefaced by a discussion of the antecedent statistical method of 'projection pursuit regression.' The inception of 'projection pursuit' is dated to 1974, and thus precedes the 1980s work on neural nets that was to receive so much attention. In _An Introduction to Statistical Learning with Applications in R_, a book whose authors include Hastie and Tibshirani, neural nets are not discussed and indeed not mentioned [@James_2013]. Textbooks written by computer scientists such as Ethem Alpaydin's _Introduction to Machine Learning_ do usually include at least a chapter on them, sometimes under different titles such as 'multi-layer perceptrons' [@Alpaydin_2010].  Willi Richert and Luis Pedro Coelho's _Building Machine Learning Systems with Python_ likewise does not mention them [@Richert_2013]. Cathy O'Neil and Rachel Schutt's _Doing Data Science_ mentions them but does not discuss them [@Schutt_2013], whereas both Brett Lantz's _Machine Learning with R_ [@Lantz_2013] and Matthew Kirk's _Thoughtful Machine Learning_ [@Kirk_2014] devote chapters to them. In the broader cannon of machine learning texts, the computer scientist Christopher Bishop's heavily cited books on pattern recognition dwell extensively on neural nets [@Bishop_1995; @Bishop_2006]. Amongst statisticians, Brian Ripley's _Pattern Recognition and Neural Networks_ [@Ripley_1996], also highly cited, placed a great deal of emphasis on them. But these specific documents against a pointillistic background of hundreds of thousands of scientific publications mentioning or making use of neural nets since the late 1980s in the usual litany of fields -- atmospheric sciences, biosensors, botany, power systems, water resource management, internal medicine, etc. This swollen publication tide attests to some kind of formation or configuration of knowledge invested in these particular techniques, perhaps more so than other I have discussed so far ( logistic regression, support vector machine, decision trees, random forests, linear discriminant analysis, etc.). 

[^7.7]: In any case, social media and search engines cannot be understood apart from the machine learning techniques that have been thoroughly woven through them since their inception.  Hence _Elements of Statistical Learning_ devotes several pages to Google's famous _PageRank_ algorithm, describing it as an unsupervised learner [@Hastie_2009, 576-578]. \index{social media platforms!machine learning as part of}

[^7.32]: Recent work on deep belief networks replaces the sigmoid function with other non-linear functions that subtly alter the way layers of neural nets relate to each other. See [TBA - the Gorlot paper] 
