
\chapter{The problem of pattern: accretion and movement in data }
\label{ch:pattern}

The contemporary form of knowledge as pattern-finding 
Finding patterns in data
Movements in machine learning: patterns of generalization
Dimensional exuberance and pattern recognition in movement


```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, tidy=TRUE, fig.height=8, 
                      echo=FALSE,  results='hide', warning=FALSE, message=FALSE, dev='pdf')
library(RSQLite)
con <- dbConnect(RSQLite::SQLite(),'../ml_lit/all_refs.sqlite3')
res = dbGetQuery(con, statement ="select * from basic_refs limit 10;")
library(ggplot2)
library(stringr)
library(xtable)
options(xtable.comment = FALSE)
simpleCap <- function(x) {
    s <- strsplit(x, " |-")[[1]]
    return (paste(toupper(substring(s, 1,1)), tolower(substring(s, 2)), sep="", collapse=" "))
  }
```

> Algorithms for pattern recognition were therefore from the very beginning associated with the construction of linear decision surfaces [@Cortes_1995, 273-4] \index{Cortes, Corinna!on pattern recognition}

> The notion of pattern involves the concept of different modes of togetherness [@Whitehead_1956, 195-6]. \index{Whitehead, Alfred North!on pattern} \index{pattern!modes of togetherness}

Should we hold machine learners accountable for their claims to recognise patterns in data?[^5.04]  \index{machine learner!pattern recognition|seealso{pattern}}  This chapter explores some major derivations of pattern developing in the last decades of the twentieth century, and in particular, how the notion of pattern itself changes as a result. We have seen the emergence of the common vector space and its epistopic transformations, the multiplication of functions and their associated partial observers, and then the re-distribution of probability that transform machine learners into populations of error-sensitive learners. Across the vectors, functions and distributions, a  diagram of  machine learning weaves and knots many points of emergence, continuity and conjunction. Formidable accumulations of infrastructure, devices and expertise accrete around machine learning. Already on several occasions problems of the density and mass of techniques, research literature, algorithms and code have appeared. What in this diagram and in the forest-like growth of techniques, projects, applications and proponents allows us to make sense of the dynamics of this accumulation? Via two highly developed and heavily used machine learners  -- decision trees and support vector machines  -- that more or less mesmerised the machine learning scientific literature between 1980-2000, some relatively novel and somewhat discontinuous forms of movement into and through data were initiated.[^5.05]  \index{machine learner!decision tree} \index{machine learner!support vector machine} \index{machine learner!neural net}   These diagrammatic movements, which we might characterise as _splitting_, and _hyper-planing_ -- animate machine learners in producing newer techniques (random forests, deep belief nets), intensifying their application, and perhaps more importantly, re-configuring what counts as pattern. \index{diagram!movement}

```{r pattern}
    q = "select * from basic_refs where TI like '%pattern%' or DE like '%pattern%'"
    q2 = "select TI, PY, TC from basic_refs where TC >500"
    res2 = dbGetQuery(con, q2)
    head(res2[order(res2$TC, decreasing=TRUE),])

    res = dbGetQuery(con, q)
    dim(res)
    hist(res$PY)
    table(res$PY)
```

The decision tree and support vector machine appear in the machine learning scientific publications of the last three decades (1980-2010) as important novelties. They also, I will suggest, embody a new enunciative modality \index{enunciative modality} in  which regularity and rarity, accumulation and sparsity do not exist in tension or opposition. Practically,  they  loom large in various contemporary accounts of machine learning as a way of knowing (for instance, in the popular machine learning  books such as _Machine Learning for Hackers_ [@Conway_2012] or _Doing Data Science_ [@Schutt_2013]). The machine learning research published in statistics, computer science, mathematics, artificial intelligence and a swathe of related scientific fields bristles with references to decision trees, support vector machine and neural networks. (As we will see, these techniques themselves ingest substantial supplies of knowledge and practice produced by  social sciences, insurance and actuarial practice, and marketing research.) The top 20 most cited publications in the field include  decision trees Ross Quinlan and Leo Breiman's papers [@Quinlan_1986; @Breiman_1984], Vladimir Vapnik and Corinna Cortes' support vector machines papers [@Vapnik_1999; @Cortes_1995],  an early textbook written by a computer scientist on machine learning [@Mitchell_1997], a textbook and software package on data mining using Java [@Witten_2005]; a textbook on pattern recognition dating from the 1970s [@Duda_2012], a tutorial on an error control technique (ROC - Receiver Operating Characteristics, first developed by the US military during WWII) and somewhat lower, another well-known textbook, this time on neural networks and pattern recognition [@Bishop_2006]. 

A striking _positivity_ takes shape across these differences. \index{positivity} I understand positivity partly in the sense of 'positivism,' an epistemological stance widely shared amongst machine learners  that views knowledge as deriving from observation.[^5.01]  The re-configuration of tensions between sparsity and accumulation, between rarity and commonality can also be understood in the sense of positivity proposed by Michel Foucault in _The Archaeology of Knowledge_ [@Foucault_1972], a book whose somewhat a-subjective approach to knowledge unexpectedly resonates with much that takes place in machine learning. Foucault theorises statements as elements _posited_ or deposited in a relatively uneven space of transformations. These transformations generate what can be thought and done within a particular discursive formation. Their positivity arises from the 'forms of an accumulation':

>To describe a group of statements not as the closed, plethoric totality of a meaning, but as an incomplete, fragmented figure; to describe a group of statements not with reference to the interiority of an intention, a thought, or a subject, but in accordance with the dispersion of an exteriority; to describe a group of statements, in order to rediscover not the moment or the trace of their origin, but the specific forms of an accumulation, is certainly not to uncover an interpretation, to discover a foundation, or to free constituent acts; nor is it to decide on a rationality, or to embrace a teleology. It is to establish what I am quite willing to call a positivity [@Foucault_1972, 125]. \index{Foucault, Michel!on positivity} \index{Foucault, Michel!on statements}  

Foucault frequently refers to positivity in those passages where he discusses  how practices develop and how statements relating to those practices take shape. The situations in which he invokes positivity involve accumulations or plethora of statements whose coherence or unity does not lie in any origin, intention, rationality, subjectivity, ideology or teleology. Like mineral strata, positivities entail accumulations without deriving them from underlying unities. Foucault's acceptance of 'positivity,' in spite of the risk of attracting criticism, pivots on his interest in, and practical commitment to, understanding accumulations and uneven distributions. A positivity is a way of describing 'a group of statements' that may be quite dispersed, discontinuous, and unevenly distributed because they make up 'a population of events in the space of discourse in general' [@Foucault_1972, 27]. If _The Archaeology of Knowledge_ is a book concerned with the tissue of relations between things, what is said, what is done and what counts as knowledge (scientific or not) in practice, then the persistent and at times forceful reiteration of _dispersion_ and _discontinuity_ might seem strangely counterintuitive. But Foucault's insistence on discourse as something irreducible to logic, language, intention, rationality, experience, historical origin or narrative, and as the way in which 'statements' ('a graph, a growth curve, an age pyramid, a distribution cloud are all statements' (82)), strategies, concepts, objects, archives and knowledges come together in 'discursive formations' is an attempt to resist any resort to a hidden order or concealed origin that critical thought could uncover or interpret. \index{positivity!accumulation in}

Positivity as accumulation helps make sense of a field such as machine learning whose many different forms, techniques, statements, institutions, values and associations often grapple with problems of much data, many variables, many possible models and a sense of an elusive but vital order or pattern discoverable amidst accumulation. Rather than seeing pattern as something discovered in data, the notion of positivity suggests we should examine the diagrammatic operations that configure the practice of machine learning, giving rise to a field of objects and subject positions.   The three machine learners that anchor this chapter are at once perhaps the most distinctive data mining, pattern recognition and predictive modelling achievements of the late twentieth century (at least judging by the citational and implementational interest they attract). They different greatly in how they move through data. At certain times, they come together (for instance, in machine learning competitions discussed in chapter \ref{ch:subjects}; or in certain formalizations such as machine learning  theory or in graphs of the bias-variance decomposition discussed in chapter \ref{ch:probability}; or in the pedagogy of machine learning discussed in chapter \ref{ch:diagram}). We might understand their differences in terms of 'enunciative modalities' [@Foucault_1972, 54] \index{enunciative modality}  -- that is, in terms of the different ways in which they give rise to statements. Every  machine learner generates statements, but from different places, by somewhat different individuals, and from the different situations they that are able 'to occupy in relation to the various domains or groups of objects' [@Foucault_1972, 52].[^5.02]

## Splitting or recursive binary partitioning and the growth of trees

> Mastering the details of tree growth and management is an excellent way to understand the activities of learning machine generally [@Malley_2011, 118].

Decision trees promise an understanding of machine learning.   The enunciative function of the decision tree partly relates to the observability and comprehensibility of machine learning. As we will see, not all machine learners readily supports observation or comprehension. The cost of comprehensibility, however, is a certain highly restricted movement through the data, a movement that has been difficult to stabilise and control. As _Elements of Statistical Learning_ puts it: 'tree-based methods partition the feature space into a set of rectangles, and then fit a simple model (like a constant) in each one. They are conceptually simple yet powerful' [@Hastie_2009, 305].

\begin{figure}
  \centering
      \includegraphics[width=0.9\textwidth]{figure/recursive_partitioning.pdf}
        \caption{Recursive partitioning of the feature space}
  \label{fig:rpart}
\end{figure}

Tree-based methods are supervised learners as they require the data to either be labelled with a class or to have some outcome value. \index{machine learning!supervised} The variable types in the feature space (or common vector space of the data) \index{feature space|see{common vector space}} can be mixed. Because the method cuts the common vector space into a tiled surface, the features or data variables can be continuous or discontinuous. The 'simple models' that tree methods construct each inhabit one of the rectangular regions or partitions of the feature space. In Figure \ref{fig:rpart}, the different regions or partitions produced by a decision tree are labelled $R1, R2$ etc. 

```{r aid, echo=FALSE, results='asis', cache=TRUE } 
    q = "select * from basic_refs where anchor = 'morgan_sonquist_WOS'"
    res = dbGetQuery(con, q)
    res$TI = sapply(res$TI, simpleCap) 
    tab  = xtable(res[1:10,c('TI', 'PY' ,'TC')], caption="References to Morgan and Sonquist's Automatic Interaction Detector", label='table:aid_citation')
    print(tab, type='latex')
   
```

Work on classification and regression techniques using decision trees goes back to the  early 1960s when social scientists James Morgan and John Sonquist at the University of Michigan's Institute for Social Research were attempting to analyse increasingly large social survey datasets [@Morgan_1963].  As Dan Steinberg describes in his brief history of decision trees [@Steinberg_2009, 180], the  'automatic interaction detector' (AID) as it was known, sought to automate the practice of data analysts looking for interactions between different variables. The variety and sheer optimism of subsequent applications of these prototype decision tree techniques  is striking.  In the 1960s and 1970s, papers that drew on the AID paper or use AID techniques can be found, as table \ref{table:aid_citation} shows, in education, politics, economics, population control, advertising, mass media and family planning. \index{machine learner!decision tree!history of} \index{machine learner!automatic interaction detector}

\begin{figure}
  \centering
      \includegraphics[width=0.9\textwidth]{figure/morgan_sonquist_p430.pdf}
        \caption{AID classifies annual earnings (Morgan & Sonquist, 1963, 430)}
  \label{fig:aid_tree}
\end{figure}

A decade after initial work, the AID was the object of trenchant criticism by statisticians and others, not for the classifications it used (see Figure \ref{fig:aid_tree}), but for its dangerous empiricism. Writing in the 1970s, statisticians in the behavioural sciences such as Hillel Einhorn at the University of Chicago castigated the use of such techniques.   The criticisms stemmed from  a general distrust of 'purely empirical methods', and scepticism focused on their positivity:

> The purely empirical approach is particularly dangerous in an age when computers and packaged programs are readily available, since there is temptation to substitute immediate empirical analysis for more analytic thought and theory building. It is also probably too much to hope that a majority of researchers will take the time to find out how and why a particular program works. The chief interest will continue to be in the output-the results-with as little delay as possible [@Einhorn_1972, 368]

Einhorn discusses AID alongside other  techniques such as factor analysis and multi-dimensional scaling (both still widely used) before concluding 'it should be clear that proceeding without a theory and with powerful data analytic techniques can lead to large numbers of Type I errors' [@Einhorn_1972, 378]. His statistical objections to AID are  particularly focused on the problematic power of the technique: 'it may make sense out of "noise"' (369). Consequently, researchers easily misuse the technique: they 'overfit' the data, and do not pay enough attention to issues of validation (369-370). Similarly the British marketing researcher Peter Doyle, criticising the use of AID in assessing store performance and site selection by operations researchers, complained that searching for patterns in data using data sets was bound to lead to spurious results and the decision trees, although intuitively appealing (that is, they could be easily interpreted), were afflicted with arbitrariness: 'a second variable may be almost as discriminating as the one chosen, but if the program is made to split on this, quite a different tree occurs' [@Doyle_1973, 465-466].[^5.03] \index{operations research!use of decision trees} \index{machine learner!overfitting} Most of these criticisms can be seen as expressing conventional statistical caution in response to threats to validity. 

## 1984: Cutbacks on recursive partitioning

As Einhorn expected, it was too much to hope that all researchers would take time to investigate how a particular program works. Some researchers however did take time, years in fact, to investigate how decision trees work.  Writing around 2000, Hastie, Tibshirani and Friedman, who could hardly by accused of not understanding decision trees,  happily recommend decision trees as the best 'off-the-shelf' classifier:  'of all the well-known learning methods, decision trees comes closest to meeting the requirements for serving as an off-the-shelf procedure for data-mining' [@Hastie_2009, 352]. We might wonder here, however, whether they damn with faint praise, since 'off-the-shelf' suggests pre-packaged, and commodified, and the term 'data-mining' itself is not without negative connotations. As for its commercial realities, in 2013, Salford Systems, the purveyors of the leading contemporary commercial decision tree software, `CART`, could claim:

 > CART is the ultimate classification tree that has revolutionized the entire field of advanced analytics and inaugurated the current era of data mining. CART, which is continually being improved, is one of the most important tools in modern data mining. Others have tried to copy CART but no one has succeeded as evidenced by unmatched accuracy, performance, feature set, built-in automation and ease of use. [Salford Systems](http://www.salford-systems.com/products/cart) \index{machine learner!CART}

What happened between 1973 and 2013?  Decision trees somehow grew out of  the statistically murky waters of social science departments and business schools in the early 1970s to inaugurate the 'current era of data mining' (which the scientific literature indicates starts in the early 1990s). \index{data mining!in 1970} This was not only a commercial innovation. As the earlier citation from U.S. National Institutes of Health biostatisticians Malley, Malley and Pajevic indicates, decision trees enjoy high regard even in biomedical research, a setting where statistical rigour is highly valued for life and death reasons. The happy situation of decision trees four decades on suggests some kind of threshold was crossed in which the epistemological, statistical, or algorithmic ('built-in automation') power of the technique altered substantially. \index{machine learner!decision tree!in medicine}

The third author of _Elements of Statistical Learning_, Jerome Friedman, worked at the  U.S. Department of Energy's Stanford Linear Accelerator during the late 1970s. Friedman  was  instrumental in rescuing  decision trees from the ignominy of profligate ease of use and pure empiricism they had endured since the late  1960s. \index{Friedman, Jerome!work on decision tree} The reorganisation and statistical retrofitting of the decision tree was not a single or focused effort. During the 1980s, statisticians such  as Friedman and Leo Breiman renovated the decision tree as a statistical tool [@Breiman_1984] at the same time as  computer scientists such as  Ross Quinlan in Sydney were re-implementing decision trees guided by an  artificial intelligence-based formalisation as rule-based induction technique [@Quinlan_1986].[^5.1] \index{Quinlan, John Ross!} \index{artificial intelligence!rule-based induction} This uneasy parallel effort between computer science and statistics  still somewhat strains relations in machine learning today. Both statisticians and computer scientists  do and use the same techniques, but often with the computer scientists focusing on optimisation and algorithmic scale and the statisticians inventing novel statistical formalizations and abstractions. The fateful embrace of statistics and computer science, the disciplinary binary that vectorizes machine learning, has been generative in the retrieval of the decision tree. \index{vectorization}

An initial symptom of the transformation of the technique appears in a name change. The term 'decision tree,' although still widely used in the research literature and machine learner  parlance was replaced by 'classification and regression tree' during the late 1970s and 1980s. The terms 'classification and regression tree' is sometimes contracted to 'CART,' and that term strictly speaking refers to a computer program described in [@Breiman_1984] as well as the title of that highly-cited monograph. \index{machine learner!CART} \index{Breiman, Leo!CART monograph} As we have seen in previous chapters, regression and classification designate the two main sides of machine learning practice, and their concatenation with 'tree' attests to the ongoing renovation of existing machine learning approaches with a more statistical facade.

The implementation of machine learning techniques in `R`  by virtue of `R`'s statistical provenance perhaps favours the statistical side of decision tree practice, but that has certain forensic virtues not offered by commercial or closed-source software often produced by computer scientists.  In this case, the name of one long-standing and widely-used `R` package itself attests to something: `rpart` is a contraction of 'recursive partitioning' and this term generally describes how the decision tree algorithm works to partition the common vector space into the form shown in Figure \ref{fig:rpart} [@Therneau_2015].  'CART,' on the other hand, is a registered trademark of Salford Systems, the software company mentioned above, who sell the leading commercial implementation of classification and regression trees. Hence, the `R` package `rpart` cannot call itself the more obvious name `cart, ` and instead invokes the algorithmic process it relies on: recursive partitioning.[^5.06] \index{algorithm!recursive partitioning} \index{programming languages!R!packages!rpart}

```{r iris_tree, echo=TRUE, cache=TRUE, message=FALSE, warning=FALSE, comment=NA, size='smallsize', results='asis' } 

    data(iris)
    library(rpart)
    iris_tree =rpart(Species ~ ., iris)
```

R.A. Fisher's _iris_ dataset, which contains measurements made in the 1930s of petal and sepal lengths of _iris virginica, iris setosa_ and _iris versicolor_ is a standard instructional example for decision trees [@Fisher_1938].[^5.12] \index{dataset!iris} The code shown here loads the _iris_ data (the dataset  is routinely installed with many data analysis tools), loads the `rpart` decision tree library, and builds a decision to classify the irises by species.  What has happened to the iris data in this decision tree? The `R` code that invokes the recursive partitioning algorithm is so brief `iris_tree =rpart(Species ~ ., iris)` that we can't tell much about how the data has been 'recursively partitioned.' We know that the _iris_ has `r nrow(iris)` rows, and that there are equal numbers of the three iris varieties. 

Code brevity indicates a great deal of formalization of practice has accrued around decision trees. \index{code!brevity of!} Some of this formalization was described in the landmark _Classification and Regression Trees_ monograph [@Breiman_1984], and is familiar (from previous chapters)  in  its use of common vector space, functions as partial observers and the re-distribution of chance between devices and worlds. For instance, Breiman and co-authors start out by placing the technique in vector space:

> Define the measurements $(x_1, x_2, ...)$ made on a case as the _measurement vector_ $\mathbf{x}$ corresponding to the case. [@Breiman_1984, 3]

In re-configuring decision trees as classification trees, Breiman first address issues of data dimensionality. \index{data!dimensionality} Next,  rather than simply invoking the notion of a classifier as a device (as AID did), the CART monograph defines a classifier in terms of a function: 

> A classifier or classification rule is a function $d(\mathbf{x})$ defined on $\mathcal{X}$ so that for every $\mathbf{x}$, $d(\mathbf{x})$ is equal to one of the numbers $1, 2, ..., J$. [@Breiman_1984, 4]

Note that 'rule,' the term that computer scientists working in artificial intelligence at the time were using, still figures here, but the classifier is formalized as a function. \index{machine learner!  The definition of the classifier as a function depends on the existence of a vector space $\mathcal{X}$, a set of measurement, feature or predictor vectors  $\mathbf{x}$ and a set of responses  $1, 2, ..., J$. Second, classification is understood in terms of a series of binary splits in the different dimensions of the common vector space. These splits comprise the tree structure diagram. The problem here is that many splits are possible. 

> The first problem in tree construction is how to use $\mathcal{L}$ to determine the binary splits of $\mathcal{X}$ into smaller pieces. The fundamental idea is to select each split of a subset so that the data in each of the descendant subsets are "purer" than the data in the parent subset [@Breiman_1984, 23].

Tree construction hinges on the notion of purity or more precisely 'node impurity', a function that measures how data labelled as belonging to different classes are mixed together at a given branch or node in a decision tree: 'that is, the node impurity is largest when all classes are equally mixed together in it, and smallest when the node contains only one class' [@Breiman_1984, 24]. As Malley and co-authors note, 'the collection of purity measures is still a subject of research' [@Malley_2011, 123], but Breiman, Friedman, Olshen and Stone promoted a particular form of impurity measure for classification trees known as 'Gini index of diversity' [@Breiman_1984, 38].  Like the cost function used in optimising linear and logistic regression models, the measures of node impurity allow the process of tree construction to be understood as a kind of non-linear movement through space. Whereas in gradient descent, the intuition was 'always go down to the valley as quickly as possible', here the intuition is more like: ' split things in ways that minimize mixing'. Good splits decrease the level of impurity in the tree. In an ideal tree with maximum purity, each terminal node -- the nodes at the base of the tree -- would contain a single class. 

The results of the application of measures of node impurity can be seen below in two plots. 

```{r iris_tree_plot, echo=FALSE,  fig.cap ='Decision tree on _iris_ dataset', cache=TRUE, message=FALSE, warning=FALSE, comment=NA, results='asis' } 

    par(mfrow = c(1,2), xpd=NA)
    plot(iris_tree, main = 'tree recursive partitioning of iris')
    text(iris_tree, cex=0.8, use.n=TRUE)
    table(iris$Species) # is data.frame with 'Species' factor
     iS <- iris$Species == "setosa"
     iV <- iris$Species == "versicolor"
     op <- par(bg = "bisque")
     matplot(c(1, 8), c(0, 4.5), type =  "n", xlab = "Length", ylab = "Width",
             main = "Petal and Sepal Dimensions in Iris Blossoms")
     matpoints(iris[iS,c(1,3)], iris[iS,c(2,4)], pch = "sS", col = c(2,4))
     matpoints(iris[iV,c(1,3)], iris[iV,c(2,4)], pch = "vV", col = c(2,4))
     legend(1, 4, c("    Setosa Petals", "    Setosa Sepals",
                    "Versicolor Petals", "Versicolor Sepals"),
            pch = "sSvV", col = rep(c(2,4), 2))
```

In Figure \ref{iris_tree_plot}, the plot on the left shows the decision tree and the plot on the right shows just _setosa_ and _versicolor_  plotted by petal and sepal widths and lengths.  As the plot on the right shows, most of the measurements are well clustered. Only the _setosa_ petal lengths and widths seem to vary widely. All the other measurements are tightly bunched. This means that the decision tree shown on the left has little trouble classifying the irises. Decision trees are read from the top down, left to right. The top level of this tree can be read, for instance, as saying, if the length of petal is less 2.45, then the iris is _setosa_.


## Conclusion

What do we learn from decision tree and their development into random forests, or from linear discriminant analysis and its re-formalization as the support vector machine? We could also have tracked the movement between the perceptron [@Rosenblatt_1958] and the 'deep learning' convolutional neural networks [@Hinton_2006] of more recent practice (see chapter \ref{ch:subjects} . Each of these shifts attests, I have been suggesting, to the emergence of a new enunciative mode in which dispersion and regularity, accumulation and rarity are no longer opposed, or treated as contradictory. A single decision tree becomes thousands in a random forest. A relatively small number of dimensions in the common vector space becomes potentially infinite in the kernelisation practiced in support vector machines. Models that sought to encompass or fit everything in the data including the outliers within a single probability distribution instead dwell on the difficult-to-classify, the erroneous or borderline instances amidst the massive normalized accumulations of event.

What counts as knowledge today?  The recent machine learners can be studied, I have been suggesting  as elements in a power-laden positivity, \index{positivity!of knowledge} that changes knowledge in specific, actually quite narrow, but nevertheless far-reaching ways. The interpretability of a decision tree cascades into the statistical regularities associated with a random forest of many thousands of trees. The separating lines and planes that allow linear models to become classifiers in the 'classic' techniques such as linear discriminant analysis find themselves displaced into hyper-planes, into newly constructed and sometimes inordinately-dimensioned feature spaces that can only be traversed by virtue of the kernel functions, and their computationally tractable inner products. These transformations are not necessarily scientific, although various sciences make great use of them (as the following chapters will explore), but they do diagram  thresholds of epistemologization and formalization in practice.  We saw in the case of the MNIST postal digits that a field of objects can furnish referentials for an enunciative mode, and that these referentials construct levels on which different machine learners move transversally. Although neural networks, support vector machines and decision trees have quite different ways of partitioning, separating or propagating differences, they together generate a group space that can attract and accumulate propositions. This group space, despite its commonalities, is not an homogeneous field. It does not have the coherence of a science, it uses different systems of formalization (the cross-entropy measures of the decision tree, the gradient descent of the neural network, the dual form Lagrange multipliers of the support vector machine), and disperses in different ways across knowledge practice (the decision tree with its commercial uptake in data mining versus the support vector machine's heavy use in image recognition and classification). 


[^5.01]: See [@Kitchin_2014] for a survey of the major epistemological commitments in 'Big Data' discourse. These largely derived from predictive and inferential techniques of machine learners. \index{Kitchin, Rob!on epistemology of 'big data'}

[^5.02]:  While Foucault tends to retain a decoupled subject-object relation in the production of statements,  I tend to see these enunciative modalities as distributed across people and things. As always, machine learner is a composite term for this distribution. 

[^5.03]:  These objections and resistances to early decision trees echo today in discussions around pattern recognition, knowledge discovery and data-mining in science and commerce.  The problem of what computers do to the analysis of empirical data is long-standing. 

[^5.04]: Many authors have suggested that algorithms should be the focus of more attention. Mike Savage, in his account of the growth of 'descriptive assemblages' based around large scale data mining of transactions, administrative records and social media practice concludes:

    It follows that a core concern might be to scrutinize how pattern is derived and produced in social inscription devices, as a means of considering the robustness of such derivations, what may be left out or made invisible from them, and so forth. We need to develop an account which seeks to criticize notions of the descriptive insofar as this involves the simple generation of categories and groups, and instead focus on the fluid and intensive generation of potential [@Savage_2009, 171] \index{Savage, Mike|on descriptive assemblage}

[^5.05]: The other major machine learner of this time, the neural net, will be discussed in chapter \ref{ch:subject}.


[^5.1]: Quinlan's papers  and book on versions of the decision tree (`ID3` and `c4.5`) are both amongst the top ten the most highly cited references in the machine literature itself. Google Scholar reports over 20,00 citations of the Quinlan's book _C4.5: Programs for Machine Learning_ [@Quinlan_1993] (although far fewer appear in Thomson Reuters Web of Science). Several years ago,  `C4.5` was voted the top data mining algorithm [@Wu_2008]. While I don't discuss Quinlan's work in much detail here, we should note as a computer scientist, Quinlan takes a much more rule-based approach to decision tree that Breiman and co-authors. \index{Quinlan, John Ross!induction tree} \index{machine learner!C4.5}

[^5.06]:  Other `R` packages such as `party` [@Hothorn_2006] and `tree` [@Ripley_2014] also use recursive partitioning, but with various tweaks and optimisations that I leave aside here. 

[^5.12]:  `iris` is a very small dataset, a pre-computational miniature.  That diminutive character makes it diagrammatically mobile.  It supports a rhizomatic ecosystem of examples scattered across the machine learning literature.    The usual framing of the classification problem is how to decide whether a given iris blossom is of the species _virginica_, _setosa_ or _versicolor_.   These irises don't grow in forests -- they are more often found in riverbanks and meadows --  but they do offer a variety of illustrations of how machine learning classifiers are brought to bear on classification problems. Here the classification problem is taxonomic - the  iris genus has various sub-genera, and sections within the sub-genera. _Setosa, _virginica_ and _versicolor_ all belong to the sub-genus _Limniris_. This botanical context is  routinely ignored in  machine learning applications. In machine learning textbooks and tutorials, iris typically would be used to demonstrate how cleanly a classifier can separate the different kinds of irises. \index{dataset!iris}
