
\chapter{The problem of pattern: accumulation and dispersions }
\label{ch:pattern}

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, tidy=TRUE, fig.height=8, 
                      echo=FALSE,  results='hide', fig.show='hide', warning=FALSE, message=FALSE, dev='pdf')
library(RSQLite)
con <- dbConnect(RSQLite::SQLite(),'../ml_lit/all_refs.sqlite3')
res = dbGetQuery(con, statement ="select * from basic_refs limit 10;")
library(ggplot2)
library(stringr)
library(xtable)
options(xtable.comment = FALSE)
simpleCap <- function(x) {
    s <- strsplit(x, " |-")[[1]]
    return (paste(toupper(substring(s, 1,1)), tolower(substring(s, 2)), sep="", collapse=" "))
  }
```

> Algorithms for pattern recognition were therefore from the very beginning associated with the construction of linear decision surfaces [@Cortes_1995, 273-4] \index{Cortes, Corinna!on pattern recognition}

> The notion of pattern involves the concept of different modes of togetherness [@Whitehead_1956, 195-6]. \index{Whitehead, Alfred North!on pattern} \index{pattern!modes of togetherness}

Should we hold machine learners accountable for their claims to recognise patterns in data?[^5.04]  \index{machine learner!pattern recognition|seealso{pattern}}  This chapter explores some major derivations of pattern developing in the last decades of the twentieth century, and in particular, how the notion of pattern itself changes as a result. We have seen the emergence of the common vector space and its epistopic transformations, the multiplication of functions and their associated partial observers, and then the re-distribution of probability that transform machine learners into populations of error-sensitive learners. On several occasions, problems of the density and mass of techniques, research literature, algorithms and code have appeared. What in this diagram and in the forest-like growth of techniques, projects, applications and proponents allows us to make sense of the dynamics of this accumulation? Across the vectors, functions and distributions, a  diagram of  machine learning weaves and knots many points of emergence, continuity and conjunction. I view the formidable accumulations of infrastructure, devices and expertise accreting around machine learning as multi-faceted abstractions, where abstraction is understood diagrammatically. \index{abstraction!diagrammatic} Two highly developed and heavily used machine learners  -- decision trees and support vector machines  -- more or less mesmerised the machine learning scientific literature between 1980-2000. They initiated relatively novel and somewhat discontinuous diagrammatic movements into data.[^5.05]  \index{machine learner!decision tree} \index{machine learner!support vector machine} \index{machine learner!neural net}   These diagrammatic movements, which we might characterise as _splitting_, and _hyper-planing_ not only animate machine learners in producing newer techniques (random forests, deep belief nets) and intensifying their application. Perhaps more importantly, they re-open the question of what counts as pattern. \index{diagram!movement} If machine learning has no single or unified idea of pattern, then claims that machine learners find hidden patterns, that they traverse the occluded volumes of data, should be evaluated carefully.[^5.08] 

The decision tree and support vector machine appear in the machine learning scientific publications of the last three decades (1980-2010) as important novelties. They also, I will suggest, embody a new enunciative modality \index{enunciative modality!see{statement!enunciative modality of}} in  which regularity and rarity, accumulation and sparsity do not exist in tension or opposition. Practically,  they  loom large in various contemporary accounts of machine learning as a way of knowing (for instance, in popular machine learning  books such as _Machine Learning for Hackers_ [@Conway_2012] or _Doing Data Science_ [@Schutt_2013]). The machine learning research published in statistics, computer science, mathematics, artificial intelligence and a swathe of related scientific fields bristles with references to decision trees and support vector machine, as well as neural networks. (As we will see, these techniques themselves ingest substantial supplies of knowledge and practice produced by  social sciences, insurance and actuarial practice, and marketing research.) The top 20 most cited publications in the field include  Ross Quinlan and Leo Breiman's papers on decision trees [@Quinlan_1986; @Breiman_1984], Vladimir Vapnik and Corinna Cortes' support vector machines papers [@Vapnik_1999; @Cortes_1995],  an early textbook written by a computer scientist on machine learning [@Mitchell_1997], a textbook and software package on data mining using Java [@Witten_2005]; a textbook on pattern recognition dating from the 1970s [@Duda_2012], a tutorial on an error control technique (ROC - Receiver Operating Characteristics, first developed by the US military during WWII) and somewhat lower, another well-known textbook, this time on neural networks and pattern recognition [@Bishop_2006]. 

The re-configuration of tensions between sparsity and accumulation, between rarity and commonality can be understood in the sense of *positivity* proposed by Michel Foucault in _The Archaeology of Knowledge_ [@Foucault_1972], a book whose somewhat a-subjective approach to knowledge unexpectedly resonates with much that takes place in machine learning. A striking _positivity_ takes shape across these differences. \index{positivity} (I understand positivity partly in the sense of 'positivism,' an epistemological stance widely shared amongst machine learners  that views knowledge as deriving from observation.[^5.01])  Foucault theorises statements as elements _posited_ or deposited in a relatively uneven space of transformations. These transformations generate what can be thought and done within a particular discursive formation. Their positivity arises from the 'forms of an accumulation':

>To describe a group of statements not as the closed, plethoric totality of a meaning, but as an incomplete, fragmented figure; to describe a group of statements not with reference to the interiority of an intention, a thought, or a subject, but in accordance with the dispersion of an exteriority; to describe a group of statements, in order to rediscover not the moment or the trace of their origin, but the specific forms of an accumulation, is certainly not to uncover an interpretation, to discover a foundation, or to free constituent acts; nor is it to decide on a rationality, or to embrace a teleology. It is to establish what I am quite willing to call a positivity [@Foucault_1972, 125]. \index{Foucault, Michel!on positivity} \index{Foucault, Michel!on statements}  

Foucault frequently refers to positivity in those passages where he discusses  how practices develop and how statements relating to those practices take shape. The situations in which he invokes positivity involve accumulations or plethora of statements whose coherence or unity does not lie in any origin, intention, rationality, subjectivity, ideology or teleology. Like mineral strata, positivities entail accumulations without deriving them from underlying unities. Foucault's acceptance of 'positivity,' in spite of the risk of attracting criticism, pivots on his interest in, and practical commitment to, understanding accumulations and uneven distributions. A positivity is a way of describing 'a group of statements' that may be quite dispersed, discontinuous, and unevenly distributed because they make up 'a population of events in the space of discourse in general' [@Foucault_1972, 27]. If _The Archaeology of Knowledge_ is a book concerned with the tissue of relations between things, what is said, what is done and what counts as knowledge (scientific or not) in practice, then the persistent and at times forceful reiteration of _dispersion_ and _discontinuity_ might seem strangely counterintuitive. But Foucault's insistence on discourse as something irreducible to logic, language, intention, rationality, experience, historical origin or narrative, and as the way in which 'statements' ('a graph, a growth curve, an age pyramid, a distribution cloud are all statements' (82)), strategies, concepts, objects, archives and knowledges come together in 'discursive formations' is an attempt to resist any resort to a hidden order or concealed origin that critical thought could uncover or interpret. \index{positivity!accumulation in} \index{statement!as diagram}

Positivity as accumulation helps make sense of a field such as machine learning whose many different forms, techniques, statements, institutions, values and associations often grapple with problems of much data, many variables, many possible models and a sense of an elusive but vital order or pattern discoverable amidst accumulation. Rather than seeing pattern as something discovered in data, the notion of positivity suggests we should examine the diagrammatic operations that configure the practice of machine learning, giving rise to a field of objects and subject positions.   The three machine learners that anchor this chapter are at once perhaps the most distinctive data mining, pattern recognition and predictive modelling achievements of the late twentieth century (at least judging by the citational and implementational interest they attract). They different greatly in how they move through data. At certain times, they come together (for instance, in machine learning competitions discussed in chapter \ref{ch:subjects}; or in certain formalizations such as machine learning  theory or in graphs of the bias-variance decomposition discussed in chapter \ref{ch:probability}; or in the pedagogy of machine learning discussed in chapter \ref{ch:diagram}). We might understand their differences in terms of 'enunciative modalities' [@Foucault_1972, 54] \index{statement!enunciative modality of}  -- that is, in terms of the different ways in which they give rise to statements. Every  machine learner generates statements, but from different places, by somewhat different individuals, and from the different situations they that are able 'to occupy in relation to the various domains or groups of objects' [@Foucault_1972, 52].[^5.02]

# Splitting or recursive binary partitioning and the growth of trees

> Mastering the details of tree growth and management is an excellent way to understand the activities of learning machine generally [@Malley_2011, 118].

Decision trees promise an understanding of machine learning.   The enunciative function of the decision tree partly relates to the observability and comprehensibility of machine learning. \index{statement!enunciative function of} As we will see, not all machine learners readily support observation or comprehension. The cost of comprehensibility, however, is a certain highly restricted movement through the data, a movement that although simple and easily understandable, has been difficult to stabilise and control. As _Elements of Statistical Learning_ puts it: 'tree-based methods partition the feature space into a set of rectangles, and then fit a simple model (like a constant) in each one. They are conceptually simple yet powerful' [@Hastie_2009, 305].

\begin{figure}
  \centering
      \includegraphics[width=0.9\textwidth]{figure/recursive_partitioning.pdf}
        \caption{Recursive partitioning of the feature space}
  \label{fig:rpart}
\end{figure}

Tree-based methods are supervised learners as they require the data to either be labelled with a class or to have some outcome value. \index{machine learning!supervised} The variable types in the feature space (or common vector space of the data) \index{feature space|see{common vector space}} can be mixed. Because the method cuts the common vector space into a tiled surface, the features or data variables can be continuous or discontinuous. The 'simple models' that tree methods construct each inhabit one of the rectangular regions or partitions of the feature space. In Figure \ref{fig:rpart}, the different regions or partitions produced by a decision tree are labelled $R1, R2$ etc. 

```{r aid, echo=FALSE, results='asis', cache=TRUE } 
    q = "select TI, PY, TC from basic_refs where anchor = 'morgan_sonquist_WOS'"
    res = dbGetQuery(con, q)
    res$TI = sapply(res$TI, simpleCap) 
    colnames(res) = c('Title', 'Year', 'Citations')
    tab  = xtable(res[1:10,], caption="References to Morgan and Sonquist's Automatic Interaction Detector", label='table:aid_citation')
    print(tab, type='latex', include.rownames=FALSE)
   
```

Work on classification and regression techniques using decision trees goes back to the  early 1960s when social scientists James Morgan and John Sonquist at the University of Michigan's Institute for Social Research were attempting to analyse increasingly large social survey datasets [@Morgan_1963].  As Dan Steinberg describes in his brief history of decision trees [@Steinberg_2009, 180], the  'automatic interaction detector' (AID) as it was known, sought to automate the practice of data analysts looking for interactions between different variables. The variety and sheer optimism of subsequent applications of these prototype decision tree techniques  is striking.  In the 1960s and 1970s, papers that drew on the AID paper or use AID techniques can be found, as table \ref{table:aid_citation} shows, in education, politics, economics, population control, advertising, mass media and family planning. \index{machine learner!decision tree!history of} \index{machine learner!automatic interaction detector}

\begin{figure}
  \centering
      \includegraphics[width=0.9\textwidth]{figure/morgan_sonquist_p430.pdf}
        \caption{AID classifies annual earnings (Morgan \& Sonquist, 1963, 430)}
  \label{fig:aid_tree}
\end{figure}

A decade after initial work, the AID was the object of trenchant criticism by statisticians and others, not for the classifications it used (see Figure \ref{fig:aid_tree}), but for its dangerous empiricism. Writing in the 1970s, statisticians in the behavioural sciences such as Hillel Einhorn at the University of Chicago castigated the use of such techniques.   The criticisms stemmed from  a general distrust of 'purely empirical methods', and scepticism focused on their positivity:

> The purely empirical approach is particularly dangerous in an age when computers and packaged programs are readily available, since there is temptation to substitute immediate empirical analysis for more analytic thought and theory building. It is also probably too much to hope that a majority of researchers will take the time to find out how and why a particular program works. The chief interest will continue to be in the output-the results-with as little delay as possible [@Einhorn_1972, 368]

Einhorn discusses AID alongside other  techniques such as factor analysis and multi-dimensional scaling (both still widely used) before concluding 'it should be clear that proceeding without a theory and with powerful data analytic techniques can lead to large numbers of Type I errors' [@Einhorn_1972, 378]. His statistical objections to AID are  particularly focused on the problematic power of the technique: 'it may make sense out of "noise"' (369). Consequently, researchers easily misuse the technique: they 'overfit' the data, and do not pay enough attention to issues of validation (369-370). Similarly the British marketing researcher Peter Doyle, criticising the use of AID in assessing store performance and site selection by operations researchers, complained that searching for patterns in data using data sets was bound to lead to spurious results and the decision trees, although intuitively appealing (that is, they could be easily interpreted), were afflicted with arbitrariness: 'a second variable may be almost as discriminating as the one chosen, but if the program is made to split on this, quite a different tree occurs' [@Doyle_1973, 465-466].[^5.03] \index{operations research!use of decision trees} \index{machine learner!overfitting} Most of these criticisms can be seen as expressing conventional statistical caution in response to threats to validity. 

# 1984: Cutbacks on recursive partitioning

As Einhorn expected, it was too much to hope that all researchers would take time to investigate how a particular program works. Some researchers however did take time, years in fact, to investigate how decision trees work.  Writing around 2000, Hastie, Tibshirani and Friedman, who could hardly by accused of not understanding decision trees,  happily recommend decision trees as the best 'off-the-shelf' classifier:  'of all the well-known learning methods, decision trees comes closest to meeting the requirements for serving as an off-the-shelf procedure for data-mining' [@Hastie_2009, 352]. We might wonder here, however, whether they damn with faint praise, since 'off-the-shelf' suggests pre-packaged, and commodified, and the term 'data-mining' itself is not without negative connotations. As for its commercial realities, in 2013, Salford Systems, the purveyors of the leading contemporary commercial decision tree software, `CART`, could claim:

 > CART is the ultimate classification tree that has revolutionized the entire field of advanced analytics and inaugurated the current era of data mining. CART, which is continually being improved, is one of the most important tools in modern data mining. Others have tried to copy CART but no one has succeeded as evidenced by unmatched accuracy, performance, feature set, built-in automation and ease of use. [Salford Systems](http://www.salford-systems.com/products/cart) \index{machine learner!CART}

What happened between 1973 and 2013?  Decision trees somehow grew out of  the statistically murky waters of social science departments and business schools in the early 1970s to inaugurate the 'current era of data mining' (which the scientific literature indicates starts in the early 1990s). \index{data mining!in 1970} This was not only a commercial innovation. As the earlier citation from U.S. National Institutes of Health biostatisticians Malley, Malley and Pajevic indicates, decision trees enjoy high regard even in biomedical research, a setting where statistical rigour is highly valued for life and death reasons. The happy situation of decision trees four decades on suggests some kind of threshold was crossed in which the epistemological, statistical, or algorithmic ('built-in automation') power of the technique altered substantially. \index{machine learner!decision tree!in medicine}

The third author of _Elements of Statistical Learning_, Jerome Friedman, worked at the  U.S. Department of Energy's Stanford Linear Accelerator during the late 1970s. Friedman  was  instrumental in rescuing  decision trees from the ignominy of profligate ease of use and pure empiricism they had endured since the late  1960s. \index{Friedman, Jerome!work on decision tree} The reorganisation and statistical retrofitting of the decision tree was not a single or focused effort. During the 1980s, statisticians such  as Friedman and Leo Breiman renovated the decision tree as a statistical tool [@Breiman_1984] at the same time as  computer scientists such as  Ross Quinlan in Sydney were re-implementing decision trees guided by an  artificial intelligence-based formalisation as rule-based induction technique [@Quinlan_1986].[^5.1] \index{Quinlan, John Ross!} \index{artificial intelligence!rule-based induction} This uneasy parallel effort between computer science and statistics  still somewhat strains relations in machine learning today. Both statisticians and computer scientists  do and use the same techniques, but often with the computer scientists focusing on optimisation and algorithmic scale and the statisticians inventing novel statistical formalizations and abstractions. The fateful embrace of statistics and computer science, the disciplinary binary that vectorizes machine learning, has been generative in the retrieval of the decision tree. \index{vectorization}

An initial symptom of the transformation of the technique appears in a name change. The term 'decision tree,' although still widely used in the research literature and machine learner  parlance was replaced by 'classification and regression tree' during the late 1970s and 1980s. The terms 'classification and regression tree' is sometimes contracted to 'CART,' and that term strictly speaking refers to a computer program described in [@Breiman_1984] as well as the title of that highly-cited monograph. \index{machine learner!CART} \index{Breiman, Leo!CART monograph} As we have seen in previous chapters, regression and classification designate the two main sides of machine learning practice, and their concatenation with 'tree' attests to the ongoing renovation of existing machine learning approaches with a more statistical facade.

The implementation of machine learning techniques in `R`  by virtue of `R`'s statistical provenance perhaps favours the statistical side of decision tree practice, but that has certain forensic virtues not offered by commercial or closed-source software often produced by computer scientists.  In this case, the name of one long-standing and widely-used `R` package itself attests to something: `rpart` is a contraction of 'recursive partitioning' and this term generally describes how the decision tree algorithm works to partition the common vector space into the form shown in Figure \ref{fig:rpart} [@Therneau_2015].  'CART,' on the other hand, is a registered trademark of Salford Systems, the software company mentioned above, who sell the leading commercial implementation of classification and regression trees. Hence, the `R` package `rpart` cannot call itself the more obvious name `cart, ` and instead invokes the algorithmic process it relies on: recursive partitioning.[^5.06] \index{algorithm!recursive partitioning} \index{programming languages!R!packages!rpart}

```{r iris_tree, echo=TRUE, cache=TRUE, message=FALSE, warning=FALSE, comment=NA, size='smallsize', results='asis' } 

    data(iris)
    library(rpart)
    iris_tree =rpart(Species ~ ., iris)
```

R.A. Fisher's _iris_ dataset, which contains measurements made in the 1930s of petal and sepal lengths of _iris virginica, iris setosa_ and _iris versicolor_ is a standard instructional example for decision trees [@Fisher_1938].[^5.12] \index{dataset!iris} The code shown here loads the _iris_ data (the dataset  is routinely installed with many data analysis tools), loads the `rpart` decision tree library, and builds a decision to classify the irises by species.  What has happened to the iris data in this decision tree? The `R` code that invokes the recursive partitioning algorithm is so brief `iris_tree =rpart(Species ~ ., iris)` that we can't tell much about how the data has been 'recursively partitioned.' We know that the _iris_ has `r nrow(iris)` rows, and that there are equal numbers of the three iris varieties. 

Code brevity indicates a great deal of formalization of practice has accrued around decision trees. \index{code!brevity of!} Some of this formalization was described in the landmark _Classification and Regression Trees_ monograph [@Breiman_1984], and is familiar (from previous chapters)  in  its use of common vector space, functions as partial observers and the re-distribution of chance between devices and worlds. For instance, Breiman and co-authors start out by placing the technique in vector space:

> Define the measurements $(x_1, x_2, ...)$ made on a case as the _measurement vector_ $\mathbf{x}$ corresponding to the case. [@Breiman_1984, 3]

In re-configuring decision trees as classification trees, Breiman first addresses issues of data dimensionality. \index{data!dimensionality} Next,  rather than simply invoking the notion of a classifier as a device (as AID did), the CART monograph defines a classifier in terms of a function: 

> A classifier or classification rule is a function $d(\mathbf{x})$ defined on $\mathcal{X}$ so that for every $\mathbf{x}$, $d(\mathbf{x})$ is equal to one of the numbers $1, 2, ..., J$. [@Breiman_1984, 4]

Note that 'rule,' the term that computer scientists working in artificial intelligence at the time were using, still figures here, but the classifier is formalized as a function. \index{machine learner!function}  The definition of the classifier as a function depends on the existence of a vector space $\mathcal{X}$, a set of measurement, feature or predictor vectors  $\mathbf{x}$ and a set of responses  $1, 2, ..., J$. Second, classification is understood in terms of a series of binary splits in the different dimensions of the common vector space. These splits comprise the tree structure diagram. The problem here is that many splits are possible. 

 > The first problem in tree construction is how to use $\mathcal{L}$ to determine the binary splits of $\mathcal{X}$ into smaller pieces. The fundamental idea is to select each split of a subset so that the data in each of the descendant subsets are "purer" than the data in the parent subset [@Breiman_1984, 23].

Tree construction hinges on the notion of purity or more precisely 'node impurity', a function that measures how data labelled as belonging to different classes are mixed together at a given branch or node in a decision tree: 'that is, the node impurity is largest when all classes are equally mixed together in it, and smallest when the node contains only one class' [@Breiman_1984, 24]. As Malley and co-authors note, 'the collection of purity measures is still a subject of research' [@Malley_2011, 123], but Breiman, Friedman, Olshen and Stone promoted a particular form of impurity measure for classification trees known as 'Gini index of diversity' [@Breiman_1984, 38].  \index{difference!Gini index of diversity} Like the planar decision surface used in machine learners based on linear regression models, recursive partitioning combined with measures of node impurity allow tree construction to be understood as a form of cutting, division or classification \index{learning!as dividing}  Whereas in linear model-based machine learners, the intuition motivating the function-finding or learning was 'find the line that best fits or separates the data, index{machine learner!logistic regression!gradient descent} here the intuition is more like: 'split things in ways that minimize mixing'. Good splits decrease the level of impurity in the tree. In an tree with maximum purity, each terminal node -- the nodes at the base of the tree -- would contain a single class. 

The results of the application of measures of node impurity can be seen below in two plots. 

```{r iris_tree_plot, echo=FALSE,  fig.cap ='Decision tree on _iris_ dataset', cache=TRUE, message=FALSE, warning=FALSE, comment=NA, results='asis' } 

    par(mfrow = c(1,2), xpd=NA)
    plot(iris_tree, main = 'tree recursive partitioning of iris')
    text(iris_tree, cex=0.8, use.n=TRUE)
    table(iris$Species) # is data.frame with 'Species' factor
     iS <- iris$Species == "setosa"
     iV <- iris$Species == "versicolor"
     op <- par(bg = "bisque")
     matplot(c(1, 8), c(0, 4.5), type =  "n", xlab = "Length", ylab = "Width",
             main = "Petal and Sepal Dimensions in Iris Blossoms")
     matpoints(iris[iS,c(1,3)], iris[iS,c(2,4)], pch = "sS", col = c(2,4))
     matpoints(iris[iV,c(1,3)], iris[iV,c(2,4)], pch = "vV", col = c(2,4))
     legend(1, 4, c("    Setosa Petals", "    Setosa Sepals",
                    "Versicolor Petals", "Versicolor Sepals"),
            pch = "sSvV", col = rep(c(2,4), 2))
```

\begin{figure}
  \centering
      \includegraphics[width=0.9\textwidth]{figure/iris_tree_plot-1.pdf}
        \caption{Decision tree on \verb{iris} dataset}
  \label{fig:iris_tree_plot}
\end{figure}

In Figure \ref{fig:iris_tree_plot}, the plot on the left shows the decision tree and the plot on the right shows just _setosa_ and _versicolor_  plotted by petal and sepal widths and lengths.  As the plot on the right shows, most of the measurements are well clustered. Only the _setosa_ petal lengths and widths seem to vary widely. All the other measurements are tightly bunched. This means that the decision tree shown on the left has little trouble classifying the irises. Decision trees are read from the top down, left to right. The top level of this tree can be read, for instance, as saying, if the length of petal is less 2.45, then the iris is _setosa_. \index{dataset!iris}

Like logistic regression models, neural networks, support vector machines or any other machine learning technique, decision trees make sense of the worlds they encounter in terms of specific qualities and logics. Recursive partitioning can split and sub-divide the common vector space to capture every minor difference between cases, and thereby achieve a perfect fit to the data. \index{algorithm!recursive partitioning} Although the partitioning or splitting rules have strong statistical justifications, they do not at all eliminate the problem of instability or variance in trees. \index{error!variance}  For instance, they easily end up 'overfitting' the data. Overfitting is a problem for all machine learning techniques. Algorithms sometimes find it hard to know when to stop. During construction of a decision tree, various features in the data are split into smaller and smaller groups. 'The goodness of the split', wrote Breiman and co-authors, 'is defined to be the decrease in impurity' [@Breiman_1984, 25]. Under this definition of goodness, the terminal nodes or leaves of the tree can end up containing a single case, or a single class of cases. The decision tree targets the singularity of the individual case to such a degree that it sees categorical differences everywhere. Driven to maximise the purity of the partitions it creates, it leans too heavily on  data it has been trained on to see relevant similarities when fresh data appears. Trees that branch too much are too sensitive (or unstable), and need to be pruned. Such a model will almost always _overfit_, since slight variations in the values of variables in a fresh case are likely to yield widely differing predictions. In the terminology of machine learning, such a decision tree will have low bias but high variance. \index{error!overfitting}

Much of the development of decision tree did not revolve around how to construct them, but how to limit their growth. An associated literature on _pruning_ decision trees using measures of tree complexity has also developed. As Hastie and co-authors put the problem in their account of classification and regression trees:

>How large should we grow the tree? Clearly a very large tree might overfit the data, while a small tree might not capture the important structure. Tree size is a tuning parameter governing the model’s complexity, and the optimal tree size should be adaptively chosen from the data. One approach would be to split tree nodes only if the decrease in sum-of-squares due to the split exceeds some threshold. This strategy is too short-sighted, however, since a seemingly worthless split might lead to a very good split below it. The preferred strategy is to grow a large tree $T_0$, stopping the splitting process only when some minimum node size (say 5) is reached. Then this large tree is pruned using cost-complexity pruning [@Hastie_2009, 307-308].

Growing a maximum decision tree, and then cutting back its branches using a cost-function is a further renovation of the decision tree device as a machine learner. The 'cost complexity pruning' reproduces the optimisation we have already discussed in relation to linear regression and logistic regression models. As in these techniques, the definition of a cost function controlling the 'complexity' of a tree -- how many branches and leaves/nodes it contains, combined with measures of how well it classifies or predicts -- opens up the possibility of iteratively testing different versions of tree against each other. 'We define the cost complexity criterion,' write Hastie and co-authors, as:

\begin {equation}
\label {eq:tree_model}
C_\alpha(T) = \sum_{m=1}^{|T|}N_mQ_m(T) + \alpha|T|
\end {equation}

The idea is 'to find, for each $\alpha$, the subtree $T_\alpha \subseteq  T$ 0 to minimize $C_\alpha(T)$' [@Hastie_2009, 308]. For present purposes, we need only recognise that the cost complexity function re-configures decision trees ($T$ in equation \ref{eq:tree_model})  as optimisation problems, in which the variation of a parameter ($\alpha$) allows minimization of a derived value (the cost $C_\alpha$).  \index{function!cost}

The shift from AID to CART enunciates a change in how patterns become visible. The decision tree algorithm combines recursive partitioning and cost-complexity pruning to configure a different mode of enunciation, a 'specific form of accumulation' [@Foucault_1972, 125]. When they cross such thresholds, machine learners create a new _referential_, a new field of rules of differentiation of individuals, facts, things and relations.[^5.101] \index{referential} The new referential -- the combination of purity and density that comes from recursive partitioning and cost-complexity pruning -- re-defines what counts as pattern. Hence, classification trees stand in a different relation to classification, and prediction than their predecessor diagrams. While the graphic form of the decision tree was, by virtue of the long-standing diagrammatic practice of tree-drawing, easy to interpret, observation of decision trees had no way of gauging the instability or variability of any given tree. \index{diagram!decision tree as} Hastie and co-authors write: 'one major problem with trees is their high variance. Often a small change in the data can result in a very different series of splits, making interpretation somewhat precarious. The major reason for this instability is the hierarchical nature of the process: the effect of an error in the top split is propagated down to all of the splits below it [@Hastie_2009, 312]. The very diagrammatic form that allows decision trees to be observed and interpreted is also the source of their instability.  

If decision trees have been heavily used in credit risk assessment as well as many biomedical models, does this stem from the visibility of the patterns they produce, even if those patterns are unstable? Or is the success of the decision tree perhaps better understood as a change in the referential underpinning of patterns more generally, of which decision trees would only be one instance among many? \index{enunciative modality} If we understand machine learners as enunciative modalities (remembering that the concept of an enunciative modality concerns the dispersion and discontinuities that surround, support and actually generate statements),   the transformation and re-modelling of the decision tree as classification and regression trees suggests a subtle, non-localizable discontinuity. We could follow here the subsequent development of the decision tree and its subsequent transmogrification into random forests [@Breiman_2001], \index{machine learner!random forest} that grow a myriad of small decision trees. Their growth and inter-comparison disperses kaleidoscopic fragments of classificatory order without unifying pattern. They intensify the 'discontinuity of the planes' from which machine learners speak. [^5.102] In such developments -- and we could also consider here techniques, models and methods of 'boosting,' 'bagging,' or 'ensemble learning' that conducts 'supervised search in a high-dimensional space of weak learners' [Hastie_2009, 603], pattern has an operational rather than visible mode of togetherness. \index{pattern!operational}

# The successful dispersion of the support vector machine

Already in the growing and pruning of the decision tree, and even more markedly in support vector machine and neural networks, a different kind of enunciative function can be discerned in which patterns plays out in dispersion and discontinuity rather than in regular geometry. \index{pattern!in dispersion} Take the case of the support vector machine. The second most highly cited reference in the last few decades of machine learning literature is a paper from 1995 by Corinna Cortes and Vladimir Vapnik of AT & T Bell Labs in New Jersey, USA entitled 'Support Vector Networks' [@Cortes_1995].  Few women's names have appeared prominently in the machine learning literature. The computing science and statistics departments at Stanford and Berkeley, the laboratories at Los Alamos and AT&T Bell between the 1960s and the 1980s were, it seems, not overly popular or populated with women scientists and engineers. Some prominent machine learning researchers at the time of writing are women (I return to this in chapter \ref{ch:subjects}), but Cortes is perhaps pre-eminent both as head of Google Research in New York (2014) and as recipient with Vapnik of an Association for Computing Machine award in 2008  for work on the support vector machine algorithm.[^5.70] \index{Cortes, Corinna} \index{Vapnik, Vladimir} \index{support vector machine|see{machine learner!support vector machine}}


```{r pattern_recognition1, echo=FALSE, cache=TRUE, message=FALSE, warning=FALSE, comment=NA, size='smallsize', results='asis' } 
    query = 'select TI, PY, TC from basic_refs where TI like "%support vector machine%" or anchor = "svm_WOS"'
    res = dbGetQuery(con, query)
    svm = res[order(res$TC, decreasing=TRUE),]
    colnames(svm) = c('Title', 'Year', 'Citations')
    tab = xtable(head(svm,20), label = 'tab:svm_lit', align = c("p{0.05\\textwidth}",  "p{0.8\\textwidth}","p{0.1\\textwidth}",  "p{0.1\\textwidth}", "p{0.15\\textwidth}"), caption = 'Most cited papers on support vector machines')
    print(tab, type='latex', include.rownames=FALSE)
```

The rapid rise to popularity of the support vector machine can be seen in the machine learning research literature, very small slice of which appears in Table \ref{table:svm_lit}. A substantial fraction of the overall research publication since the mid-1990s accumulates around this single technique, and as usual ranges across credit analysis, land cover prediction, protein structures, brain states and face recognition. The support vector machine covers the normal biopolitical gamut of life, labour and language. The influence of the technique  can also be seen in  adjacent and overlapping fields such as pattern recognition and data mining, where [@Cortes_1995] and similar papers rank near the top-cited papers.[^5.112]  This kind of growth betokens high levels of interest, identification and investment on the part of the researchers, and presumably more widely. What is it about this technique?

```{r svm-demo, echo=FALSE, cache=TRUE, message=FALSE, warning=FALSE, comment=NA, engine='python' } 
import numpy as np
import pylab as pl
import seaborn
from sklearn import svm, datasets

iris = datasets.load_iris()  
X = iris.data[:, :2] 
Y = iris.target

h = .02  # step size in the mesh

C = 1.0  # SVM regularization parameter
svc = svm.SVC(kernel='linear', C=C).fit(X, Y)
rbf_svc = svm.SVC(kernel='rbf', gamma=0.7, C=C).fit(X, Y)
poly_svc = svm.SVC(kernel='poly', degree=3, C=C).fit(X, Y)
lin_svc = svm.LinearSVC(C=C).fit(X, Y)

x_min, x_max = X[:, 0].min() - 1, X[:, 0].max() + 1
y_min, y_max = X[:, 1].min() - 1, X[:, 1].max() + 1
xx, yy = np.meshgrid(np.arange(x_min, x_max, h),
                     np.arange(y_min, y_max, h))

titles = ['SVC with linear kernel',
          'SVC with RBF kernel',
          'SVC with polynomial (degree 3) kernel',
          'LinearSVC (linear kernel)']


for i, clf in enumerate((svc, rbf_svc, poly_svc, lin_svc)):
    pl.subplot(2, 2, i + 1)
    Z = clf.predict(np.c_[xx.ravel(), yy.ravel()])
    Z = Z.reshape(xx.shape)
    pl.contourf(xx, yy, Z, cmap=pl.cm.Paired)
    pl.axis('off')
    pl.scatter(X[:, 0], X[:, 1], c=Y, cmap=pl.cm.Paired)
pl.title(titles[i])

pl.savefig('figure/contour.pdf')
```

\begin{figure}
  \centering
      \includegraphics[width=1.0\textwidth]{figure/contour.pdf}
        \caption{SVM on Anderson's iris dataset}
  \label{fig:svm_iris}
\end{figure}

The shift from decision trees to classification tree (and then random forests) illustrate an enunciative modality anchored in a dispersed referential. \index{referential!dispersed} The support vector machine shown in Figure \ref{fig:svm_iris} demonstrates perhaps more strikingly how this dispersed referential changes what counts as pattern. The typical regions of the different sub-graphs have different contours, contours that project a much more dispersed referential. While the name is somewhat forbiddingly technical compared to more familiar terms such as 'decision tree' or even 'neural network,' the underlying intuition of the technique is much older, and can be found in the models developed by the British statistician R. A. Fisher during the 1930s. R.A Fisher developed the 'first pattern recognition algorithm' [@Cortes_1995, 273], the 'linear discriminator function' [@Fisher_1936],  to deal with problems of classification, and demonstrated its efficacy on the taxonomic problem of  discriminating or classifying the irises observed in W.E. Anderson's irises in `iris` dataset (see above). \index{dataset!iris} \index{function!linear discrimant} In his 1936 article in the _Annual Review of Eugenics_, Fisher comments on similar classification work carried out in craniometry and other related settings: so-called 'discriminant functions' had been successfully used to distinguish populations. Fisher wrote: 'when two or more populations have been measured in several characters, ... special interest attaches to certain linear functions of measurements by which the populations are best discriminated' [@Fisher_1936, 179]. \index{Fisher, Ronald Ayre}  The discriminant functions divide the common vector space into 'a collection of regions labeled according to classification' [@Hastie_2009, 101]. 'Decision boundaries' (or sometimes 'decision surfaces' \index{classification!decision boundary} often appear as straight lines that divide the common vector space into regions of constant classification. \index{function!discriminant}  The route by which these long-standing discriminant functions were reconstructed during the 1990s in the form of the support vector machine highlights shifts in practice that give rise to new statements about data, statements that can be glimpsed in table \ref{table:svm_lit} in the range of 'referents,' things, facts and beings running through the titles of the papers. 


# The decision boundary blurs?

The linear decision boundary changes in two ways in support vector machines. It blurs and bends. \index{pattern!dispersion of}  Nearly all expositions of the support vector machine including [@Cortes_1995]  highlight the 'soft margin' that runs in parallel to the solid decision boundary. The support vector machine addresses the problem of how to model differences when differences are blurred. An oft-repeated illustration of how the support vector machine traverses data appears in Cortes and Vapnik's initial publication simply entitled 'Support Vector Networks' [@Cortes_1995]. They demonstrate how the support vector machine classifies handwritten digits drawn from a dataset supplied by the US Postal Service [@LeCun_2012]. Like `iris`, the US Postal Service digits and a larger version from the US National Institute of Standard (NIST) are  standard machine learning dataset. They have been frequently used to measure the performance of competing learning algorithms. \index{dataset!U.S. Postal digits|see{MNIST}}  In contrast to _iris_, the US Postal Service Database is high dimensional. Each digit in the dataset is stored as 16x16 pixel element. Image classification typically treats each pixel as a feature  or variable in the input space. So each digit as represented by a 16x16 pixels amounts to a 256 dimensional input space. By comparison, `iris` has five dimensions. Unsurprisingly, there are also many more digits in the US Postal Service Database than in flowers in _iris_. The NIST dataset has around 70,000. Aside from this dimensional growth, the handwritten digits aptly convey the non-linearity of the classification problem. On the one hand, human can recognise handwritten digits fairly easily and with few errors. This is despite the many variations in handwriting that skew, morph and distort the ideal graphic forms of numbers.[^5.23]  


\begin{figure}
  \centering
      \includegraphics[width=0.9\textwidth]{figure/mnist_test6.jpg}
        \caption{MNIST Postal Digits}
  \label{fig:postal_digits}
\end{figure}

In their experiments with digit recognition (shown in figure \ref{fig:postal_digits}, Cortes and Vapnik contrast the error rates of decision trees (CART and C4.5), neural networks and the support vector machine working at various level of dimensionality. This occurs in two different ways. On the one hand, the support vector machine develops Fisher's linear discriminant analysis since it searches for a separating hyperplanes in the data. \index{hyperplane|see{decision surface}} While  linear discriminant analysis constructs that hyperplane by finding the most likely boundary between classes based on all the data, the support vector machine searches for a hyperplane resting only on those cases in the data that lie near the boundary. It introduces the intuition that the best hyperplane separating different classes will run near to the cases -- the _support vectors_ --  that are most difficult to classify. These hard-to-classify cases become the support vectors whose relative proximities tilt the decision surface in various directions. In contra-distinction to the $N=\forall{X}$ proposition (discussed in chapter \ref{ch:probability}), the machine learner discards much of the data. As Alpayadin writes, 'we do not care about correctly estimating the densities [probability distributions of variables] inside class regions; all we care about is the correct estimation of the _boundaries_ between the class regions' [@Alpaydin_2010, 210]. \index{Alpayadim, Ethem!on decision boundaries} \index{machine learner!linear discriminant analysis}

```{r svm_margins, cache=TRUE, dev='pdf'}
library(kernlab)
x = rbind(matrix(rnorm(100,),,2), matrix(rpois(100, 0.3),,2))
c = matrix(c(rep(-1, 50), rep(1, 50)))
svp = ksvm(x, c, type='C-svc')
plot(svp, data=x)
w <- colSums(coef(svp)[[1]] * x[unlist(alphaindex(svp)),])
b <- b(svp)
abline(b/w[1],-w[2]/w[1])
abline((b+1)/w[1],-w[2]/w[1],lty=2)
abline((b-1)/w[1],-w[2]/w[1],lty=2)
text(x=2, y=2,cex=0.7, labels='support vector')
```

\begin{figure}
  \centering
      \includegraphics[width=0.9\textwidth]{figure/svm_margins-1.pdf}
    \caption{Margins in a support vector machine}
  \label{fig:svm_margins}
\end{figure}

Figure \ref{fig:svm_margins} has appeared in many slight variations in the last two decades. Such figures diagram classes by different point shapes, and the diagrammatic work of the classifier takes the shape of diagonal lines, the solid line marking the decision surface or hyperplane and the dotted lines marking the soft margins that separate the two classes. In Figure \ref{fig:svm_margins}, the dotted lines represent a margin on either side of a hyperplane (the solid line).  The support vector machine finds the hyperplane for which that margin or perpendicular distance between the margins is greatest. Of all the slightly different planes that might run between the two classes shown in that figure, the maximum separating hyperplane lies at the greatest distance from all the points of the different classes. The support vector classifier modifies the idea of the optimal separating hyperplane by allowing classification of inseparable or overlapping classes.

While the geometrical intuition here is that some cases will be allowed to lie on the opposite side of the decision surface to where they should be, the distance they lie on the wrong side of the separating hyperplane will be as small as possible. How are the lines showing in Figure \ref{fig:svm_margin} calculated? Combining the optimal separating hyperplane and a limited number of permitted misclassifications entails a complicated optimisation problem. As Hastie and co-authors, following Cortes and Vapnik's formulations,  formalize it, the problem can be stated as a minimization problem:

\begin {equation}
\label {eq:lagrange_primal}
L_P = \frac{1}{2}\parallel\beta\parallel^2 + C\sum_{ i=1}^{N}xi_i - \sum \alpha[y_{i}(x_i^T\beta + \beta_0) - (1-xi_i)]- \sum_{i=1}^{N}\mu_ixi_i
\end {equation}
[@Hastie_2009, 420]

In equation \ref{eq:lagrange_primal}, the optimisation problem is to minimize $L_P$, the Lagrange primal function with respect to $\beta, \beta_0$ and $x_i$ [@Hastie_2009, 420]. In this complicated optimisation problem (one that is difficult to understand without extensive mathematical background), the key elements are the now familiar $\beta$ parameters in the linear form of $x_i^T\beta + \beta_0$, which is the equation defining the hyperplane,  as well as the multiple occurrences of sums ($\sum$) of all the values $x_i$, which calculate the distance that each case is on the wrong side of the margin. Correctly classified cases will, therefore, have $x_i =0$. As always with these mathematical propositions, their diagrammatic relations, and the way in which they contain both the generalizating regularities (the linear equation, the repeated summation of all values, the shaping parameters) and forms of variation (the presence of the misclassification measures $x_i$) should be the focus of our attention. What if  elements that lie on the wrong side of the hyperplane were allowed? If that were possible, the support vector machine could deal with in-separable or overlapping classes. As Hastie and co-authors put it in _Elements of Statistical Learning_, 'hence we are able to find the hyperplane that creates the biggest margin between the training points for class 1 and -1' [@Hastie_2009, 418]. The support vector machine permits instances that lie on the wrong side of the separating hyperplane. They function not as errors (as they would appear in most linear classifiers such as linear discriminant analysis and logistic regression), but as components of a 'soft margin' designed to accommodate inseparability and indistinctness. Equations such as equation \ref{eq:lagrange_primal} connect the diagrammatic intuition of a separating hyperplane with a set of movements controlled by parameters such as $C$, which effectively controls the size of the margin, and $\alpha$, which effectively bounds the proportion by which a predicted instance can be on the wrong side of the margins that define the hyperplane. In other words, as we have seen previously in cost-function optimisation (see chapter \ref{ch:function}), the learning done by the machine consists in finding a way of moving according to these constraints, whose shape has already been inscribed by the functions entering into the diagram. \index{diagram!forms of movement} 

# Bending the decision boundary

Substitutions and alterations of the common vector space affect the decision boundary in a second sense. In the abstract of their 1995 paper, Cortes and Vapnik briefly describe the how the support vector machine affects the linear decision surface: 

> The support-vector network is a new learning machine for two-group classification problems. The machine conceptually implements the following idea: input vectors are non-linearly mapped to a very high-dimension feature space. In this feature space a linear decision surface is constructed [@Cortes_1995, 273]

This 'very high-dimension feature space' is explicitly made to support 'a linear decision surface,' just as Fisher's linear discriminant analysis had. But this linear decision surface is now the product of a non-linear mapping of the data.[^5.300] In Cortes and Vapnik's support vector machine, a 'linear decision surface is constructed' but in a new domain - 'a very high dimension feature space' --  where differences exist less linearly. The dimensions of the feature do not index new sources or kinds of data. Instead, the support vector machine creates them from the existing features of the common vector space.   From the standpoint of pattern recognition, this sometimes vastly augmented vector space should make it harder to see patterns. In fact, it suggests that not all machine learning reduces the dimensionality of data to diagrammatically flat surfaces. A linear or planar decision surface in high dimensional space maps onto a curving even labyrinthine decision boundary when projected back into the original common vector space. In certain cases, machine learners multiply dimensions in data in the name of differentiation, classification, and prediction. Many of the techniques that have accumulated or been gathered into machine learning do flatten variations and differences into lines and planes, but not always by reducing them. In fact, random forests, neural networks and support vector machines exemplify a counter-movement  that maximises variety in the name of differentiation.[^5.22]  Research in machine learning, whether it has been primarily statistical, mathematical or computational,  countenances  and addresses problems of non-linear classification through _dimensional expansion_.  \index{common vector space!dimensionality} As Vapnik writes in the preface to the second edition of _The Nature of Statistical Learning Theory_ [@Vapnik_1999, vii], 'in contrast to classical methods of statistics where in order to control performance one decreases the dimensionality of a feature space, the SVM dramatically increases dimensionality' (vii). \index{Vapnik, Vladimir!dimensional increase}

The powerful variation characteristic of the support vector machine can be understood as a diagrammatic substitution. \index{diagrammatic!substitution} Consider the expression show below:

\begin {equation} 
L_D = \sum_{i=1}^{N} \alpha_i - \frac{1}{2}\sum_{i=1}^{N}\sum_{i'=1}^{N}\alpha_i \alpha_{i'} y_{i} y_{i'}\langle h(x_i), h(x_{i'})\rangle
\label {eq:kernel_model}
\end {equation}
[@Hastie_2009, 423]

In equation \ref{eq:kernel_model}, a re-mapping of the common vector space occurs. All of the data is re-mapped using some function $h(X)$ into a new higher dimensional space. What would be the value of a more complicated space? As Leo Breiman writes in his account of the development of the support vector machine: 

>In two-class data, separability by a hyperplane does not often occur. However, let us increase the dimensionality by adding as additional predictor variables all quadratic monomials in the original predictor variables. ... A hyperplane in the original variables plus quadratic monomials in the original variables is a more complex creature. The possibility of separation is greater. If no separation occurs, add cubic monomials as input features. If there are originally 30 predictor variables, then there are about 40,000 features if monomials up to the fourth degree are added. [@Breiman_2001a, 209] \index{Breiman, Leo!on support vector machine}

The extravagant dimensionality released in the shift from 30 to 40,000 variables vastly expands the number of possible decision surfaces or hyperplanes that might be instituted in the vector space. The support vector machine, however, corrals and manages this massive and sometimes infinite expansion at the same time by only allowing this expansion to occur along particular lines marked out by _kernel functions_. While the support vector machine maintains a commitment to the separating hyperplane, a linear form, albeit with soft margins, it re-constitutes that plane in newly created vector spaces constrained by certain key structural features that render them computationally tractable. On the one hand, a promise of infinite expansion and associated freedom from the rigidity of lines, and on the other hand, a mode of expansion can only countenance a limited range of movements given by the kernel functions. \index{function!kernel}

Hastie and co-authors put it this way:

>We can represent the optimization problem  and its solution in a special way that only involves the input features via inner products. We do this directly for the transformed feature vectors $h(x_i)$. We then see that for particular choices of $h$, these inner products can be computed very cheaply [@Hastie_2009, 423]

The terminology here takes us back the common vector space (see Chapter 2) that machine learning inhabits. The 'inner product' or 'the convolution of the dot-product' described by Cortes and Vapnik come from this space, in which the distances or alignments between whatever can be rendered as a vector can be calculated en masse. In _Top 10 Algorithms for Data Mining_ [@Wu_2008], a widely cited computer science account of data mining, the operation is described in this way:

> The kernel trick is another commonly used technique to solve linearly inseparably problems. This issue is to define an appropriate kernel function based on the _inner product_ between the given data, as a nonlinear transformation of data from the input space to a feature space with higher (even infinite) dimension in order to make the problems linearly separable. The underlying justification can be found in _Cover's theorem_ on the separability of patterns; that is, a complex pattern classification problem case in a high-dimensional space is _more likely_ to be linearly separable than in a low dimensional space [@Wu_2008, 42].

The 'kernel trick' that overcomes inseparability remaps an already transformed vector space -- the inner product of all the vectors in the data -- into a higher dimensional space defined by functions such as $f(x) = x_i^2 + x_i^3$. The trick is no simple technical trick, since as Cortes and Vapnik point out it relies on substantial mathematical developments in the 1960s. 'The idea of constructing support-vector networks comes from considering general forms of the dot-product in a Hilbert space (Anderson & Bahadur, 1966)', write Cortes and Vapnik [@Cortes_1995, 283]. It is a trick, however, in the sense that it is 'can be computed very cheaply' as Hastie et. al. put it.[^5.81]

What does the transformed feature space combined with the computational short cut of the inner product do in practice? In describing what happens to the generalization error -- the errors made when a model classifies hitherto unseen data -- Cortes and Vapnik highlight the growth in dimensionality introduced by the technique of the support vector. They describe how the technique exponentially increases the dimensionality of the feature space and how the error rate on difficult to classify handwritten digits drops correspondingly. When the feature space has 256 dimensions (the given dimensions of the 16x16 pixel digits), the error rate is around 12%. As the dimensionality grows to 33,000, then a million, a billon, a trillion and so forth (up to $1 x 10^16$ dimensions), the error rate drops to just over 4%, which is close to the errors made by 'human performance' (2.5%) [@Cortes_1995, 288].


# Patterns as institutional not spatial-temporal

> The rule of materiality that statements necessarily obey is therefore of the order of the institution rather than of the spatio-temporal localization; it defines possibilities of reinscription and transcription (but also thresholds and limits) , rather than limited and perishable individualities [@Foucault_1972, 103] \index{Foucault, Michel!on statements!materiality of}

Patterns in machine learning are a rule of materiality ordering contemporary statements (remembering that for Foucault statements can be machinic: 'a graph, a growth curve, an age pyramid, a distribution cloud are all statements' [@Foucault_1972, 82]). \index{statement!materiality of}  The spatio-temporality of patterns resides in the 'order of institution' rather the localization because patterns exist in their reinscription and transcription. Pattern recognition is pattern repetition.   Patterns inherently distribute differences. The form of movement engineered into various machine learning techniques grapple with the distribution of differences. But they do it in different ways. Sometimes that take for granted the possibility of separation in a pattern, as if the pattern itself was intrinsically divisible or cleavable along certain lines or contours. At other times, intrinsic inseparability is taken into account as part of the learning.  The power of support vector machine to do this is limited, but instructive. It can deal with various forms of inseparability by taking the difficult-to-classify boundary cases as the basis of the model. It deals with problems of non-linearity by increasing the dimensionality of the data, and looking for separations in the higher-dimensional space. 

What do we learn about pattern from decision tree and their development into random forests, or from linear discriminant analysis and its re-formalization as the support vector machine? We could also have tracked the movement between the perceptron [@Rosenblatt_1958] and the 'deep learning' convolutional neural networks [@Hinton_2006] of more recent practice (see chapter \ref{ch:subjects} . Each of these shifts attests, I have been suggesting, to the emergence of a new enunciative mode in which dispersion and regularity, accumulation and rarity are no longer opposed, or treated as contradictory. A single decision tree becomes thousands in a random forest. A relatively small number of dimensions in the common vector space becomes potentially infinite in the convolutional dot products and kernel functions of support vector machines. Models that sought to encompass or fit everything in the data including the outliers within a single probability distribution instead dwell on the difficult-to-classify, the erroneous or borderline instances amidst the massive normalized accumulations of event.

What counts as pattern today?  Recent machine learners can be studied, I have been suggesting  as elements in a power-laden positivity, \index{positivity!of knowledge} that changes knowledge in specific, actually quite narrow, but nevertheless far-reaching ways. The interpretability of a decision tree cascades into the statistical regularities associated with a random forest of many thousands of trees. The separating lines and planes that allow linear models to become classifiers in the 'classic' techniques such as linear discriminant analysis find themselves displaced into hyper-planes, into newly constructed and sometimes inordinately-dimensioned feature spaces that can only be traversed by virtue of the kernel functions, and their computationally tractable inner products. These transformations are not necessarily scientific, although various sciences make great use of them, but they do diagram  thresholds of epistemologization and formalization in practice.  Despite their quite different ways of partitioning, separating or propagating differences, support vector machines and decision trees define possibilities of grouping and spacing, sometime through purifying, sometimes through bending and blurring. These groupings and spacings attract, generate and accumulate propositions. This grouping-spacing positivity,  space, despite its commonalities, is not an homogeneous field. It does not have the coherence of a science, it uses different systems of formalization (the cross-entropy measures of the decision tree, the gradient descent of the neural network, the dual form Lagrange multipliers of the support vector machine), and disperses in different ways across knowledge practice (the decision tree with its commercial uptake in data mining versus the support vector machine's heavy use in image recognition and classification). As we will see in the following two chapters, deep ordering differences such as living and non-living, saying and seeing, making and using might begin to take on new patterns. 


[^5.01]: See [@Kitchin_2014] for a survey of the major epistemological commitments in 'Big Data' discourse. These largely derived from predictive and inferential techniques of machine learners. \index{Kitchin, Rob!on epistemology of 'big data'}

[^5.02]:  While Foucault tends to retain a decoupled subject-object relation in the production of statements,  I tend to see these enunciative modalities as distributed across people and things. As always, machine learner is a composite term for this distribution. 

[^5.03]:  These objections and resistances to early decision trees echo today in discussions around pattern recognition, knowledge discovery and data-mining in science and commerce.  The problem of what computers do to the analysis of empirical data is long-standing. 

[^5.04]: Many authors have suggested that algorithms should be the focus of more attention. Mike Savage, in his account of the growth of 'descriptive assemblages' based around large scale data mining of transactions, administrative records and social media practice concludes:

    It follows that a core concern might be to scrutinize how pattern is derived and produced in social inscription devices, as a means of considering the robustness of such derivations, what may be left out or made invisible from them, and so forth. We need to develop an account which seeks to criticize notions of the descriptive insofar as this involves the simple generation of categories and groups, and instead focus on the fluid and intensive generation of potential [@Savage_2009, 171] \index{Savage, Mike|on descriptive assemblage}

[^5.05]: The other major machine learner of this time, the neural net, will be discussed in chapter \ref{ch:subject}.


[^5.1]: Quinlan's papers  and book on versions of the decision tree (`ID3` and `c4.5`) are both amongst the top ten the most highly cited references in the machine literature itself. Google Scholar reports over 20,00 citations of the Quinlan's book _C4.5: Programs for Machine Learning_ [@Quinlan_1993] (although far fewer appear in Thomson Reuters Web of Science). Several years ago,  `C4.5` was voted the top data mining algorithm [@Wu_2008]. While I don't discuss Quinlan's work in much detail here, we should note as a computer scientist, Quinlan takes a much more rule-based approach to decision tree that Breiman and co-authors. \index{Quinlan, John Ross!induction tree} \index{machine learner!C4.5}

[^5.06]:  Other `R` packages such as `party` [@Hothorn_2006] and `tree` [@Ripley_2014] also use recursive partitioning, but with various tweaks and optimisations that I leave aside here. 

[^5.08]: _Elements of Statistical Learning_ uses the term 'pattern' only occasionally. The term appears 33 times there, and mainly in the bibliography. Apart from Brian Ripley's _Pattern Recognition and Neural Networks_ [@Ripley_1996], statisticians largely eschew the term. Computer scientists like it more, and particularly in work on the classification of images (see Christopher Bishop _Pattern Recognition and Machine Learning_ [@Bishop_2006]. Hastie, Tibshirani and Friedman, as statistical  machine learners, confine their use of pattern to the term 'pattern recognition'. \index{pattern!as a term in machine learning}

[^5.12]:  `iris` is a very small dataset, a pre-computational miniature.  That diminutive character makes it diagrammatically mobile.  It supports a rhizomatic ecosystem of examples scattered across the machine learning literature.    The usual framing of the classification problem is how to decide whether a given iris blossom is of the species _virginica_, _setosa_ or _versicolor_.   These irises don't grow in forests -- they are more often found in riverbanks and meadows --  but they do offer a variety of illustrations of how machine learning classifiers are brought to bear on classification problems. Here the classification problem is taxonomic - the  iris genus has various sub-genera, and sections within the sub-genera. _Setosa, _virginica_ and _versicolor_ all belong to the sub-genus _Limniris_. This botanical context is  routinely ignored in  machine learning applications. In machine learning textbooks and tutorials, iris typically would be used to demonstrate how cleanly a classifier can separate the different kinds of irises. \index{dataset!iris}

[^5.101]: Foucault assigns all statement to referentials. He states that 'the referential of the statement forms the place, the condition, the field of emergence, the authority to differentiate between individuals or objects, states of things and relations that are brought into play by the statement itself; it defines the possibilities of appearance and delimitation of that which gives meaning to the sentence, a value as truth to the proposition' [@Foucault_1972, 91].  \index{statement!referential of}

[^5.102]: Chapter \ref{ch:subjects} discusses subject positions and machine learning. The current discussion concerns changes in referentiality associated with machine learning. 

[^5.70]: The support vector machine is distinctive in its mode of movement, and without being swamped by the technical details, we might grasp something of this from Vapnik's writings. Vapnik trained and worked of decades in the former USSR as a mathematician and statistician. His writings on the problems of pattern recognition contrast greatly with other engineers, statisticians and computer scientists  in their  robustly theoretical formalism. A highly cited 1971 publication with  Alexey Chervonenkis 'On the uniform convergence of relative frequencies of events to their probabilities' (published in Russian in 1968 ) [@Vapnik_1971] sets the formal tone of this work. In ensuing publications in Russian and then in English after Vapnik moved from Moscow to AT&T's New Jersey Bell Labs in 1990, Vapnik's work remains quite formally mathematical. Although it pertains to 'learning machines,' machine here are understood mathematically simply as 'the implementation of a set of functions' [@Vapnik_1999, 17]. \index{Vapnik, Vladimir} The way that Vapnik develops a theory of learning owes little visible debt to actual attempts to work with data or experience in doing statistics in any particular domain. This contrasts greatly for instance with the work of statisticians like Breiman or Friedman or even computer scientists like Quinlan or Le Cun whose work lies much closer to fields of application.  Vapnik's work, like that of the Russian mathematician Andrey Kolmogorov he draws on, differs from many other contributions to machine learning partly by virtue of this formality and its efforts to derive insight into machine learning by theorising learning. The _Vapnik-Chervonenkis dimension_( VC dimension), a very widely used way of defining the capacity of a particular machine learning technique to recognise patterns in data dates from his work in the 1960s and underpins a general theory of 'learning.' Vapnik writes in 1995, 

    > The VC dimension of the set of functions (rather than the number of parameters) is responsible for the generalization ability of learning machines. This opens remarkable opportunities to overcome the "curse of dimensionality [@Vapnik_1999, 83].

    As we will see in this chapter, Vapnik's attempts to overcome dimensionality also re-shape what counts as pattern. 

[^5.112]: _Elements of Statistical Learning_ also devotes a chapter to support vector machines [@Hastie_2009, Chapter 14]

[^5.22]: Despite the in-principle commitment to any form of function, machine learning strongly prefers forms that can either be visualised on a plane (using the visual grammar of lines, dots, axes, labels, colours, shapes, etc.), or can be computed in form of matrix or vectorised calculations focused on planes. \index{vectorisation} Many of the techniques that grapple with complicated datasets seek to reduce their dimensionality so that lines, planes and regular curves can be applied to them: multi-dimensional scaling (MDS),  factor analysis, principal component analysis (PCA), or self-organising maps (SOM) are just a few examples of this. \index{function!linear}

[^5.23]: Neural network researchers have heavily used the MNIST dataset. I discuss some of that work in chapter \ref{ch:subjects}.  The handwritten MNIST also appear in _Elements of Statistical Learning_ , where they are used to compare the generalization error (see previous chapter) of a _k_ nearest neighbours, convolutional neural network, and a 'degree-9 polynomial' support vector machine  [@Hastie_2009, 408]. \index{dataset!MNIST} What about the handwritten digits attracts so many machine learning techniques? The logistics of the US Postal Service aside (since the MNIST datasets continue to be used by machine learners well after the problem of scrawl on letters has been sorted), the variations, the regularities, and the banal everydayness of these digits  furnish a referential locus, whose existence as a facts, things or events in the world is less important than the relations of similarity and differences it poses. \index{referential} 
The field of digitals becomes a site of differentiation not only of digits -- the machine learners attempt to correctly classify the digits -- but of the authority of different machine learning techniques and approaches. They become ways of announcing and delimiting the authority, the knowledge claims or 'truth' associated with the machine. The many uses of the MNIST data documented by [@LeCun_2012] suggests something of the referential force of such datasets.

[^5.300]:  The 'linear decision surface' is an old and familiar entity in statistics. As we have seen on several occasions, the common vector space invites a certain form of classification based on the search for the best line, the line of best fit, or the most discriminating line, the line that best divides things from each other. Linear regression is not called 'linear' for no reason. And Fisher's 'discriminant functions' were later called 'linear discriminant analysis' for the same reason: they divide the vector space into different regions ('decision regions') separated by 'linear decision boundaries' [@Alpaydin_2010, 53]. Almost all machine learners are aware of and try to address the idealism or abstraction of the line or plane. \index{abstraction!of line and plane}

[^5.81]: This cheapness appeared already in the cat machine learner  discussed in the introduction. Heather McArthur's`kittydar` cat image classifier implemented a support vector machine in Javasript that runs in a browser. Cats are classified cheaply there. \index{machine learner!kittydar}
