\chapter{The vector space and its products}
\label{ch:vector}

# The vector space and its products

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, tidy=TRUE, fig.height=8, 
                      echo=FALSE,  warning=FALSE, message=FALSE, dev='pdf')

library(RSQLite)
con <- dbConnect(RSQLite::SQLite(),'../ml_lit/all_refs.sqlite3')
res = dbGetQuery(con, statement ="select * from basic_refs limit 10;")
library(ggplot2)
library(stringr)
library(xtable)
options(xtable.comment = FALSE)
```


Do machine learners  want to go everywhere in the world, take on any problem, and see no data as too difficult to learn on? This formulation from Matthew Kirk's _Thoughtful Machine Learning_ is typical:

> Machine learning is an amazing application of computation because it tackles problems that are straight out of science fiction. These algorithms can solve voice recognition, mapping, recommendations, and disease detection. The applications are endless, which is what makes machine learning so fascinating [@Kirk_2014, 11] \index{Kirk, Matthew}

It is harder but not impossible to find dissenting or sceptical voices.  Computer scientists such as Anand Rajaraman and David Ullman in their _Mining Massive Datasets_ voice doubts about the general power of machine learning algorithms and contrast these algorithms with specialized search and retrieval techniques [@Rajaraman_2012]. Similarly, the mathematician Cathy O'Neil trenchantly criticises the rapid extension of machine learning approaches in finance. Regardless of reservations and criticisms, the appearance of 'endless' applications is not misleading. Machine learning has, once you start looking for it, become astonishingly generalized. The epistemic diversity of  its operation is not only due to the generic mathematical, statistical or computational advances. It attests to the advent of a new model of truth, which opens onto an expanding epistemic continuum. The continuum arises, I will suggest, from a very specific diagrammatization of data. This diagrammatic movement generates the continuum (a continuum that has already surfaced in the perceptron's treatment of NAND truth table as discussed in chapter \ref{ch:diagram}) that *vectorises* data. Data moves vectorally; that is, in terms of displacements in specific directions. \index{data!vectorisation|see{common vector space!vectorisation}

Often data is  represented as if it is an homogeneous mass or a continuous stream. If there is any such homogeneity,  we should examine the transformations that allow different shapes and densities of data to mix. Data in its local complexes spaces out in many different density shapes, depending on how the data has been generated or instanced [^2.104]. Whatever the starting point -- a measuring instrument, people clicking and typing on websites, a device like a camera, a random number generator, etc. --  a given machine learning algorithm, data visualisation or database query will need the data to be in specific *vectorised* shapes (vectors,  matrices, arrays, etc), and will scale it within a certain volume.  \index{vectorisation} \index{data!volume}  The process of composing data for statistical, visual, predictive or even storage purposes, maps a concrete situation, some state of affairs, onto forms imbued with various geometrical, probabilistic, and decisionist abstractions often expressed in terms of functions or mathematical models. (This is sometimes referred to as 'data cleaning' but this term covers over some important transformations.)  If, as I have been proposing, we seek contact with the practices or concrete value-situations, then the reshaping and reflowing of densities matters greatly.   This forming and reforming of data is evidence of  implicated relations. These practices attest to what the philosopher A.N. Whitehead called 'strain': 'a feeling in which the forms exemplified in the datum concern geometrical, straight, and flat loci will be called a "strain." In a strain, qualitative elements, other than the geometrical forms, express themselves as qualities implicated in those forms' [@Whitehead_1960, 310]. \index{Whitehead, A. North}  \index{data!strain} In many machine learning models the exemplified forms are straight or flat loci (as we see in chapter \ref{ch:function}).  Yet  different practices also seek to elicit relations that strain the linear shaping of data, that show how it does not fit.  Some  practices working in the name of linear forms effectively trace strain feelings. These feelings are affective connectors between the abstractions and the concrete-value situations in which data arises. 

Statistical modelling, data-mining, pattern recognition, recommendation systems, network modelling and machine learning rely very much on  the transformation called 'fitting a model.' \index{model!fitting} The basic machine learning work of 'fitting a model' (or many models)  to data is often literally implemented, as we will see, by  constraining data within a continuum, which I term the *common vector space.* \index{common vector space}  Familiar narratives of Moore's Law increases in  speed or efficiency of computation do not account for transformations of   data in  `R` and other computing environments (for instance, the popular `Map-Reduce` architecture invented at Google Corporation to speed up its search engine services [@Mackenzie_2011] \index{data!architecture!map-reduce}).  Vectoral movements into data tap into and are interwoven with the transformation of data along much more diagrammatic lines.[^2.1] 

```{r elemstatlearn_data_2, echo=FALSE, results = "asis"}
library(ElemStatLearn)
library(xtable)
datasets_results <-data(package='ElemStatLearn')$results[,3:4]
datasets = as.data.frame(datasets_results)
dataset_count <- nrow(datasets)
dataset_table = xtable(datasets, caption='Datasets in _Elements of Statistical Learning_', label = 'tab:datasets_elemstatlearn')
print.xtable(dataset_table, type='latex')
```

In _Elements of Statistical Learning_,  data appears in various forms. Plots of New Zealand fishing patterns lie next to plots of factors in South African heart disease. The `r dataset_count` datasets  shown in table \ref{tab:datasets_elemstatlearn'} typify the variety of domains found in [@Hastie_2009; @Hastie_2001]. They span scientific, clinical, commercial and media fields. Note that they include many patches and pathways of everyday life -- speaking, seeing, writing and reading -- as well as specific scientific objects of knowledge such as galaxies, cancer, climate and national economies. This span of situations is not unusual for machine learners.  To give another example, the statistician Leo Breiman's 2001 article on random forests [@Breiman_2001] \index{random forest} is one of the most cited journal papers in the machine learning literature and displays a similarly diagonal line across human and non-human worlds: \index{Breiman, Leo}

>Glass,  Breast cancer,  Diabetes,  Sonar,  Vowel,  Ionosphere,  Vehicle,  Soybean,  German credit,  Image,  Ecoli,  Votes,  Liver,  Letters,  Sat-images,  Zip-code,  Waveform,  Twonorm,  Threenorm,  Ringnorm,  [@Breiman_2001, 12].

In some ways, the improbable conjunction of spam email and cancer detection in machine learning simply continues what statistics as a field has always done: rove  across scattered fields ranging from astronomy to statecraft, from zoology to epidemiology, gleaning data as it goes (see [@Stigler_1986; @Hacking_1990] for samples of this trekking movement). \index{statistics} But in _Elements of Statistical Learning_ and the field of machine learning more generally, something more is moving through and associating these datasets. The colligation of datasets -- vowels, ozone, bone density, marketing, prostate cancer, and spam -- in all their diversity  demonstrates the mobility of machine learners. The diversity of the data could be seen as almost aleatory, as if the datasets were the fruit of random derivè in the world.  Rather than random drifting, I think it more likely that the diversity of datasets allows machine learners to perform something like one of its own statistical procedures in 'bagging' (bootstrap aggregating [@Hastie_2009, 300]) datasets in order to average its predictions out across different domains. \index{bagging} The repeated sampling of the world bootstraps machine learning as a field into its operational mode of generalization.\index{generalization}  

The tabular form, the practices of naming and labelling, and the sorting of different data types exemplified in these diverse datasets can tell us a lot about how machine learning organises and energises vectoral movement through data.  The different shapes and composition of the datasets provide indications of the functioning of machine learners as they make knowledge or operationalise power relations. Datasets have visible forms of alignments or architectures that we should pay close attention to as they are learned machinically. Already in table \ref{tab:datasets_elemsstatlearn} differences between datasets labelled 'training' and 'test' suggest something of this learning. We might also notice differences between data called 'simulated' and other kinds of data. If machine learning can be understood as a constantly evolving diagram or an abstracting multiplicity, the way it moves through data and associates different forms of data matter. This movement and connecting is somewhat guided by the shape of the data in some ways (for instance, very many demonstrations of machine learning use R.A. Fisher's `iris` dataset [@Fisher_1936]). \index{dataset!iris}  But as we will see in this chapter, it also generates data in an increasingly extensive, heavily conventional space, the  common vector space.  Building on the diagrammatic account of the _Elements of Statistical Learning_ developed in the previous chapter, this chapter attends to the conventions, regularities and the mobilities of data in the common vector space. \index{common vector space!conventions of}

# Mixing media, medicine, business and biology

>References to things act simultaneously as reference to (and within) activities. [@Lynch_1993, 193] \index{referential!things and activities}

'This book is about learning from data' write Hastie, Tibshirani and Friedman on the first page of _Elements of Statistical Learning_, as they rapidly begin to iterate through some datasets. On the second page of the book [@Hastie_2009, 2], a table of spam email word frequencies appears (and the problem of spam classification is canonical in the machine learning literature - we return to in chapter \ref{ch:probability}). \index{dataset!spam} They come from the dataset `spam` [@Cranor_1998]. On the third page, a complicated data graphic appears (Figure 1.1, [@Hastie_2009, 3]. It is a scatterplot matrix of the `prostate` dataset included in the `R` package `ElemStatLearn,` the companion `R` package for the book. In a third example, a set of scanned handwritten numbers appears. These scans are images of zipcode or postcode numbers written on postal envelopes taken from the dataset `zip` [@LeCun_2012], and they differ from both the `spam` table and `prostate` plots because they directly resemble something in the world, which, however, happens to be numbers, and is, therefore, probably already recruited into data-making and data-circulating processes (a dataset we return to in chapter \ref{ch:subject}).\index{dataset|zip}  The final example in the introduction, 'Example 4: DNA Expression Microarrays,' draws this time from biology, and particularly, high-throughput genomic biology, the kind of science that produces large amounts of data about something in the world by running many tests, or by constructing devices that generate many measurements, in this case, a DNA microarray.[^2.10] The image shown here is perhaps most striking. No numbers are shown, only a colorful heatmap with brighter colours standing for higher levels of gene expression, and darker colours for lower levels. \index{graphic!heatmap} For all its dense color, the data shown here is a sample of 100 of the approximately 7000 genes in the dataset `nci`  [@Hastie_2009, 6]. In comparison to the medical data from the `prostate` dataset with its 97 rows of 10 columns, the microarray dataset has 64 columns of data.[^2.11] Both are cancer datasets, but the `nci` dataset cannot be shown in its entirety because it refers to 60 different cell lines ranging across colon, breast, prostate, ovarian and renal cancers, as well as the thousands of different genes.\index{cancer} \index{dataset!nci}

The combination of datasets deriving from network media, from medicine, from business administration and from cutting-edge life science (c.2000) suggests a tremendous, indeed almost spectacular miscibility, one that in principle could surprise us because there is otherwise little mixing between the places these datasets come from. In passing, we should note that the _Elements of Statistical Learning_ is not alone in this juxapositional opening. Very similar example sets can be seen arrayed in most machine learning publications. To give just a few examples: Andrew Ng's CS229 lectures, for instance, treat spam  in Lecture 5 [@Ng_2008h]; Rachel Schutt and Cathy O'Neil's _Doing Data Science_ discusses spam in Chapter 4 [@Schutt_2013], and Peter Flach's _Machine Learning The Art and Science of Algorithms that Make Sense of Data_ [@Flach_2012] introduces spam in the first few pages. \index{dataset!spam} Similar parallels exist around the image recognition problem (exemplified by handwritten digits), around the measurements dataset (`prostate` in the case of [@Hastie_2009], and in relation to the larger, impossible-to-see-the-patterns datasets of the cancer microarray data). How is this mixing, conformation and homogenisation being done? The repetition of data sets, the juxtaposition of discontinuous domains, and forms of movement construct and order continuities in the service of various forms of predictive and inferential knowledge. The miscible juxtapositions we are dealing with here produce, it seems, a _regularity_ or a continuous common space, a space that displays strong tendencies to expand. 

# Truth is no longer in the table?

'Things, in their fundamental truth,' writes Foucault in _The Order of Things_ 'have now escaped from the space of the table' [@Foucault_1992, 239]. \index{table!history |(} Foucault of course was writing in these pages about the fabled emergence of life, labour and language as the anchoring vertexes of a new triangle of knowledge and power structuring the figure of the 'human' in the 19th century. The 'Classical' table as a space of order  was abandoned because it could no longer serve as the 'common place for representation and things, for empirical visibility and for the essential rules' (239). Before the emergence of the characteristic sciences of the human -- political economy, linguistics and biology -- knowledges such as natural history, the general grammars, and philosophies of wealth (such as Adam Smith's work)  had ordered empirical materials of diverse provenance in tables or grids. These tables themselves supplanted earlier much more variegated Renaissance tabulations. In pre-Classical tables, an image or figure from myth might lie alongside a measurement or a count of occurrences, and this proximity was ordered by systems of analogical association that spanned what we might today, in the light of the 19th century, still want to separate.  While the history of tables as data forms reaches a long way back (see [@Marchese_2013] for a broad historical overview that reaches back to Mesopotamia), and tables still definitely abounded and indeed multiplied in the 19th and then 20th centuries, Foucault argues that the system of grids that permitted ranking, sorting and ordering in tables had between the end of the Renaissance and the early nineteenth been based on 'buried similitudes' and 'invisible analogies' [@Foucault_1992, 26]. The Classical Age grid or table was a way of bringing plural and diverse resemblances into exhaustive enumeration where they could be subjected to analysis by counting and comparison. The table as space of order did not stand in isolation. It served a particular epistemic function. While algebra or _mathesis_ more generally applied to 'simple natures' (planets in movement, dynamics of falling bodies, etc.), table-based knowledges such as taxonomy dealt with more complex natures. Even in the tables, systems of signs -- the groupings established by the eighteenth century taxonomist Carl Linnaeus -- sought to reduce complex natures (plants, animals, etc.) to simpler forms as columns and rows in a table based on resemblances and similarities. \index{Linnaeus, Carl} \index{similarity}  Importantly, the table as space of order was a space of imagination in that one could begin to see continuities and the genesis of things (including grammar, nature and wealth) by carefully ordering and scanning the table. 'Hedged in by calculus and genesis,' Foucault suggest, 'we have the area of the _table_' [@Foucault_1992, 73]. \index{calculus} Note in passing that calculus  and calculations bound the table only in relation to 'simple natures' whose identity and difference can be understood in the form of movements, rates and change in position. This is an important limitation since it is precisely the complex natures in genesis that machine learning tries to engage with using algebra, calculus, statistics and computation.

In the nineteenth century, a different form of ordering shattered tabulation based on similarity and resemblance. Foucault figures this change as shattering:

>The space of order, which served as a _common place_ for representation and for things, for empirical visibility and for the essential rules, which united the regularities of nature and the resemblances of imagination in the grid of identities and differences, which displayed the empirical sequence of representations in a simultaneous table, and made it possible to scan step by step, in accordance with a logical sequence, the totality of nature's elements thus rendered contemporaneous with one another -- this space of order is from now on shattered: there will be things, with their own organic structures, their hidden veins, the spaces that articulates them, the time that produces them; and then representation, a purely temporal succession, in which those things address themselves (always partially) to a subjectivity [@Foucault_1992, 239-240].

Life, labour and language -- the figure of the human  -- famously replace the exhaustive, aggregative, synoptic classificatory tables of the Classical age. Tables still abound in newer orderings (almost any episode from the history of nineteenth and twentieth century statistics will confirm that; see [@Stigler_1986; @Stigler_2002]) \index{statistics!history}, but from now on there are things, organically structured, relating to a place and changing in time, and there are representations addressed to (and constitute) of a subjectivity. Double interiority takes over. Things such as a language, a species or an economic system have their own genesis, and our knowledges and indeed experience too become finite, historical, with their own dynamics and internal life. Knowledge is, for instance,  epistemologized in various domains so as to become scientific (through the use of experimental practice, certain ways of writing, etc.; see [@Shapin_1989] for one account of this change), but in this change, the table itself is no longer the foundation of knowledge. It is one apparatus amongst many. \index{table!history|)}

This brief résumé of one of the main arguments in _The Order of Things_ might help us reassess what machine learning  does in data. The question here is how we make sense of these tables in terms of their internal structures and their relation to what lies outside them. Presumably the `spam`, `prostate`, `image` and `microarray` data do not operate in the way Linnaeus would have put together a table of living things based on similarities and resemblances. But, as we will soon see, measures of similarity and resemblance operate strongly in machine learning as it moves through tables. In this respect, the Classical table perhaps returns with renewed relevance. Moreover, the repeated juxtaposition of tables of diverse provenance alongside each other suggests we should be looking for alignments and resemblances between whole tables as well as between particular cells, rows, columns or margins of a table.  The topics of these tables -- as we have seen, in the opening pages of _Elements of Statistical Learning_ they concern work, life, language and economy in various ways -- maps very readily onto the anchor points, the new 'empiricities' (Foucault's term for the empirical problems)  of labour, life and language or biopower \index{biopower} that took root at this time [@Foucault_1991]. \index{empiricities}  Things may no longer be hedged in by the table. Their fundamental truth may have escaped from the space of similarities found in the Classical table. And yet we are now confronted by tables whose expansion and assimilation derives from  calculations of similarity and distance. In many practical respects, we are hedged in by tables, and many different mechanisms animate and multiply tables around us.[^2.22] In certain respects the tables we see in the first pages of [@Hastie_2009] are much more classical in their scale and in their relatively immutability. In the flagship settings such as social media platforms or genomics in which machine learning operates, tables change rapidly in scale and sometimes in organisation. Regardless of any  differences between the tables of data used to demonstrate  machine learning techniques and data tables used in contemporary operational environments, the multiplication and juxtaposition of different tables today suggests that we might be seeing the advent of a non-classical common space of order for representation and things, for regularities and resemblances, for nature and imagination, or science and media. If that is so, then the question will be what kind of scanning, what kind of step-by-step, row-by-row movement, traverses them.


# The epistopic fault line in tables

```[r ElemStatLearn_data, echo=TRUE}
library('ElemStatLearn')
datasets = data(package='ElemStatLearn')
datasets$results[,3]
```

The `ElemStatLearn` `R` package brings, as we have seen in the previous chapter, with it around 20 different datasets, including the four we see in the opening examples. At this point, I'm not so much interested in gauging the rhetorical effect of this mixing of differences,  as in exploring how  _Elements of Statistical Learning_ goes into and between data tables.  On the one hand, every the data table indexes a localised complex of activities (clinical, social media platform, financial transactions, etc.). On the other hand, for machine learners, the table is a space of similarities and differences both within the table itself (e.g. how much does row number 1000 differ from row 1,000,000) and between tables (e.g. how much does this table of clinical test results differ from that table of microarray data?).  These internal and external differences entwine with each other in ways that create a fault line. It is this fault or fold line, I propose, that diagrammatically transforms data tables into the expansive and moving substrate of the common vector space. \index{diagrammatic movement} 

How can we see this fault line? In edging along this fault, we might proceed *epistopically*.  The term 'epistopic' \index{epistopic} comes from the work of the science studies scholar Mike Lynch \index{Lynch, Mike}, whose account of scientific practice is usefully focused on ordinariness\index{practice!scientific}. He proposed the 'epistopic' as a way of connecting 'familiar themes from epistemology and general methodology' [@Lynch_1993, 280] with localized practices, or with the local achievement of coherence. In other words, as the term itself suggests, an epistopic connects general epistemic themes such as validity, precision, specificity, error, confidence, expectation, likelihood, uncertainty, or approximation with a place, a 'local complex of activities' (281). \index{machine learning!statistical aspects}  This emphasis on epistemic-placedness is very useful in several ways, especially if it can be brought to bear on the problem of what happens when the 'local complex' of a specific dataset encounters a generalizing epistemic practice such as machine learning with its concerns with generalization, error, and prediction.  It's interest in the couplings between 'graphism' and practice should help make sense of some of the diagrams (plots, equations, tables) we see in _Elements of Statistical Learning_ as forms of movement that epistemologize data, or that inscribe thresholds of epistemologization in data. \index{epistemologization!threshold of|seealso{epistopic}} A focus on the connections between down-to-earth practices such as drawing a line, labelling a point or sorting a list of values and seemingly generic scientific-epistemological statements about  error, bias, variation, confidence, expectation and likelihood is a good antidote to the sometimes ballistic epistemological trajectories fuelled by statistical machine learning (e.g. the infamous and heavily cited comments by the former _Wired_ magazine editor, Chris Anderson, on the 'end of theory').[^2.41] \index{epistemology} \index{Anderson, Chris}

# Surface and depths: the problem of volume in data 

Every distinct column  in a table effectively adds a new dimension to the common vector space. Since the 1950s, problems of classification and prediction in high-dimensional dataspaces have been the object of mathematical interest. The mathematician Richard Bellman coined the term ‘the curse of dimensionality’ to describe how partitioning becomes more unstable  as the dimensions of the dataspace increase  [@Bellman_1961]. \index{Bellman, Richard!curse of dimensionality} The problem is that while the volume of a space increases exponentially with dimensions,  the number of data points (actual measurements or observations) usually does not usually increase at the same rate. In high dimensional spaces, the data becomes more thinly spread out. It is hard to partitions sparsely populated spaces because they accommodate many different boundaries. \index{common vector space!dimensionality!curse of}

In the epistopic mode of diagramming, machine learners  find edges in data tables that are capable of combining or conjugating with each other, even if they are not reducible to each other. \index{machine learner!epistopic mode}  Let us see how Hastie, Tibshirani and Friedman begin to move through the datasets. First of all, they highlight in the first three (`spam`, `prostate` and `zip`, leaving aside the cancer genes dataset (`nci`)),  what they have in common:

>The first three examples described in Chapter 1 have several components in common. For each there is a set of variables that might be denoted as inputs, which are measured or preset. These have some influence on one or more outputs. For each example the goal is to use the inputs to predict the values of the outputs. This exercise is called supervised learning. We have used the more modern language of machine learning.[@Hastie_2009, 9].

In the ensuing discussion,  they begin to dissect the tables and point to various discontinuities traversing the tables. The major discontinuity is already remarked in the common cut between 'inputs which are measured' and values that might called 'outputs' that inputs 'influence.' Not all tables are cut by machine learners in this way, but this cut is epistopic since the table now shows that some things are measured and some might not be (the values of the outputs). In addition to this vertical discontinuity dividing tables, more diverse horizontal discontinuities can be seen (and Hastie et. al. go on to describe it). Some of the variables in the table are continuous numerical values, others are discrete (they can only take on certain values such as a whole number), and indeed may be categorical or nominal (for instance, when predicting the actual digit from the handwritten digits, the predicted digit itself is a categorical variable, since the prediction must choose one of the digits 0,1,2,3,4,5,6,7,8 or 9). \index{data!variable!types} Discrete qualitative data appears frequently in machine learning, or wherever classification problems can be found. \index{classification} (I will have much more to say about classification and machine learning classifiers in later chapters. For the moment, we need only note that predicting the class to which things belong is a major topic of interest.)

```{r prostate_plot, echo=TRUE, message=FALSE, results = 'hide', fig.cap=''}
library('ElemStatLearn')
data('prostate')
pairs(prostate[, -10], main='Scatterplot matrix of the `prostate` dataset', cex=0.2)
```

```{r prostate_data, results='asis'}

library(xtable)
print(xtable(head(prostate), label= 'tab:prostate', caption='First rows of the \`prostate\` dataset'))
```

\begin{figure}
  \centering
      \includegraphics[width=0.9\textwidth]{figure/prostate_plot-1.pdf}
        \caption{Scatter plot matrix of prostate data}
  \label{fig:prostate_plot_matrix}
\end{figure}

Of the three example datasets  (`prostate`, `spam` and `zip`), the authors return most frequently to `prostate.` This dataset derives from the work of urologists working at Stanford [@Stamey_1989], and, as [@Hastie_2009, 67] points out, concerns various clinical measurements performed on men who were about to undergo radical prostatectomy. The measurements range across the volume and weight of the prostate, as well as levels of various prostate-related biomarkers such as PSA -- prostate specific antigen. Several rows from the dataset are shown in Table \ref{tab:prostate}. In the first pages of the book, they had already quite exhaustively plotted all the variables in the dataset against each other using a scatter plot matrix [@Hastie_2009,3] (show in \ref{fig:prostate_plot_matrix}, and they return to the same data on almost a dozen occasions in the course of the book. \index{dataset!prostate} \index{graphics!scatterplot matrix}

The transformation between the Table \ref{tab:prostate} and the Figure \ref{fig:prostate} already evinces a vector \index{vector} of movement commonly found around such datasets. On the one hand, the table arrays all the different data types indifferently in rows and columns. The relation between the different data types (the log of the weight of prostate - `lwp` and `age` for instance) is quite hard to see. Moreover, different kinds of variables stand side by side. `svi`, short for 'seminal vesicle invasion' is a categorical variable. It takes the values 'true' or 'false,' shown here as `1` or `0`, but the other variables either measure or count things (years, sizes, or levels of antigens).  On the other hand, the scatter plot matrix also takes the form of a grid-like figure, but the squares of the grid are not occupied by numbers but by `x-y` plots of different pairs of variables in the `prostate` dataset. The 'matrix' of figures shows of 72 plots is mirrored across the diagonal that runs from the top-life to bottom right in figure \ref{fig:prostate}. Taking this folding into account, we see 36 unique plots with different data in each one. Each sub-plot displays the relation between two variables in the dataset as a scatter plot. Note how certain variables such as `svi` are not very amenable to plotting in this way. More importantly perhaps, note how certain combinations of variables result in visible forms can be seen as signs of relations between different variables. The matrix constructs a space in which contrasts between the pairs of variables that yield scattered clouds of points and those that show distinct clusters or tendencies can be seen. In the light of these contrasts (and I use 'light' here in an almost literal sense to refer to the way in which the architecture of the figure creates a space in which light scatters in varying patterns), th `prostate` dataset begins to expose relations that might be worth knowing about. We have moved on from the bare table of the dataset to a transformed tabulation, from a textual-numerical grid to a geometrical-textual grid. Everything remains on the surface of a grid here, but the grid functions differently in the scatterplot matrix. 

All of this lies somewhat in a pre-machine learning space. Similar tables and plots are part and parcel of statistical data analysis more generally (see [@Beniger_1978] for an historical account of quantitative graphics in statistics \index{statistics!graphics|seealso{graphics}). But the point of the scatterplot matrix is not to exhaust the dataset, but rather to highlight the need to constantly revisit it (12 times in the _Elements of Statistical Learning_) in order to tease out the multiple relations or influences that remain opaque to even the most exhaustive matrices of plots. We cannot clearly see in the scatterplot matrix beyond pairs of variables in relation. If the crucial diagnostic measure in this case is the level of the PSA (prostate specific antigen), how do we know what combinations of other measurements might be associated with its elevation? What is multiple variables affect the level of PSA?[^2.31] This question can be pursued by scanning the matrix of plots but not very stably since different data analysts might see different associations combining with each other there. Different statements could be supported by the same figure. Perhaps worse, the very question of relation between multiple variables and the predicted levels of PSA suggests the existence of a hidden volume, an occluded or internal space that cannot be seen in a data table, or that cannot be brought to light in the geometry of a plot. This volume is not the measured volume of the prostate but the virtual volume suggested by both the dataset table and the scatterplot matrix, a nine dimensional space subtended by treating each of the nine variables in the dataset as occupying a dimension. \index{common vector space!dimension}  When Foucault wrote of truth escaping the table, he might well have pointed towards the higher-dimensional volumes of the _vector space_ into which lines were already beginning to regress even in the late 18th century.[^2.33]. A different basis of order -- the common vector space -- \index{common vector space|(}  begins to take shape when the variables (usually columns) become dimensions.  Strenuous efforts are made in machine learning to ensure that as much as possible goes into and comes of that common space. 

# Vector spaces expand

To show this space in its epistopic making, we might bfollow what happens to just one or two columns of the `prostate` data in the common vector space as it is prepared for machine learning. In the `prostate` dataset, some variables are continuous quantitative values, and some are categorical (they represent membership in a group or category) or ordinal variables (they represent a ranking or order). But a variable such as `svi` has `True` or `False` values. How can such values be positioned in the vector space, and thus be subject to calculation in the same way as a measure of prostate gland volume? In order to put classifications or categories into vector space, they need to be translated into the same _basis_ as the quantitative variables with their rather more obvious geometrical and linear coordinate values. \index{classification} How does one geometrically or indeed algebraically render a category so that it can be mobilised in the way that equation \ref{eq:linear_regression} (see chapter \ref{ch:diagram}), the vector form of the linear regression model, suggests it might be? \index{linear regression model} The problem is solved via a form of binary coding:

>Qualitative variables are typically represented numerically by codes. The easiest case is when there are only two classes or categories, such as “success” or “failure,” “survived” or “died.” These are often represented by a single binary digit or bit as 0 or 1, or else by 1 and 1 ... When there are more than two categories, several alternatives are available. The most useful and commonly used coding is via dummy variables. Here a K-level qualitative variable is represented by a vector of K binary variables or bits, only one of which is “on” at a time [@Hastie_2009, 12] 

Again, the details are not so important here as the transformations that the common space permits once things inhabit it. Note that a single qualitative or categorical variable expands into 'a vector of K binary variables or bits.' \index{data!variable!qualitative|seealso{data!variable!categorical}} The dimensions of the vector space expand accordingly, and the machine learners  treat these added dimensions as variables to be included in the model. Qualitative data, once coded in this way, can be multiplied, added, and in short, handled algebraically using the same aggregate operations we saw in discussing linear algebra more generally. Note also that not only has the vector space expanded here, this expansion smooths over important gaps or differences that figure large in the dataset. Complex natures become simple natures. The different kinds of variables -- qualitative and quantitative, discrete and continuous, nominal and ordinal -- can be accommodated by adding dimensions to the vector space. \index{common vector space|)}

# Traversing behind the light 

Adding dimensions to the common vector space makes seeing the volumes and densities of data distributed in this space more challenging.   The character of the transformations in `prostate` that ensue in _Elements of Statistical Learning_ are difficult to summarise. They are the locus of machine learning as an epistopic form of movement. \index{machine learning!epistopic}  In terms of the historical transformations of the table, they can be seen as subtending differences in a common vector space of simple natures understood as things with direction and displacement, and hence susceptible to calculation as vectors. \index{common vector space!dimension}  Once this hidden virtual volume in the data is glimpsed, strenuous efforts will be made to bring it to light, even sometimes blatantly disregarding the local complex of activities that originally produced the data. These efforts will proceed along different lines. Sometimes this space is treated as one filled with constantly varying proximities and similarities, and machine learning techniques gather and order these differences (for instance, as in the _k_ nearest neighbours model \index{\texit{k}-nearest neighbours model}) or in unsupervised methods such as k-means clustering [@Hastie_2009, 513]\index{\testit{k}-means clustering}.  More commonly, machine learning draws lines or flat surfaces that con-strain through the volume using some version of linear regression, perhaps the most important modelling technique in modern statistics\index{linear regression}. \index{common vector space|strain}

\index{linear model} The importance of the linear modelling, finding lines of best fit,  should not be under-estimated amidst the plethora of machine learners that attract more recent attention (neural nets, support vector machines, random forests). Linear regression lays down the basis of common vector space that renders all differences as distances and directions of movement. Drawing lines or flat surfaces at various angles and directions is perhaps the main way in which the volume of data is traversed, and a relation between input and output, between predictors and prediction, consolidated.[^2.351] While it is more or less obvious to the eye from the scatterplot matrix that lines could be drawn through the clouds of points as a way of defining directions of movement, the way to draw  these lines in higher dimensions is not obvious. Yet the line of best fit has a readily graspable geometrical intuition to it, even in higher dimensions, and that line can be diagrammed by making use of the equations of linear algebra, the field of mathematics that operates on lines in spaces of arbitrary dimensions. \index{linear algebra|(}  It is no accident that linear algebra is such a taken-for-granted part of  machine learning that its techniques for finding intersections between lines and planes, of manipulating collections of lines and surfaces through mappings and transformations (rotations, displacements or translations, skewing, and scaling), and above all, treating systems of equations using aggregate forms such as matrices and vectors. It brings with it a set of formalisations -- vector space, dimension, matrix, determinant, coordinate system, linear independence, eigenvectors and eigenvalues, inner-product space, etc. -- that machine learners resort to constantly.[^2.36]

Many of these formalisations quickly become difficult to geometrically figure. We will have reason to examine some important ways in which linear algebra structures machine learning practices in later chapters, but for the moment, it might be understood as offering a way to draw lines through spaces that can only be expressed diagrammatically in the form of equations, not in the geometrical form of figures.[^2.41b] Let us return to the equations for linear regression models a third time (remembering that both C.S. Pierce and Andrew Ng advocate returning often to simple expressions). The 'mainstay of statistics,' the linear regression model, usually appears diagrammatically in the form of linear algebra: 

\begin {equation}
\label {eq:linear_model}
\hat{Y}=\hat{\beta_0}  + \sum^p_{j=1}X_j\hat{\beta_j}
\end {equation}


\begin {equation}
\label {eq:linear_model_vector}
\hat{Y} = X_T\hat{\beta}
\end {equation}

The concision of this way of diagramming the drawing of line through a high dimensional space derives largely from linear algebra. Reading Equation \ref{eq:linear_model} from left to right, the expression $\hat{Y}$ already points to a set of calculated, predicted values, or a vector of $y$ values, such as all the `lpsa` or PSA readings included in the `prostate` dataset. Similarly, the term $X_j$ points to the table of all the other variables in the `prostate` dataset. \index{dataset!prostate} Since there are 8 other variables, and close to 100 rows, $X$ is a *matrix*  -- a higher dimensional table -- of values, addressable by coordinates. Finally $\beta_j$ are the pivotal coefficients or multiplying quantities that determine the slope or direction of the lines drawn. \index{deep learning} The second expression Equation  \ref{eq:linear_model_vector} relies more fully on linear algebra. This is the linear model written in 'vector form' [@Hastie_2009, 11]. The right hand side comprises two operations $X^T$, the transpose or rotation of the data, and implicitly -- multiplication is hardly ever shown, but diagrammed by putting terms alongside each other -- an _inner product_ of the $X$ matrix and the $\beta$ parameters (to use model talk) or coefficients (to use linear algebra talk). \index{parameters|seealso{model!coefficient}}

So, in the expression for Equation \ref{linear_model} we can begin to grasp the diagramming of a line that cannot be fully drawn in any figure, only projected onto the dimensions of a plot. In spite of these limitations, it can be readily imagined and conceptualised within the expandable dimensions of vector spaces. While that line can never fully come to light, the diagrammatics of equations \ref{eq:linear_model} and \ref{eq:linear_model_vector}  expresses a way of constructing it calculating the coefficients $\hat\beta$. Much effort, many techniques, and sub-fields of science attach themselves to various ways of constructing and making statements about such lines. The problem of calculating optimal values of the coefficients $\beta$ \index{model!coefficient} has attracted and continues to attract the attention of statisticians for a long time (at least two centuries) and today lies at the heart of cutting machine learners such as deep learning neural nets.  Such expressions are epistopic in that they connect the local complex of activities indexed as tabulated data together through the diagonal diagrammatic element of a line or plane angling through common vector space. This diagrammatic element is not necessarily reductive, although it does render differences as distances in vector space.\index{common vector space|distance} Between the statement of the data and the system of coefficients defining a fitted surface, the diagrammatic expression of the linear model creates a new kind of anamorphic or diagonal space, 'constituting hundreds of points of emergence or creativity, unexpected conjunctions or improbable continuums,' as Deleuze puts it in describing such diagrams [@Deleuze_1988,35].\index{Deleuze, Gilles!diagrams} 

# Drawing lines in a common space of transformation

The common vector space generated by the epistopic operation of machine learners has several facets. First of all, it invites mathematical analysis in terms of matrix operations and linear algebra \index{common vector space!linear algebra|seealso{linear algebra}}. In high school mathematics, at least since WWII, students have been taught how to solve systems of linear equations, first using algebra, and then using matrix operations. Solving a system of linear equations means finding those values of the variables that satisfy all the equations in the system. If such values can be found, we know that the lines expressed by equations of the form $y = ax_1 + bx_2 + c$ intersect at a point, along a line, in some sub-space. In all of these cases, diligent mathematics students solve to find the common space (shown in figure \ref{fig:common_space}), even if it consists in a single point, inhabited by the elements of the system.

\begin{figure}
  \centering
      \includegraphics[width=0.5\textwidth]{figure/intersection.pdf}
        \caption{Common space in linear equations}
  \label{fig:common_space}
\end{figure}

Similarly, although the mathematics is slightly more complicated than high school level (although not by much), drawing the line of best fit through a set of data points can be seen as solving a system of linear equations. \index{linear algebra|)} Viewed in terms of linear algebra, the 'solution' -- or 'closed form solution' -- to the linear model is given in equation \ref{eq:linear_model_solution}:

\begin {equation}
\label {eq:linear_model_solution}
\hat{\beta} = (\mathbf{X}^T\mathbf{X})^-1\mathbf{X}^T\mathbf{y}
\end {equation}

In this expression, the data shown as $\mathfb{X}$ supports the calculation of the coefficients $\hat{\beta}$ that define a flat surface or plane cutting across the common vector space.[^2.38a] The derivation of the analytical 'ordinary least squares' solution \index{ordinary least squares} relies on some differential calculus \index{calculus!differential} as well as a range of linear algebra operations such as matrix transpose, inner product and matrix inversion, the details of which need not trouble us here. (As usual, and in keeping the diagrammatic reading of these forms, these operations all consist of ways of either moving or combining numbers in practice.) The relevant point is that equation \ref{eq:linear_model_solution} calculates some values for the linear regression model coefficients $\hat\beta$ and therefore assigns a slope and intercept, a direction and displacement to a line or flat surface traversing the density-shape of a dataset in its full dimensional vector space (nine dimensions in the case of `prostate`).[^2.38] 

# `-ply`: vectorisation and the transformation of common space 

The common vector space \index{common vector space!vectorisation} appears and operates just as directly in code, especially in `R` and other programming languages designed for data practice (Octave, Matlab, Python's NumPy, or C++ Armadillo). _Vectorised_ transformations of data lie are the moving substrate of machine learning as it expands.\index{programming languages}  As usual, `R` coding practice instantiates this in miniature. As a programming language, `R` is striking for its many  *vectorised* constructs and operations. In vectorised languages such as `R`, transformations of a data structure  expressed in one line of code  simultaneously affect all the elements of the data structure. As the widely used _R Cookbook_ puts it, 'many functions [in `R`] operate on entire vectors ...  and return a vector result' [@Teetor_2011, 38]. Or as _The Art of `R` Programming: A Tour of Statistical Software Design_ by Norman Matloff puts it, 'the fundamental data type in `R` is the _vector_' [@Matloff_2011, 24], and indeed in `R`, all data is vector. There are no individual data types,  only varieties of vectors in `R`. Or, as _R in a Nutshell_  puts the point: 'in `R`, any number that you enter ... is interpreted as a vector' or as 'an ordered collection of numbers' [@Adler_2010, 17]. In these operations, all data is a vector. That is, all data operations treat data as a movement with direction and magnitude. This implicitly links data to a spatial dimension, even if it has no obvious spatial index  or reference.\index{common vector space!vector}   There are many vectorised operations in the `R` core language and many to be found in packages (the popular 'plyr' package; versions of 'ply' can also be found in recent Python data analysis libraries such as `numpy` or `pandas` [@McKinney_2012]).  In many cases, these vectorised operations occur implicitly. `R` sometimes presents difficulties for programmers trained to code using so-called procedural programming languages because it so thoroughly embraces the notion of the _vector_ -- and hence, regards all data as inhabiting vector space. In many mainstream programming languages, transformations of data rely on loops and array constructs in which some operation is successively repeated on each element of a data structure. 


```{r vectorisation, echo=TRUE, cache=TRUE, message=FALSE, warning=FALSE, comment=NA, size='smallsize', results='markup' } 
 
     vector1 <- c(0,1,2,3,4,5,6,8,9,10)
     vector2 <- c(0,1,2,3,4,5,6,8,9,10)

     #procedural programming-style looped addition
     result_looped = vector()
     for (i in vector1) {
        result_looped[i] = vector1[i] + vector2[i]
     }
     result_looped

     #vectorised addition
     result_vectorised  <- vector1 + vector2
     result_vectorised
```

The practical difference between the two approaches to moving through data is illustrated in the code listing \ref{vectorisation} above in which two 'vectors' of numbers are added together, in the first case using a classic `for`-loop construct, and in the second case using an implicitly vectorised arithmetic operation `+`. The difference between adding $1 ... 10$ using a loop or vector arithmetic is completely trivial here, but when millions of numbers are handled, these implementation differences significantly affect the work of both programmers and computing machinery.  This simultaneity is only apparent, since somehow the underlying code has to deal with all the individual elements, but vectorised programming languages take advantage of hardware optimisations or carefully-crafted low-level linear algebra libraries. More importantly, this is a different mode of movement. Operations now longer step through a series of coordinates that address data elements, but wield plans, surfaces, cross-sections or stratifications of the vector space. 

A further level of vectorisation appears in the vectorisation of functions in `R`. Specific `R` constructs such as `apply`, `sapply`, `tapply`, `lapply`, `mapply` exemplify these vectorised functions. All of the `-ply` constructs have a common feature: they take some collection of things (it may be ordered in many different ways – as a list, as a table, as an array, etc.), do something to it, and return something else. This hardly sounds startling. But while most programming languages in common use offer constructs to help deal with collections of things sequentially (for instance, by accessing each element of a list in turn and doing something with it), `R` offers ways of expressing a simultaneous operation  on  them all. The `-ply` constructs ultimately derive from the functional logic developed by the mathematician Alonzo Church in the 1930s [@Church_1936;@Church_1996]. \index{Church, Alonzo}  The functional programming style of applying functions to functions seems strangely abstract. \index{programming!functional} These -ply constructs  and the implicit vectorisation of many operations on data structures can be written concisely and tersely.  They  transform data in ways that readily adapt to and indeed motivate  increasingly parallel contemporary chip architectures, clusters of computers, reallocation of computation to GPU (Graphic Processing Unit), FPGA (Field Programmable Gate Arrays) and, perhaps most significantly,  the increasingly Cyclopean infrastructures of cloud computing and associated data centres. \index{cloud computing} Many of these condensing and expanding movements of data are diagrammed in miniature in the `R` constructs as operators in vector space. The `-ply` operations reduce both data and computational frictions. The real stake in -plying data, is not speed but transformation. -plying makes working with data less like iteration (number, text, table, list, etc.), and more like folding a pliable material. Such shifts in feeling for data are very mundane yet crucial to the epistopic movements in data. While the examples of vectorised computation that I have shown are relatively trivial --  adding sequences of numbers -- these vectorised operations exert  both infrastructural and epistemic strains for the common vector space. \index{common vector space!vectorisation!function}  

The infrastructural implication we have just seen:  the implicit and explicit vectorisation of data and operations feeds directly into the scaling up of computational  work on data. \index{common vector space|scaling} \index{common vector space|infrastructure} Not only vectorised computation, but the various styles of programming and infrastructure that lend themselves to simultaneous transformations of large matrices or tables of data lie at the kernel of not only machine learning, but of the intensive processing of data in many different places. The rendering of moving images in  digital video and in computer game animation depends implicitly on vectorised  transformations of matrices of numbers [@Mackenzie_2010a]. The more abstract implications of vectorisation and the forms of movement it encourages and  proliferates bring us back to the problem of what it means for machine learning to learn. \index{machine learning!learning}. In many respects, it entails drawing a line through the data distributed in the common vector space. In short, vectorising computation makes the common vector space, which we might understand as a resurgent form of the table, operationally concrete and machinically abstract. It is no longer a formal diagram, but a machinic process that multiplies and propagates into the world along many diagonal lines.  


# 'We fit a linear model'

'We fit a linear model' write Hastie and co-authors, referring to the `prostate` data. Models constrain movement in the fluxing dimensionality of common vector space. \index{common vector space!dimensionality} In `R` this might look like the code excerpt shown below:

```{r prostate_model, echo=TRUE}
library(ElemStatLearn)
library(xtable)
data(prostate)
columns_to_standardize = c(1,2,3,4,6,8,9)
prostate_standard = as.matrix(prostate[, columns_to_standardize]) 
prostate_standard = as.data.frame(scale(prostate_standard))
prostate_standard = cbind(prostate_standard, gleason=prostate$gleason, svi = prostate$svi, train = prostate$train)
train = prostate$train ==TRUE
prostate_model = lm(lpsa~., prostate_standard[train,-10])
tab1 = xtable(summary(prostate_model), caption = 'Fitting a linear model', label='tab:prostate_linear_regression')
print(tab1)
```

The epistopic character of Table \ref{tab:prostate_linear_regression} surfaces around the coefficients  $\hat{\beta}$  that define the direction of a flat surface running through the common vector space of the `prostate` data.[^2.101] \index{dataset!prostate} This vector is a product of operations in that space. As Hastie and co-authors write, the 'unique solution' to the problem of fitting a linear model to a given dataset using the most popular method of 'least squares' [@Hastie_2009, 12] is given by the operations we have seen in equation \ref{eq:linear_model_solution}. This tightly coiled expression calculates the parameters $\hat{\beta}$ for a linear model. The two matrices involved are the $X$ input variables (all of the `prostate` variables apart from the variable chosen as the response variable, in this case `lpsa`, the log of the PSA level) and a 1-dimensional matrix $y$, the `lpsa` values. These matrices are multiplied, transposed (a form of rotation that swaps rows for columns) and inverted (a more complex operation that finds another matrix). Epitomising the vectorised code often used in machine learning, calculating $\hat{\beta}$ for  the `prostate` data only requires one line of `R` code:

```{r beta_hat, eval = FALSE}
beta_hat = ginv(t(X) %*% X) %*% t(X) %*% y
```

The implicit vectorisation of the `R` code in the code listing \ref{beta_hat}, the fact that it already concretely operates in the common vector space, operationalizes the concise diagrammatizing of equation \ref{eq:linear_model_solution} as a machine process. \index{common vector space!vectorisation} More importantly, the vector of values that result from the vectorised multiplication, transposition and inversion of matrices creates the new vector defined by $\hat{\beta}$ whose estimated values will be subjected to the statistical tests of significance, variation, and error that we see in Table \ref{tab:prostate_linear_regression}. We will have occasion to return to these statistical estimates (in chapter \ref{ch:probability}), since the play of values that starts to appear even in just fitting one model will become much more significant when fitting hundreds or thousands of models, as some machine learners do.[^2.300] 


# The vectorised table?

>The table has the function of treating multiplicity itself, distributing it and deriving from it as many effects as possible [@Foucault_1997, 149] \index{table} \index{Foucault, Michel}

In  *Surveiller et Punir* or *Discipline and Punish* [@Foucault_1977], Foucault returned to the question of the table in a slightly different context: an account of the operation of power in disciplinary institutions and knowledges. In these knowledges, the table becomes generative of 'as many effects as possible.' A version of this generative function appears in the machine learning shown by equations \ref{eq:linear_model} and  \ref{eq:linear_model_solution} and in the code listing \ref{beta_hat}.  In the construction of the common vector space in linear algebra, and then its practice in code rather than solely in the analytical space of equations, machine learning opens new volumes or dimensionalities that could not be seen, or rendered in either tables of data or graphic figures of data. Every   machine learner inhabits and moves through the common vector space. Sometimes their operations flatten the vector space down into lower dimensions, sometimes they expand the vector space into a great many new dimensions (as we saw with 'dummy variables' that embody categories, and as we will see with support vector machine classifiers in a chapter \ref{ch:pattern}). 

The generalized application of machine learning relies on and reinforces a broad, powerful yet quite subtle transformation of data into vectors and a vector space. This remaps the grid of the table into the expanding dimensions of the common vector space. This is not the first such expansion of the table. We need only think of the relational database systems of the late 1960s, and their multiplication of tables [@Mackenzie_2013b]. But perhaps in the vectorised and matrix-form practices of the  common vector space, machine learners produce for the first time a quasi-tabular or meta(s)table volume that cannot be surfaced on a page or screen, only diagrammatically traversed through vectorised operations.  \index{table} \index{common vector space!metatable}

The epistopic transformation of the table pulls and re-aligns communication and infrastructures. It acts as a powerful tensor on knowledges and operations of many different kinds since it transposes, inverts, and re-maps local complexes of activity. In following what happens to vectors, lists, matrices, arrays, dictionaries, sets, dataframes, or series or tuples in data, we might get a sense of how the epistemic operations of predictive models, the supervised and unsupervised learners, the classifiers, the decision trees and the neural networks fold and refold matrices and vectors in a common vector space. The  challenge lies in attending to this expansive data vectorisation in ways that maintain and indeed augment its value-relevance, its concreteness, and attachments to lives and places. We lack good intuitions of how to do that because of the ways in which data as vector space has been constructed and described. \index{machine learning!epistopic}

Does the act of unwinding some of these operations, for instance, by seeing how  matrix multiplication ripples through different treatments of data in a linear regression model (even such an small one as the `prostate` data), bring us closer to the vitality of the multiple/multiplying concrete value-situations that connect calculation?  My suggestion is that the act of diagramming how machine learners  vectorise data densities  begins to locate and unravel the processes of knowing, predicting and deciding on which many aspects of the turn to data rely.  As we will see, the vectoral operations we have just been viewing are themselves organised by other layers of diagrammatic movement \index{diagrammatic movement} that shape surfaces in more convoluted forms.    

What is at stake in the common vector space? The broad stake, at the risk of over-emphasising it, is the production of new kinds of realities occasioned by the mobility of machine learners. \index{machine learner!mobility of}  The machine learner recognising zip codes might be found in the navigation system of an autonomous vehicle or the methods section of a paper published in *Nature.* The datasets that _Elements of Statistical Learning_ ranges sides by side evince this improbable diagonal movement. 


[^2.38]: As we will see in the following chapter (chapter \ref{ch:function}), it is not always possible to calculate the parameters of a model analytically. Especially in relation to contemporary datasets that have very many variables and many instances (rows in the table), linear algebra approaches become unwieldy in their attempt to produce exact results, and machine learning steps in with a variety of computational optimisation techniques. \index{coefficients}

[^2.38a]: Perhaps more importantly, the linear algebraic expression of these operations presupposes that all the data, both the values used to build the model and the predicted values the model may generate as it is refined or put into operation somewhere, are contained in a common space, the vector space, a space whose formation and transformation can be progressively ramified and reiterated by various lines that either separate volumes in the space, or head in a direction that brings along most of the data. Not all of these lines are bound to be straight, and much of the variety and dispersion visible in machine learning techniques comes from efforts to construct different kinds of lines or different kinds of 'decision boundaries' (in the case of classification problems) in vector space (for instance, the k-nearest neighbors method does not construct straight lines, but somewhat meandering curves that weave between nearby vectors in the vector space; see [@Hastie_2009, 14-16]). Whether they are straight or not, the epistopic aspect of these lines remains prominent. Typically, many different statistical tests (Z-scores or standard errors, F-tests, confidence intervals, and then prediction errors) will be applied to any estimate of the coefficients of even the basic linear regression model, well before most advanced or sophisticated models and techniques (cross-validation, bootstrap testing, subset and shrinkage selection) begin to re-configure the model in more radical ways. \index{statistics!tests}


[^2.10]: This kind of data will be the focus of a later chapter \ref{ch:genome}. Machine learning during the 1990s and 2000s was in some ways boosted heavily by the advent of genomic biology with its large, enterprise style knowledge endeavours such as the human genome project.

[^2.11]: The data derives from a publication in _nature genetics_ [@ross_2000] analysing gene expression in the cancer cell lines maintained as experimental models by the us national cancer institute.

[^2.41]: Learning machine learning, and learning to implement machine learning techniques, is largely a matter of implementing series of matrix multiplications. As Andrew Ng advises his students,

    > Almost any programming language you use will have great linear algebra libraries. And they will be high optimised to do that matrix-matrix multiplication very efficiently including taking advantage of any parallelism your computer. So that you can very efficiently make lots of predictions of lots of hypotheses [@ng_2008a, 10:50]\index{Ng, Andrew}

    In other parts of his teaching, and indeed throughout the practice exercises and assignments, Ng stresses the value of implementing machine learning techniques for both understanding them and using them properly. But this is one case where implementation does not facilitate learning. Ng advises his learners against implementing their own matrix handling code.    They should instead  use the 'great linear algebra libraries' found in 'almost any programming language.' 'linear algebra libraries' multiply, transpose, decompose, invert and generally transform matrices and vectors. \index{linear algebra} they will be 'highly optimised' not because every programming language has been prepared for the advent of machine learning on a large scale, but rather more likely because matrix operations are just so widely used in image and audio processing.  Happily, Ng observes, that means that 'you can make lots of predictions' [@Ng_2008e].   It seems that generating predictions and hypotheses outweighs the value of understanding how things work on this point. \index{prediction}


[^2.351]: One sign of the centrality of the line in machine learning can be seen, for instance, from the contents page of the book [@hastie_2009, xiii-xxii]. After the introduction of the linear model in the first chapter and its initial exposition in chapter 2 ('overview of supervised learning'), it forms the central topics of chapter 3 ('linear methods for regression'), chapter 4 ('linear methods for classification'), chapter 5 ('basis expansions and regularization'), chapter 6 ('kernel smoothing methods'), much of chapter 7 ('model assessment and selection'), chapter 8 ('model inference and averaging'), major parts of chapter 9 ('additive models, trees and related methods'), important parts of chapter 11 ('neural networks' -- neural networks can be understood as a kind of regression model), the anchoring point of chapter 12 ('support vector machines and flexible discriminants') and the main focus in the final chapter ('high dimensional problems'). A similar topic distribution can be found in Andrew Ng's Cs229 lectures on machine learning. More than half of the 20 lectures concern linear models and their variants. See [@Ng_2008a; @Ng_2008b;@Ng_2008c;@Ng_2008d].


[^2.36]: Along with statistics and probability, linear algebra is a such an important part of machine learning that many books and courses recommend students complete a linear algebra course before they study machine learning. Cathy O'Neill and Rachel Schutt advise:
    When you’re developing your skill set as a data scientist, certain foundational pieces need to be in place first—statistics, linear algebra, some programming [@Schutt_2013, 17]

[^2.31]: Despite the intensive work that Hastie and co-authors conduct on the `prostate` data, all with a view to better predicting PSA levels using volumes and weights of prostates, etc., Stamey and other urologists more than a decade or so concluded that PSA is not a good biomarker for prostate cancer. \index{dataset!prostate} Stamey writes in 2004:
    What is urgently needed is a serum marker for prostate cancer that is truly proportional to the volume and grade of this ubiquitous cancer, and solid observations on who should and should not be treated which will surely require randomized trials once such a marker is available. Since there is no such marker for any other organ confined cancer, little is likely to change the current state of overdiagnosis (and over-treatment) of prostate cancer, a cancer we all get if we live long enough unbound points in the matrix [@Stamey_2004, 1301]
\index{dataset!prostate}
[^2.33]: Carl Friedrich Gauss \index{Gauss, Carl Friedirch} and Adrien-Marie Legendre's work on linear regression at this time is well-known. The first independent use of linear regression was Gauss' prediction of the location of an 'occluded volume,' the position of the asteroid Ceres after it reappeared in its orbit behind the sun. [@Stigler_2002] -- TBA page ref


[^2.1]:  Later chapters will discuss various ways in which the vectoral dimensionality of data or its rendering as _vector space_ scales up and scales down in machine learning. In terms of multiplying matrices, dimensionality both constrains and enables many aspects of the prediction. Perhaps on the grounds of data dimensionality alone, we should attend to dimensionality practices in  `R`. A fuller discussion of dimensionality of data is the topic of chapter \ref{ch:pattern}. There I discuss how machine learning re-dimensions data in various ways, sometimes reducing dimensions and at other times, multiplying dimensions.

[^2.22]: By and large, I am not discussing networking and database infrastructures here. In other work, I have attempted to account for the multiplication of tables in databases but also in mundane practices of ordering. See [@Mackenzie_2011; @Mackenzie_2012]. My analysis there pivots mainly on a set-driven account of data derived from the work of Alain Badiou. \index{Badiou, Alain} Although I've already been using some set terminology here -- as in 'dataset' -- I'm persisting with a more geometrical and algebraic account of datasets in order to better deal with some of the graphisms common in machine learning. 

[^2.41b]: We might add also approach the epistopic fault line in machine learning topologically \index{topology}. Over a decade ago, the cultural theorist Brian Massumi wrote that 'the space of experience is really, literally, physically a topological hyperspace of transformation' [@Massumi_2002, 184] \index{Massumi, Brian}. Much earlier, Gilles Deleuze had conceptualised Michel Foucault's philosophy as a topology, or 'thought of the outside' [@Deleuze_1988], as a set of movements that sought to map the diagrams that generated a 'kind of reality, a new model of truth' [@Deleuze_1988, 35]. More recently, this topological thinking has been extended and developed by Celia Lury amongst others. In 'The Becoming Topological of Culture,' Lury, Luciana Parisi and Tiziana Terranova suggests that 'a  new rationality is emerging: the moving ratio of a topological culture' [@Lury_2012, 4] \index{{Lury, Celia}. \index{Parisi, Luciana} \index{Terranova, Tiziana} In this new rationality, practices of ordering, modelling, networking and mapping co-constitute culture, technology and science [@Lury_2012, 5].  At the core of this new rationality, however, lies a new ordering of continuity. The 'ordering of continuity,' Lury, Parisi and Terranova propose, takes shape 'in practices of sorting, naming, numbering, comparing, listing, and calculating' (4). The phrase 'ordering of continuity' is interesting, since we don't normally think of continuities as subject to ordering. In many ways, that which is continuous bears within it its own ordering, its own immanent seriation or lamination. But in the becoming topological of culture, movement itself undergoes a transformation according to these authors. Rather than movement as something moving from place to place relatively unchanged (as in geometrical translation), movement should be understood as more like an animation, a set of shape-changing operations. These transformations, I would suggest, should be legible in the way that machine learning, almost the epitome of the processes of modelling and calculation that Lury, Parisi and Terranova point to, itself moves through the data. And indeed, the juxtaposition of spam, biomedical data, gene expression data and handwritten digits already suggests that topological equivalences, and a 'radical expansion' of comparison might be occurring. Bringing epistopics and topologies together might, I suggest, help trace, map and importantly diagram some of the movements into the data occurring today.

[^2.101]: From the epistopic viewpoint, the most obvious result of fitting a linear model is the production not of a line on a diagram or in a graphic. As we have seen, such lines cannot be easily rendered visible. Instead, the model generates a new table of coefficients (see Table \ref{tab:prostate_linear_regression}) and some new numbers, _statistics_.  This table is not as extensive as the original data, the $\mathbf{X}$ and $Y$ vectors. But the names of the variables in the dataset appear as rows in the new table, a table that describes something of how a line has been fitted by the linear model to the data. The columns of the table now bear abbreviated and much more statistical names such as `estimate` (the estimated values of $\hat{\beta}$, the key parameters in any linear model), `Std. Error`, `t value`, and the all important _p_ values written as `Pr(|t|)`.  The numerical values ranging along the rows mostly range from -1 to 1, but the final column includes values that are incredibly small: `1.47e-06` is a few millionths. Other statistics range around the outside the table: the `F-statistic`, the `R-squared` statistic, and the `Residual standard error`. I don't propose discussing these in any great detail here. No matter how we understand these statistics, they constitute a transformation of the `prostate` dataset. The numbers of the \ref{tab:prostate} become epistopic here, since they now appear as a set of standard errors, estimates, t-statistics, and _p_ values, that together indicate how likely the estimated values of $\beta$ are, and therefore how well the diagonal line expresses the relations between different dimensions of the dataset in the common vector space. How have these statistics, which all act as qualifications and qualifications on the line whose direction and point of entry is given the $\hat{\beta}$ coefficents, arisen? We seem to have crossed some threshold between drawing a line through an occluded data volume and generating propositions about the way that line passes through the data.


[^2.300]: This is an important differentiation: it is not typical machine learning practice to construct one model, characterised by a single set of statistics (F scores, R^2 scores, _t_ values, etc.). In practice, most machine learning techniques construct many models, and the efficacy of some predictive techniques derives often from the multiplication or indeed proliferation of models. Techniques such as neural networks, cross-validation, bagging, shrinkage and subset selection, and random forests, to name a few, generate many statistics, and navigating the multiple or highly variable models that result becomes a major concern. An epistopic abundance will appear here -- bias, variance, precision, recall, training error, test error, expectation, Bayesian Information Criteria, etc. as well as graphisms such as ROC (Receiver-Operator-Characteristics) curves. Put simply, the proliferation of models start to drive the dimensional expansion of the common vector space. At the same time, the multiplicity of models multiplied by the machine learners  becomes the topic of statistical analysis. 


[^2.104]:   I loosely borrow the term 'density' from statistics, where *probability density functions* are often used to describe the hardly ever uniform distribution of probabilities of different values of a variable. Sensing density as a form of variation matters greatly both in machine learning itself, where algorithms seek purchase on unevenly distributed data, and in any broader diagram of data. Probability densities are discussed in much more detail in Chapters \ref{ch:probability}. The collection _"Raw Data" is an Oxymoron" [@Gitelman_2013] evinces some of these different densities and distributions of data.  
