\chapter{Vectorisation and its consequences}
\label{ch:vector}

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, tidy=TRUE, fig.height=8, 
                      echo=FALSE,  fig.show='hide', results='hide', warning=FALSE, message=FALSE, dev='pdf')
library(RSQLite)
con <- dbConnect(RSQLite::SQLite(),'../ml_lit/all_refs.sqlite3')
library(ggplot2)
library(stringr)
library(xtable)
options(xtable.comment = FALSE)
options(xtable.size = 'tiny')
options(xtable.tableplacement = '!htbp')
options(xtable.rownames = FALSE)
simpleCap <- function(x) {
    s <- strsplit(x, " |-")[[1]]
    return (paste(toupper(substring(s, 1,1)), tolower(substring(s, 2)), sep="", collapse=" "))
}
```

>The table has the function of treating multiplicity itself, distributing it and deriving from it as many effects as possible [@Foucault_1997, 149] \index{table} \index{Foucault, Michel}

>We call _any_ set that satisfies these properties (or axioms) a *vector space*, and the objects in the set are called *vectors*. [@Larson_1996, 166] \index{vector} \index{vector space|see also {mathematics!linear algebra}}

>All things are vectors [@Whitehead_1960,309]\index{Whitehead, Alfred North!on vectors}

The operational power of  machine learning locates data practice in an expanding epistemic space. The space derives, I will suggest, from a  specific operational diagram that maps data into  a \gls{vector space}. It  \textit{\glspl{vectorize}} data according to axes, coordinates, and scales. Machine learning, in turn, inhabits a vectorised space, and its  operations vectorise data.  \index{data!vectorization|see {vector space!vectorization}}

Often data is  represented as an homogeneous mass or a continuous flowing stream. My aim here, however, is to archaeologically examine some of the transformations that allow different shapes and densities of data, whether in the form of numbers, words or images, to become machine learnable. Data in its local complexes spaces out in many different density shapes, depending on how the data has been generated or instanced.[^2.104] \index{data!density of} Whatever the starting point -- a measuring instrument, people clicking and typing on websites, a device like a camera, a random number generator, etc. --  machine learners only ever encounter data in specific *vectorised* shapes (vectors,  matrices, arrays, etc.), mapped to  a geometrically coordinate volume. The mapping and forming, when mentioned at all,  is sometimes referred to as 'data cleaning' but that term covers over important but largely taken for granted and constitutive transformations. \index{data!cleaning} The archaeology of data shapes presented in this chapter explores a range of transformations focused around the table or row-column grid. \index{archaeology!of tables}

The reshaping and re-flowing of data densities into vectors deeply affects machine learning.   This forming and reforming of data is evidence of  implicated relations. The philosopher A.N. Whitehead called 'strain':

>a feeling in which the forms exemplified in the datum concern geometrical, straight, and flat loci will be called a "strain." In a strain, qualitative elements, other than the geometrical forms, express themselves as qualities implicated in those forms [@Whitehead_1960, 310]. \index{Whitehead, A. North}  \index{data!strain}

In many machine learning models the exemplified forms are straight or flat loci (as we see in chapter \ref{ch:function}).  Yet  different practices also elicit relations that strain the linear shaping of data and these divergent relations sometimes combine in generative and provocative ways. 

#  Vector space and geometry

Statistical modelling, data-mining, pattern recognition, recommendation systems, network modelling and machine learning rely very much on the operation called 'fitting a model.' \index{model!fitting} Fitting as a spatial practice has some elements that resemble the phenomenologist Edmund Husserl's account of the origin of geometry. Husserl writes:

>First to be singled out from the thing-shapes are surfaces -- more or less "smooth," more or less perfect surfaces; edges, more or less rough or fairly "even"; in other words, more or less pure lines, angles, more or less perfect points; then again, among the lines, for examples, straight lines are especially preferred, and among surfaces, the even surfaces. ... Thus the production of even surfaces and their perfection (polishing) always plays its role in praxis [@Derrida_1989, 178] \index{Husserl, Edmund!on thing-shapes in geometry}

Husserl here refers is attempting to describe something of the way in which forms such as planes, lines, circles, triangles, squares, and points became objects of geometrical practice. A similar polishing and smoothing of surfaces is certainly taking place today in the thing-shapes we call data. The basic machine learning work of 'fitting a model' (or many models)  to data is often literally implemented, as we will see, by  constraining data within a coordinate, discretized space, which I term the \textit\gls{vector space}. \index{vector space} 

Critical thought from phenomenology to social theory has a long-standing  nervousness about the power of geometry and its gradual movement away from shapes and things towards mathematical operations.   The philosopher Hannah Arendt, for instance, observes: 

>decisive is the entirely un-Platonic subjection of geometry to algebraic treatment, which discloses the modern idea of reducing terrestrial sense data and movements to mathematical symbols [@Arendt_1998,265] \index{Arendt, Hannah!on geometry and algebra}

The crux of the problem rests on the 'treatment' or operations that 'reduce terrestrial sense data and movements' to symbols. One challenge for contemporary thought is how to orient itself to such operations, particularly in their data-intensive forms,  without assuming that the familiar story of scientific and technical reduction of sense and movement to the lines and planes of modern geometry is simply reinforced in contemporary data practice. \index{critical thought!relation to geometry} If an archaeology of data vectorization does anything, it needs to offer an alternative to that reduction.  

They also, as we will see, reach down into the practices of programming, infrastructure and hardware production in ways that differ somewhat from increases in computational power or speed. Familiar narratives of Moore's Law increases in  speed or efficiency of computation do not account for transformations of   data in  `R` and other computing environments (for instance, the popular `Map-Reduce` architecture invented at Google Corporation to speed up its search engine services [@Mackenzie_2011] \index{data!architecture!map-reduce}).  Vectorisation transforms data along more diagrammatic lines.[^2.1]  

# Mixing places

```{r elemstatlearn_data_2, echo=FALSE, results = "asis"}
library(ElemStatLearn)
library(xtable)
datasets_results <-data(package='ElemStatLearn')$results[,3:4]
datasets = as.data.frame(datasets_results)
dataset_count <- nrow(datasets)
dataset_table = xtable(datasets, caption='Datasets in \\textit{Elements of Statistical Learning}', label = 'tab:datasets_elemstatlearn', table.placement = '!ht')
print.xtable(dataset_table, type='latex')
```

Data appears in _Elements of Statistical Learning_ in multiple forms. \index{\textit{Elements of Statistical Learning}!datasets in} Maps of New Zealand fishing patterns lie next to plots of factors in South African heart disease. The `r dataset_count` datasets  shown in table \ref{tab:datasets_elemstatlearn} typify the variety found in [@Hastie_2009]. They span scientific, clinical, commercial and media fields. Note that they include many patches and pathways of everyday life -- speaking, seeing, writing and reading -- as well as specific scientific objects of knowledge such as galaxies, cancer, climate and national economies. This mixture is not unusual for machine learners.  To give another example, the statistician Leo Breiman's 2001 article on random forests [@Breiman_2001], \index{machine learner!random forest} perhaps the most cited journal paper in the machine learning literature, displays a similarly diagonal line across human and non-human worlds: \index{Breiman, Leo}

>Glass,  Breast cancer,  Diabetes,  Sonar,  Vowel,  Ionosphere,  Vehicle,  Soybean,  German credit,  Image,  Ecoli,  Votes,  Liver,  Letters,  Sat-images,  Zip-code,  Waveform,  Twonorm,  Threenorm,  Ringnorm,  [@Breiman_2001,12].

In some ways, the improbable conjunction of spam email and cancer detection in machine learning continues what statistics as a field has always done: rove  across scattered fields ranging from astronomy to statecraft, from zoology to epidemiology, gleaning data as it goes (see [@Stigler_1986; @Hacking_1990] for samples of itineraries). \index{statistics} 

In _Elements of Statistical Learning_ and the field of machine learning more generally, something more is moving through and coordinating these datasets. The coordination might be rhetorical: the colligation of datasets -- vowels, ozone, bone density, marketing, prostate cancer, and spam -- in all their diversity  suggests the mobility of machine learners. The combination of datasets deriving from network media, from medicine, from business administration and from cutting-edge life science (c.2000) suggests a tremendous, indeed almost spectacular miscibility, one that in principle could surprise us because there is otherwise little mixing between the settings and knowledge domains these datasets come from. How is this mixing, conformation and homogenisation being done? The repetition of data sets, the juxtaposition of discontinuous domains, and forms of movement construct and order continuities in the service of various forms of predictive and inferential knowledge. The miscible juxtapositions we encounter in machine learning enter into, it seems, a _regularity_ or a common space, a space that displays strong tendencies to expand, accumulate and archive relations. Rather than peripatetic learning, the accumulation of diverse datasets attests to a prior ordering of data to afford its traversal. 


The practices of naming and ordering, the sorting of different data types, the addressing and expansion of data  exemplified in these diverse datasets can tell us a lot about how machine learning  learns from data. \index{learning!from data}  The shapes, compositions and loci formed  from the datasets enable the functioning of machine learners as they operate to generate statements, classifications and decisions.  If machine learning can be understood as a constantly evolving diagram of practical abstractions,  the way it draws on and relates to different forms of data matter.  As we will see, machine learners transpose data in an increasingly extensive, heavily coordinated space, the vector space.  


# Truth is no longer in the table?

'This book is about learning from data' write Hastie, Tibshirani and Friedman on the first page of _Elements of Statistical Learning_, as they rapidly begin to iterate through some datasets. \index{\textit{Elements of Statistical Learning}!on learning from data} On the second page of the book, a table of spam email word frequencies appears (and the problem of spam classification is canonical in the machine learning literature - we return to in chapter \ref{ch:probability}). \index{dataset!spam} They come from the dataset `spam` [@Cranor_1998]. On the third page, a complicated data graphic appears (Figure 1.1, [@Hastie_2009, 3]. It is a scatterplot matrix of the `prostate` dataset included in the `R` package `ElemStatLearn,` the companion `R` package for the book. In a third example, a set of scanned handwritten numbers appears. These scans are images of zipcode or postcode numbers written on postal envelopes taken from the dataset `zip` [@LeCun_2012], and they differ from both the `spam` table and `prostate` plots because they directly resemble something in the world, which, however, happens to be numbers, and is, therefore, probably already recruited into data-making and data-circulating processes (a dataset we return to in chapter \ref{ch:subjects}).\index{dataset!\texttt{zip}}  The final example in the introduction, 'Example 4: DNA Expression Microarrays,' draws from biology, and particularly, high-throughput genomic biology, a science that produces large amounts of data about biological processes  by running many tests, or by constructing devices that generate many measurements, in this case, a DNA microarray.[^2.10]


The table or the row-column addressable grid is common to all of these datasets. And yet, as we are about to see, machine learning in many ways deals with the collapse or liquidation of tabular datasets. 'Things, in their fundamental truth,' writes Foucault in _The Order of Things_ 'have now escaped from the space of the table' [@Foucault_1992, 239]. \index{Foucault, Michel!on table} \index{table!history |(} Foucault  writes in these pages about the fabled emergence of life, labour and language as the anchoring vertexes of a new triangle of knowledge and power structuring the figure of the 'human' in the 19th century.

Before the emergence of the characteristic sciences of the human -- political economy, linguistics and biology -- knowledges such as natural history, the general grammars, and philosophies of wealth (such as Adam Smith's work)  had ordered empirical materials of diverse provenance in tables or grids. While the history of tables as data forms reaches a long way back (see [@Marchese_2013] for a broad historical overview that reaches back to Mesopotamia),  Foucault argues that the Classical age first developed the system of grids that permitted ranking, sorting and ordering in tables. These grids replace the Renaissance tabulations based on 'buried similitudes' and 'invisible analogies' [@Foucault_1992, 26]. In pre-Classical tables, an image or figure from myth might lie alongside a measurement or a count of occurrences, and this proximity was ordered by systems of analogical association that spanned what we might today, in the wake of the 19th century, might see as incongruous (for instance, associations between medicine and Biblical prophecy). 

The Classical Age grid or table, by contrast, brought plural and diverse resemblances into exhaustive systematic visible enumeration. As Foucault puts it:

> The space of order, which served as a _common place_ for representation and for things, for empirical visibility and for the essential rules, which united the regularities of nature and the resemblances of imagination in the grid of identities and differences, which displayed the empirical sequence of representations in a simultaneous table, and made it possible to scan step by step, in accordance with a logical sequence, the totality of nature's elements thus rendered contemporaneous with one another [@Foucault_1972, 239]

The table as space of order did not stand in isolation. It served a localized epistemic function in conjunction with other of knowledge such as experiment and mathematical proof. Since algebra or experimental _mathesis_ only applied to 'simple natures' (planets in movement, dynamics of falling bodies, etc.), \index{mathematics!application to nature} table-based knowledges such as taxonomy dealt with more complex natures. In the tables, systems of signs -- for instance, the groupings established by the eighteenth century taxonomist Carl Linnaeus -- sought to reduce complex natures (plants, animals, etc.) to simpler forms as columns and rows in a table based on resemblances and similarities. \index{Linnaeus, Carl} \index{differences!taxonomy of}  Importantly, the table as space of order was a space of imagination in that one could begin to see continuities and differences between things (organisms, words, nations)  by carefully ordering and scanning the table. 'Hedged in by calculus and genesis,' Foucault suggest, 'we have the area of the _table_' [@Foucault_1992, 73]. \index{calculus|see{mathematics!calculus}} Note in passing that calculus  and calculations bound the table only in relation to 'simple natures' whose identity and difference can be understood in the form of movements, rates and change in position.[^2.333] This is an important limitation since it is precisely the complex natures in genesis that machine learning tries to engage using algebra, calculus, statistics and computation.


[^2.333]: In  *Surveiller et Punir* or *Discipline and Punish* [@Foucault_1977], Foucault returned to the question of the table in a slightly different context: an account of the operation of power in disciplinary institutions and knowledges. In these knowledges, the table becomes generative of 'as many effects as possible.' \index{Foucault, Michel!tables!in disciplinary power}

Finally (at least for our purposes), in the nineteenth century, a different form of ordering shattered tabulation based on enumerated similarity and ordered resemblance. Foucault figures this change as shattering the table into shards of order:

>this space of order is from now on shattered: there will be things, with their own organic structures, their hidden veins, the spaces that articulates them, the time that produces them; and then representation, a purely temporal succession, in which those things address themselves (always partially) to a subjectivity [@Foucault_1992, 239-240].

Life, labour and language -- Foucault's famous historically emergent triadic figure of the human  -- replace the enumerative, synoptic classificatory tables of the Classical age. Tables still abound in newer temporal, genetic orderings (almost any episode from the history of nineteenth and twentieth century statistics will confirm that; see [@Stigler_1986; @Stigler_2002])\index{statistics!history}, but from now on tables are localized, relating to a place and changing in time, and functioning as shards of representational order addressed to (and often constitutive) of subject position. \index{subject position!human!historical constitution of} A double interiority takes over. Things such as a language, a species or an economic system have their own genesis, their own temporality and historical existence. Our knowledges and indeed experience of them also become finite, historical, with their own dynamics and internal life. In this change, the table itself is no longer the foundation or distillation of knowledge. It is one knowledge apparatus amongst many. \index{table!history|)} Tables, we might say, become merely data, inscriptions pendant on hidden structures and their genesis. \index{data!tables become}

This brief résumé of a thread of argument in _The Order of Things_ might help us reassess how machine learning  goes into the data. The tables of `spam`, `prostate`, `image` and `microarray` data in contemporary machine learning do not operate in the way Linnaeus would have tabulated a table of living things based on similarities and resemblances or a Renaissance medical textbook might assemble analogical resemblances between disease and astronomy. As I will argue, measures of similarity and resemblance still operate strongly in machine learning as it moves through tables. In this respect, the Classical table and perhaps even the pre-Classical table returns with extended relevance.  The repeated juxtaposition of tables of diverse provenance alongside each other, the incorporation of 'complex natures,' and the  operational superimposition of different tables suggests machine learners still suggest synoptic alignments, but along different lines than the row-column grid of the Renaissance, Classical or even the later life-labour-language tables of the human sciences.  Even if the tables in the opening pages of _Elements of Statistical Learning_ concern work, life, language and economy and  map very readily onto the anchor points, the new 'empiricities' (Foucault's term for the empirical problems),  of labour, life and language or biopower \index{biopower} that took root at this time [@Foucault_1991], the ways machine learners traverse them may not be recognisable in terms of these empiricities. \index{empiricities} 

Instead,  we are now confronted by kaleidoscopically transmuted tables whose expansion and open margins afford many  formulations of similarity and difference. Already in the handwritten digits and the microarray data, scale and dimensions thwart tabular display of the data.   In settings such as social media platforms or genomics in which machine learning operates, tables change rapidly in scale and sometimes in organisation.  The multiplication and juxtaposition of tables suggests that we might be seeing the advent of a post-order space for regularities and resemblances, for simple and complex natures, encompassing knowledges ranging from humanities to traffic engineering.  

# The epistopic fault line in tables

```{r ElemStatLearn_data, echo=FALSE}
library('ElemStatLearn')
datasets = data(package='ElemStatLearn')
datasets[['results']][,3]
```

The `ElemStatLearn` `R` package brings, as we have seen in the previous chapter, with it around 20 different datasets, including the four mentioned in the introduction to _Elements of Statistical Learning_.  On the one hand, every data table indexes a localised complex of activities (clinical research, social media platform, financial transactions, etc.) with possible referential importance.\index{referentiality}  On the other hand, for machine learners, the table is a space of potential similarities and differences  both internal to the table itself (e.g. how much does row number 1000 differ from row 1,000,000) and associated with other tables (e.g. how much does this table of clinical test relate to that table of microarray data?). \index{differences!between datasets}  These internal and external differences entwine with each other in ways that create a fault line, an unstable yet generative line of diagonal movement. It is this fault or fold line, I propose, that diagrammatically distributes data tables into the expansive and moving substrata of the  vector space. \index{diagrammatic movement} 

The table has been vectorised. We might say that the vectorization of the table is epistopic.   The term 'epistopic' \index{epistopic} comes from the work of the science studies scholar Mike Lynch \index{Lynch, Mike!epistopics} whose account of scientific practice is particularly focused on ordinariness\index{practice!scientific}. Akin to what   Foucault in the _Archaeology of Knowledge_  terms a threshold of epistemologization [@Foucault_1972, 195] \index{epistemologization!threshold of} Lynch characterizes the 'epistopic' as connecting  localized practices ('topics')  with 'familiar themes from epistemology and general methodology' in the local achievement of coherence in knowledge [@Lynch_1993, 280]. \index{knowledge!local coherence of} In other words, as the term itself suggests, an epistopic connects general epistemic themes such as validity, precision, specificity, error, confidence, expectation, likelihood, uncertainty, or approximation with a place, a 'local complex of activities' (281). \index{machine learning!statistical aspects}  This emphasis on epistemic location frames the problem of how  the 'local complex' of a specific dataset encounters a generalizing epistemic practice such as machine learning.   \index{epistemology}

# Surface and depths: the problem of volume in data 

The local complex of activities for associating datasets shatters tables from a different direction than that described by Foucault in _The Order of Things_. Algebra,  linear algebra in particular, organizes and distributes differences in vector space. _Mathesis_ in the form of algebraic operations of addition and multiplication of  collections of tabular elements such as rows and columns, now re-defined as vectors, re-structure tabular data as a vector space, as a 'set' whose membership is only limited by the applicability of the constructing relations. These operations absorb and subtend differences in quality, type, kind and quantity. \index{vector space!ramification in}

```{r prostate_plot, echo=FALSE, message=FALSE, results = 'hide', fig.show='hide'}
library('ElemStatLearn')
data('prostate')
pairs(prostate[, -10],  cex=0.2)
```

```{r prostate_data, results='asis'}

library(xtable)
print(xtable(head(prostate, 3), label= 'tab:prostate', caption='First rows of the \`prostate\` dataset'))
```

\begin{figure}
  \centering
      \includegraphics[width=0.9\textwidth]{figure/prostate_plot-1.pdf}
        \caption{Scatter plot matrix of \texttt{prostate} data}
  \label{fig:prostate_plot_matrix}
\end{figure}

Of the three example datasets  (`prostate`, `spam` and `zip`), _Elements of Statistical Learning_  returns most frequently to `prostate.` This dataset derives from the work of urologists working at Stanford [@Stamey_1989], and concerns various clinical measurements performed on men who were about to undergo radical prostatectomy. The measurements range across the volume and weight of the prostate, as well as levels of various prostate-related biomarkers such as PSA -- prostate specific antigen. Several rows from the dataset are shown in Table \ref{tab:prostate}. The first pages of the book had already exhaustively plotted all the variables in the dataset against each other using the table-related form of a scatter plot matrix [@Hastie_2009,3] (shown in figure \ref{fig:prostate_plot_matrix}, and they return to the same data on almost a dozen occasions in the course of the book, subjecting it to repeated vectorization. \index{dataset!prostate} \index{graphic!scatterplot matrix}

The contrast between the Table \ref{tab:prostate} and the Figure \ref{fig:prostate_plot_matrix} already depends on a transformation intrinsic to vectorization.  On the one hand, the table arrays all different data types in rows and columns. In the table, the relation between the different data types (the log of the weight of prostate - `lwp` and `age` for instance) is quite hard to see. Moreover, different kinds of variables stand side by side. `svi`, short for 'seminal vesicle invasion' is a categorical variable. \index{data!type!categorical} It takes the values 'true' or 'false,' shown here as `1` or `0`, but the other variables either measure or count things (years, sizes, or levels of antigens). 

On the other hand, the scatter plot matrix also takes the form of a grid-like figure, but the cells of the grid are not occupied by numbers but by `x-y` plots of different pairs of variables in the `prostate` dataset. The 'matrix' of figures shows of 72 plots is mirrored across the diagonal that runs from the top-life to bottom right in figure \ref{fig:prostate_plot_matrix}. Taking this folding into account, we see 36 unique plots with different data in each one. Each sub-plot displays the relation between two variables in the dataset as a scatter plot. Some variables such as `svi` are not very amenable to plotting in this way. More importantly perhaps, certain combinations of variables appear as flat loci that can be read as signs of relations between different variables. The scatter plot matrix constructs a tabular space in which relational contrasts between pairs of variables start to appear. In the light of these contrasts (and I use 'light' here in an almost literal sense to refer to the way in which the architecture of the figure creates a space in which light scatters in varying patterns), the `prostate` dataset begins to expose relations that might be worth knowing about. We have moved on from the bare table of the dataset to a transformed tabulation, from a textual-numerical grid to a geometrical-numerical grid. Everything remains on the surface of a grid here, but the grid permits differences in relationality to begin to appear. \index{differences!visibility in data}

All of this somewhat precedes the operational formation of machine learning. Similar tables and plots are part and parcel of statistical data exploration more generally (see [@Beniger_1978] for an historical account of quantitative graphics in statistics). \index{statistics!graphics| seealso {graphics}} The scatterplot matrix does not exhaust differences or relationality in the  `prostate` dataset, but highlights a tendency to approach it from different angles (12 times in the _Elements of Statistical Learning_) in order to map the multiple relations or influences that remain opaque to even the most exhaustive matrices of plots. The scatterplot matrix shows pairs of variables in relation. If the crucial diagnostic factor in this case is an elevated level of the PSA (prostate specific antigen), how do we know what combinations of other measurements might be associated with its elevation? What if multiple variables affect the level of PSA?[^2.31] This question can just about be pursued by scanning the matrix of plots,  but not very stably since different data analysts might see different associations combining with each other there. Different statements or  epistopics could be supported by the same figure. \index{statements!epistopic elements of}

The very question of relation between multiple variables and the predicted levels of PSA suggests the existence of a hidden, occluded or internal space that cannot be seen in a data table, and that cannot be brought to light even in the more complex geometry of a plot. This volume contains the locus of multiple relations, a locus inhering in a higher dimensional space, in this case, the nine dimensional space subtended by treating each of the nine variables or columns in the `prostate` dataset as occupying its own dimension. A different basis of order -- the vector space -- \index{vector space|(}  begins to take shape when dataset variables (usually columns in a table) become dimensions.[^2.717]

[^2.717]: Every distinct column  in a table practically adds a new dimension to the vector space. Since the 1950s, problems of classification and prediction in high-dimensional spaces have been the object of mathematical interest. The mathematician Richard Bellman coined the term ‘the curse of dimensionality’ to describe how partitioning becomes more unstable  as the dimensions of the space increase  [@Bellman_1961]. \index{Bellman, Richard!curse of dimensionality} The problem is that while the volume of a space increases exponentially with dimensions,  the number of data points (actual measurements or observations) usually does not usually increase at the same rate. In high dimensional spaces, the data becomes more thinly spread out. It is hard to partitions sparsely populated spaces because they accommodate many different boundaries. \index{vector space!dimensionality!curse of}

# Vector space expansion

To show how this space opens up, we might follow what happens to just one or two columns of the `prostate` data in the vector space as it is vectorized. \index{vectorization!of data} In the `prostate` dataset, some variables are continuous quantitative values, some are categorical (they represent membership in a group or category) and some are ordinal variables (they represent a ranking or order). \index{data!type!continuous} \index{data!type!ordinal} \index{data!type!categorical}  How can different data types be located in vector space? In order to put classifications or categories into vector space, they need to be translated into the same _basis_ as the quantitative variables with their rather more obvious geometrical and linear coordinate values. \index{vector space!basis of} How does one geometrically or indeed algebraically render a category or a qualitative difference?  The problem is solved via an expansion of the vector space through a form of binary coding that generates a new variable and hence a new dimension for each category:

>Qualitative variables are typically represented numerically by codes. The easiest case is when there are only two classes or categories, such as “success” or “failure,” “survived” or “died.” These are often represented by a single binary digit or bit as 0 or 1, or else by 1 and 1 ... When there are more than two categories, several alternatives are available. The most useful and commonly used coding is via dummy variables. Here a K-level qualitative variable is represented by a vector of K binary variables or bits, only one of which is “on” at a time [@Hastie_2009, 12] \index{data!type!categorical!coding of}

Again, the details are not so important here as the transformations that the vector space accommodates. A single qualitative or categorical variable expands into 'a vector of K binary variables or bits.' Qualitative data, once coded in this way, can be multiplied, added, and in short, handled algebraically using the same aggregate operations applied to numerical or continuous variables. Not only has the vector space expanded here, its expansion smooths over important fault-lines of difference that vertically divided the tabular data. Complex natures become simple natures. The different kinds of variables -- qualitative and quantitative, discrete and continuous, nominal and ordinal -- can be accommodated by adding dimensions to the vector space. \index{vector space|)} As Whitehead says, 'all things are vectors' [@Whitehead_1960, 309] 

Adding dimensions to vector space subsumes differences, but makes seeing the geometrically regular loci -- lines, planes, smooth surfaces -- in data distributed in this space more challenging.   The many transformations in `prostate` that ensue in _Elements of Statistical Learning_ become the locus of machine learning. \index{machine learning!epistopic}  In a historically significant transfiguration of the table, these expansions -- and we will see others, including _de novo_ creations of constructed dimensions --  subtend differences in a vector space comprising elements defined purely by coordinate position and vectoral (having direction and extent) movement.   \index{vector space!dimension}  Once this hidden, expandable and transformable (by rotation, displacement, or scaling) distribution of elements in space exists,  strenuous efforts will be made to bring loci to light.  Machine learners search for these loci or or feel for \gls{data strain}, to use Whitehead's term, \index{data!strain} along different lines. Sometimes a machine learner prehends vector space as filled with constantly varying proximities. It gathers and orders these proximities (for instance, as in the _k_ nearest neighbours model \index{machine learner!\textit{k}-nearest neighbours model}) or in unsupervised methods such as \textit{k}-means clustering [@Hastie_2009, 513]\index{machine learner!\textit{k}-means clustering}. 

More commonly, machine learning draws lines or flat surfaces that constrain the volume. \index{vector space!strain} \index{machine learner!linear regression model} The importance of lines and flat surfaces can hardly be under-estimated in machine learning. Finding lines of best fit underpins many of the machine learners that attract more  attention (neural nets, support vector machines, random forests). Linear regression with its pursuit of the straight line or plane projects the basic alignments of vector space.   It renders all differences as distances and directions of movement. Drawing lines or flat surfaces at various angles and directions is perhaps the main way in which the volume of data is traversed, and a relation between input and output, between predictors and prediction, consolidated as a loci or data strain loci or data strain.[^2.351]  The line of best fit has a ready generalization to higher dimensions, and a line can be diagrammed in the equations of linear algebra, the field of mathematics that operates on lines in spaces of arbitrary dimensions. \index{linear algebra|(}  Linear algebra operations exist for finding intersections between lines and planes, for manipulating collections of elements and aggregate forms such matrices \index{data!matrix as} through mappings and transformations (rotations, displacements or translations, skewing, and scaling), and above all, handling whole vector spaces as operational sets. It brings with it a set of formalisations -- vector space, dimension, matrix, determinant, coordinate system, linear independence, eigenvectors and eigenvalues, inner-product space, etc. -- that machine learners constantly and implicitly resort to invoke.[^2.36] \index{machine learning!reliance on linear algebra}

Many of these operations quickly become difficult to geometrically figure. [^2.41b] Let us return to the equations for linear regression models (remembering that both C.S. Pierce and Andrew Ng advocate returning often to equations). The 'mainstay of statistics,' the linear regression model, usually appears diagrammatically in a more or less algebraic form: 

\begin {equation}
\label {eq:linear_model}
\hat{Y}=\hat{\beta_0}  + \sum^p_{j=1}X_j\hat{\beta_j}
\end {equation}


\begin {equation}
\label {eq:linear_model_vector}
\hat{Y} = X_T\hat{\beta}
\end {equation}

Equations \ref{eq:linear_model} and \ref{eq:linear_model_vector} express a plane (or hyperplane) in increasingly diagrammatic abstraction.  The possibility of diagramming a high dimensional space derives largely from linear algebra. Reading Equation \ref{eq:linear_model} from left to right, the expression $\hat{Y}$ already points to a set of calculated, predicted values, or a vector of $y$ values, such as all the `lpsa` or PSA readings included in the `prostate` dataset. Similarly, the term $X_j$ points to the table of all the other variables in the `prostate` dataset. \index{dataset!prostate} Since there are 8 other variables, and close to 100 rows, $X$ is a *matrix*  -- a higher dimensional table -- of values, addressable by coordinates. \index{mathematics!linear algebra!matrix}  Finally $\beta_j$ are the pivotal coefficients or multiplying quantities that determine the slope or direction of the lines drawn. \index{deep learning} The second expression Equation  \ref{eq:linear_model_vector} relies more fully on linear algebra. This is the linear model written in 'vector form' [@Hastie_2009, 11], or vectorized. The right hand side comprises two operations $X^T$, the transpose or rotation of the data, and implicitly -- multiplication is hardly ever shown, but diagrammed by putting terms alongside each other -- an _inner product_ of the $X$ matrix and the $\beta$ parameters (to use model talk) or coefficients (to use linear algebra talk). \index{parameters|seealso{model!coefficient}}[^2.33] \index{mathematics!linear algebra!dot or inner product}

The vector form in equation \ref{eq:linear_model} diagrams an inclined  plane that cannot be fully drawn in any figure, only projected perspectivally onto the surface of a graphic plot. While that line can never fully come to light, the diagrammatics of equations \ref{eq:linear_model} and \ref{eq:linear_model_vector}  express a way of constructing it and orienting it in vector space.  Such expressions are epistopic in that they connect the local complex of activities indexed as tabulated data together through the diagonal diagrammatic element of a line or plane angling through vector space. 

# Drawing lines in a common space of transformation

\begin{figure}
  \centering
      \includegraphics[width=0.5\textwidth]{figure/vector_space.png}
        \caption{Vector space comprises transformations}
  \label{fig:common_space}
\end{figure}

Once data is distributed in vector space, machine learners operate as transformations of that space into other vector spaces, the flat loci. Indeed from the perspective of vector space, machine learners are simple transformations or operations that map vector spaces into different ones, usually of lower but sometimes of higher dimensionality.   \index{machine learning!as transformation of vector space}  For instance,  'drawing' the line of best fit through the `prostate` data or 'fitting a line' can be understood as a purely algebraic operation (although in practice, most machine learners are not purely algebraic -- they optimise and probabilise, as we will see).  \index{linear algebra|)} Viewed in terms of linear algebra, the analytical or 'closed form solution' for the parameters of the linear model is given in equation \ref{eq:linear_model_solution}: \index{model!fitting of} \index{machine learner!linear regression!closed form solution to}

\begin {equation}
\label {eq:linear_model_solution}
\hat{\beta} = (\mathbf{X}^T\mathbf{X})^{-1}\mathbf{X}^T\mathbf{y}
\end {equation}

In this expression, linear algebraic operations on the data shown as $\mathbf{X}$ calculate the coefficients $\hat{\beta}$ that orient a plane cutting through the vector space.[^2.38a] The derivation of the analytical 'ordinary least squares' solution \index{machine learner!linear regression!ordinary least squares} relies on some differential calculus \index{mathematics!calculus!differential} as well as a range of linear algebra operations such as matrix transpose, inner product and matrix inversion, the details of which need not trouble us here. The relevant point is that equation \ref{eq:linear_model_solution} constructs a plane  -- a new vector -- that traverses the density-shape of a dataset in its full dimensional vector space (nine dimensions in the case of `prostate`).[^2.38] 

# Implicit vectorization in code and infrastructures

Vectorised transformations of data lie are the moving substrate of machine learning as it expands, but they are largely taken for granted as given space or commonsense ground.\index{programming languages}  `R` coding practice instantiates vectorization in multiple ways, and is sometimes described as a 'vectorised programming language.' The vector space \index{vector space!vectorization} appears and operates just as directly in other programming languages designed for data practice (Octave, Matlab, Python's NumPy, or C++ Armadillo). 

In vectorised languages such as `R`, transformations of a data structure  expressed in one line of code  simultaneously affect all the elements of the data structure. As the widely used _R Cookbook_ puts it, 'many functions [in `R`] operate on entire vectors ...  and return a vector result' [@Teetor_2011, 38]. Or as _The Art of `R` Programming: A Tour of Statistical Software Design_ by Norman Matloff puts it, 'the fundamental data type in `R` is the _vector_' [@Matloff_2011, 24], and indeed in `R`, all data is vector. There are no individual data types,  only varieties of vectors in `R`.   \index{vectorization!in code}   There are many vectorised operations in the `R` core language and many to be found in packages (the popular 'plyr' package; vectorised operations can also be found in recent Python data analysis libraries such as `numpy` or `pandas` [@McKinney_2012]). \index{Python!packages!\textit{pandas}} The fact that many of these vectorised operations occur implicitly suggests how pervasive vector space has become in data practice.[^2.934]

[^2.934]: `R` sometimes presents difficulties for programmers trained to code using so-called procedural programming languages because it so thoroughly embraces the notion of the _vector_ -- and hence, regards all data as inhabiting vector space. In many mainstream programming languages, transformations of data rely on loops and array constructs in which some operation is successively repeated on each element of a data structure. \index{programming languages!vectorized} 

\begin{lstlisting}[language=R, caption={Vectorising code}, label={lst:vectorizing}]
     vector1 <- c(0,1,2,3,4,5,6,8,9,10)
     vector2 <- c(0,1,2,3,4,5,6,8,9,10)

     #procedural programming-style looped addition
     result_looped = vector()
     for (i in vector1) {
        result_looped[i] = vector1[i] + vector2[i]
     }
     result_looped

     #vectorised addition
     result_vectorised  <- vector1 + vector2
     result_vectorised
\end{lstlisting}

```{r vectorization, cache=TRUE, message=FALSE, warning=FALSE, comment=NA, size='smallsize', results='markup' } 
 
     vector1 <- c(0,1,2,3,4,5,6,8,9,10)
     vector2 <- c(0,1,2,3,4,5,6,8,9,10)

     #procedural programming-style looped addition
     result_looped = vector()
     for (i in 1:length(vector1)) {
        result_looped[i] = vector1[i] + vector2[i]
     }
     result_looped

     #vectorised addition
     result_vectorised  <- vector1 + vector2
     result_vectorised
```

The practical difference between the two approaches to moving through data is illustrated in the code listing \ref{lst:vectorizing} in which two 'vectors' of numbers are added together, in the first case using a classic `for`-loop construct, and in the second case using an implicitly vectorised arithmetic operation `+`. The difference between adding $1 ... 10$ using a loop or vector arithmetic is completely trivial here, or as we will soon see, nested operations are involved,   these differences in coding significantly affect human-machine relations. \index{code!as human-machine relation}  This simultaneity is only apparent, since somehow the underlying code has to deal with all the individual elements, but vectorised programming languages take advantage of hardware optimisations or carefully-crafted low-level linear algebra libraries.[^2.41] More importantly, this is a different mode of movement. Operations now longer step through a series of coordinates that address data elements, but wield planes, diagonals, cross-sections and whole-space transformations. Vectorized code reduces both data and computational frictions. The real stake in vectorizing data is not speed but transformation. It  makes working with data less like iteration through data structures (lists, indexes, arrays, fields, dictionaries, variables),   and more like folding a pliable material. Such practical shifts in feeling for data are mundane yet crucial to the epistopic movements in data. \index{vector space!vectorization!function}  [^2.455]

[^2.455]:A further level of vectorization appears in specific `R` constructs such as `apply`, `sapply`, `tapply`, `lapply`, and `mapply`. \index{programming languages!R!vector operations} All of the `-ply` constructs have a common feature: they take some collection of things (it may be ordered in many different ways – as a list, as a table, as an array, etc.), do something to it, and return a collection. While most programming languages in common use offer constructs to help deal with collections of things sequentially (for instance, by accessing each element of a list in turn and doing something with it), `R` offers ways of expressing a simultaneous operation  on  them all. The `-ply` constructs ultimately derive from the functional logic developed by the mathematician Alonzo Church in the 1930s [@Church_1936;@Church_1996]. \index{Church, Alonzo!on functional logic}  The functional programming style of applying functions to functions seems strangely abstract. \index{code!functional programming}

Vectorization also motivates  increasingly parallel contemporary chip architectures, clusters of computers such as `hadoop` or `spark`, reallocation of computation to GPUs (Graphic Processing Unit), \index{hardware!GPU} data-centre usage of FPGAs (Field Programmable Gate Arrays) \index{hardware!FPGA} and various other Cyclopean infrastructures of cloud computing.\index{vectorization!infrastructural}  \index{infrastructure!cloud computing} Many of these condensing and expanding movements of data are diagrammed in miniature in the `R` constructs as operators in vector space. 

# Lines traversing behind the light 

How does the combination of algebraic vector space and vectorised code play out in data? 'We fit a linear model' write Hastie and co-authors, referring to one epistopic operation on `prostate` data in _Elements of Statistical Learning_.  In `R` this might look like the code excerpt shown below:

\begin{lstlisting}[language=R, caption={Building a \texttt{prostate} model}, label={lst:prostate_lm}]
library(ElemStatLearn)
data(prostate)
columns_to_standardize = c(1,2,3,4,6,8,9)
prostate_standard = as.matrix(prostate[, columns_to_standardize]) 
prostate_standard = as.data.frame(scale(prostate_standard))
prostate_standard = cbind(prostate_standard, gleason=prostate$gleason, svi = prostate$svi, train = prostate$train)
train = prostate$train ==TRUE
prostate_model = lm(lpsa~., prostate_standard[train,-10])
\end{lstlisting}

```{r prostate_model}
library(ElemStatLearn)
data(prostate)
columns_to_standardize = c(1,2,3,4,6,8,9)
prostate_standard = as.matrix(prostate[, columns_to_standardize]) 
prostate_standard = as.data.frame(scale(prostate_standard))
prostate_standard = cbind(prostate_standard, gleason=prostate$gleason, svi = prostate$svi, train = prostate$train)
train = prostate$train ==TRUE
prostate_model = lm(lpsa~., prostate_standard[train,-10])

```

```{r prostate_model_output, echo=FALSE, results = 'asis'}
tab1 = xtable(summary(prostate_model), caption = 'Fitting a linear model to the \texttt{prostate} dataset', label='tab:prostate_linear_regression', type='latex')
print(tab1, include.rownames = FALSE)
```

Table \ref{tab:prostate_linear_regression} displays estimates of the coefficients or parameters \gls{$\beta}$}  that define the direction of a flat surface running through the vector space of the `prostate` data.[^2.101] \index{dataset!prostate} This new vector is a product of operations in the vectorized `prostate` data. Some vectorizing operations can be seen in `R` code in listing \ref{lst:prostate_lm} (for instance, `as.matrix` or `scale(prostate_standard)`). 

The 'unique solution' to the problem of fitting a linear model to a given dataset using the popular method of 'least squares' [@Hastie_2009, 12] is given by the operations we have seen in equation \ref{eq:linear_model_solution}. This tightly coiled expression calculates the $\hat{\beta}$ parameters that set the slope and location of a flat surface or plane in nine-dimensional vector space using all of the `prostate` variables apart from one variable chosen as the response or predicted variable, in this case `lpsa`. $X$ and $y$  matrices are multiplied, transposed (a form of rotation that swaps rows for columns) and inverted (a more complex operation that finds another matrix) in a series of linear algebra transformations. Epitomising the implicitly vectorised code often seen in machine learning, calculating $\hat{\beta}$ for  the `prostate` data only requires one line of `R` code: \index{code!implicit vectorization of}

\begin{lstlisting}[language=R, caption={Closed form evaluation of linear model parameters}, label={lst:beta_hat}]
beta_hat = ginv(t(X) %*% X) %*% t(X) %*% y
\end{lstlisting}


The implicit vectorization of the `R` code in the code listing \ref{lst:beta_hat}, the fact that it already concretely operates in the vector space, operationalizes the concise diagrammaticism of equation \ref{eq:linear_model_solution} as a machine process. \index{vector space!vectorization} More importantly, the vectorised multiplication, transposition and inversion of data creates the new vector $\hat{\beta}$ whose variations can be explored, observed, graphed and varied in ways that go well beyond the statistical tests of significance, variation, and error reported in Table \ref{tab:prostate_linear_regression}. \index{statistics!tests of significance}  (We will have occasion to return to these statistical estimates in chapter \ref{ch:probability}.) The play of values that starts to appear even in fitting one linear model will become much more significant when fitting hundreds or thousands of models, as some machine learners do.[^2.300] \index{parameters!estimation of}

# The vectorised table?

I started out from the observation that _Elements of Statistical Learning_ mixes many datasets. \index{machine learning!many datasets in} The more abstract implications of vectorization and the forms of transformation movement it encourages and  proliferates bring us back to the problem of how machine learning mixes datasets that span different settings.  In short, vectorising computation makes the vector space, which we might understand as a resurgent form of the pre-Classical table, a table that tolerates all many of relations and similarities,  operationally concrete and machinically abstract. It is no longer a visible diagram, but a machinic process that multiplies and propagates into the world along many diagonal lines.  

Machine learning relies on a broad but subtle transformation of data into vectors and a vector space. Slightly re-purposing Foucault's archaeology of tables in _The Order of Things_, I have suggested that vectorization remaps the grid of the table into the expanding dimensions of the vector space. This space accommodates both simple and complex natures. This is not the first such expansion of the table. We need only think of the relational database systems of the late 1960s, and their multiplication of tables [@Mackenzie_2012]. But in the vectorised and matrix-form practices of the  vector space, machine learning produces for the first time a meta(s)table volume that cannot be surfaced on a page or screen.  \index{table} \index{vector space!metatable} 

Does the vectorization of data lie a 'a long way from sense data' as Arendt suggest? In the diagrammatic operations of linear algebra on data, and the vectorization of code,  machine learning traverses dimensions that, as Arendt observes, cannot be immediately sensed. Whitehead's notion of data strain \index{data!strain}  as 'a complex distribution of geometrical significance' suggests, however,  that vectorization is not a complete loss of feeling.   Every   machine learner inhabits and moves through the vector space along different strains. Sometimes their operations flatten the vector space down into lower dimensional sub-spaces  as in the many 'dimensional reduction' machine learners such as principal component analysis, Latent Dirichlet Allocation or indeed the linear regression model that maps an irregular volume onto a plane. \index{machine learner!principal component analysis} \index{machine learner!Latent Dirichlet Allocation} Sometimes they expand the vector space into a great many new dimensions or 'features' \index{vector space!features in}  (as we saw with 'dummy variables' that embody categories, and as we will see with support vector machine classifiers in  chapter \ref{ch:pattern} or the deep learners of chapter \ref{ch:subjects}). \index{machine learner!support vector machine}

The epistopic transformation of datasets and tables into vector space reaches into and  re-aligns communication and infrastructures. It acts as a powerful tensor on knowledges and operations of many different kinds as it transposes, inverts, and re-maps local complexes of activity. In following what happens to vectors, lists, matrices, arrays, dictionaries, sets, dataframes, series or tuples in data, we might get a sense of how the epistemic operations of predictive models, the supervised and unsupervised learners, the classifiers, the decision trees and the neural networks have purchase in data.

What is at stake in vectorizing data? \index{vectorization}  It produces a common space that juxtaposes and mixes localized complex realities. The `prostate` dataset could be aggregated and melded as vectors with a microarray, heart disease or bone density datasets. In vector space, identities and differences change in nature. \index{differences!vector space} Similarity and belonging no longer rely on resemblance or a common genesis, but on measures of proximity or distance, on flat loci that run as vectors through the space. Vectorization, the deep saturation of the table  by algebra, constitutes all relations as movements of transformation, diagonalization, inversion or rotation. The epistopic power of vectorization takes root in the elementary practices of machine learning, and engenders many variations amongst machine learners. Vectorisation also strains the production of knowledge through a loss of the visible geometry of tabular comparison. This loss of visibility is, as we will see met by the production of new groups of statements,  new visible forms and operational devices and infrastructures that accommodate the dimensional expansion of vector space. Infrastructural vectorization has often been called 'big data.' \index{vectorization!infrastructure} \index{data!'big'|see{data!all of}} 

The fascination of machine learning, its seemingly endless applications (I refer the reader back to the diagram of machine learning's vastness in chapter \ref{ch:diagram}), owes much to the vector feeling, with its twin lures of ideal operationality -- everything is a vector operation -- and its tantalizing tendency to expand and to move. This feeling, the vector feeling, we might note, is not surprising. ' characteristically for Whitehead 'Feelings are "vectors"; for they feel what is there and transform it into what is here' [@Whitehead_1960, 87]. \index{Whitehead, Alfred North!feeling!as vector} \index{vector!as feeling}

Expansive data vectorization challenges contemporary critical to develop intuitions and value-relevant concepts describing vector-feelings or data strains.  We lack good intuitions of how to do that partly because machine learning implicitly vectorizes its practice in code, in infrastructures and in highly condensed diagrammatic forms. My aim in undertaking an archaeology of the transformations of tables into vector spaces is to unwind or de-diagonalise some of the operations rippling through different treatments of data. \index{archaeology!of transformation}  The act of diagramming how machine learners  vectorise data densities  begins to locate and unravel the processes of knowing, predicting and deciding on which many aspects of the turn to data rely.  The vectoral operations we have just been viewing are themselves organised and aligned by other lines of diagrammatic movement \index{diagrammatic movement} that shape surfaces in more convoluted forms.    



[^2.38]: As we will see in the following chapter (chapter \ref{ch:function}), it is not always possible to calculate the parameters of a model analytically. Especially in relation to contemporary datasets that have very many variables and many instances (rows in the table), linear algebra approaches become unwieldy in their attempt to produce exact results, and machine learning steps in with a variety of computational optimisation techniques. \index{coefficients}

[^2.38a]: Perhaps more importantly, the linear algebraic expression of these operations presupposes that all the data, both the values used to build the model and the predicted values the model may generate as it is refined or put into operation somewhere, are contained in a common space, the vector space, a space whose formation and transformation can be progressively ramified and reiterated by various lines that either separate volumes in the space, or head in a direction that brings along most of the data. Not all of these lines are bound to be straight, and much of the variety and dispersion visible in machine learning techniques comes from efforts to construct different kinds of lines or different kinds of 'decision boundaries' (in the case of classification problems) in vector space (for instance, the k-nearest neighbors method does not construct straight lines, but somewhat meandering curves that weave between nearby vectors in the vector space; see [@Hastie_2009, 14-16]). \index{machine learner!\textit{k}-nearest neighbours}  Whether they are straight or not, the epistopic aspect of these lines remains prominent. Typically, many different statistical tests (Z-scores or standard errors, F-tests, confidence intervals, and then prediction errors) will be applied to any estimate of the coefficients of even the basic linear regression model, well before most advanced or sophisticated models and techniques (cross-validation, bootstrap testing, subset and shrinkage selection) begin to re-configure the model in more radical ways. \index{statistics!tests}


[^2.10]: Some genomic data will be the focus of a later chapter \ref{ch:genome}. Machine learning during the 1990s and 2000s was in some ways boosted heavily by the advent of genomic biology with its large, enterprise style knowledge endeavours such as the Human Genome Project.

[^2.11]: The data derives from a publication in _nature genetics_ [@Ross_1916_2000] analysing gene expression in the cancer cell lines maintained as experimental models by the us national cancer institute.

[^2.41]: Learning machine learning, and learning to implement machine learning techniques, is largely a matter of implementing series of matrix multiplications. As Andrew Ng advises his students,

    > Almost any programming language you use will have great linear algebra libraries. And they will be high optimised to do that matrix-matrix multiplication very efficiently including taking advantage of any parallelism your computer. So that you can very efficiently make lots of predictions of lots of hypotheses [@Ng_2008a, 10:50]\index{Ng, Andrew}

    In other parts of his teaching, and indeed throughout the practice exercises and assignments, Ng stresses the value of implementing machine learning techniques for both understanding them and using them properly. But this is one case where implementation does not facilitate learning. Ng advises his learners against implementing their own matrix handling code.    They should instead  use the 'great linear algebra libraries' found in 'almost any programming language.' Linear algebra libraries multiply, transpose, decompose, invert and generally transform matrices and vectors. \index{linear algebra} They will be 'highly optimised' not because every programming language has been prepared for the advent of machine learning on a large scale, but rather more likely because matrix operations are just so widely used in image and audio processing.  Happily, Ng observes, that means that 'you can make lots of predictions' [@Ng_2008e].   It seems that generating predictions and hypotheses outweighs the value of understanding how things work on this point. \index{prediction}


[^2.351]: One sign of the centrality of the line in machine learning can be seen, for instance, from the contents page of the book [@Hastie_2009, xiii-xxii]. After the introduction of the linear model in the first chapter and its initial exposition in chapter 2 ('overview of supervised learning'), it forms the central topics of chapter 3 ('linear methods for regression'), chapter 4 ('linear methods for classification'), chapter 5 ('basis expansions and regularization'), chapter 6 ('kernel smoothing methods'), much of chapter 7 ('model assessment and selection'), chapter 8 ('model inference and averaging'), major parts of chapter 9 ('additive models, trees and related methods'), important parts of chapter 11 ('neural networks' -- neural networks can be understood as a kind of regression model), the anchoring point of chapter 12 ('support vector machines and flexible discriminants') and the main focus in the final chapter ('high dimensional problems'). A similar topic distribution can be found in Andrew Ng's Cs229 lectures on machine learning. More than half of the 20 lectures concern linear models and their variants. See [@Ng_2008a; @Ng_2008b;@Ng_2008c;@Ng_2008d].


[^2.36]: Along with statistics and probability, linear algebra is a such an important part of machine learning that many books and courses recommend students complete a linear algebra course before they study machine learning. Cathy O'Neill and Rachel Schutt advise:
    When you’re developing your skill set as a data scientist, certain foundational pieces need to be in place first—statistics, linear algebra, some programming [@Schutt_2013, 17]

[^2.31]: Despite the intensive work that Hastie and co-authors conduct on the `prostate` data, all with a view to better predicting PSA levels using volumes and weights of prostates, etc., Stamey and other urologists more than a decade or so concluded that PSA is not a good biomarker for prostate cancer. \index{dataset!prostate} Stamey writes in 2004:

    What is urgently needed is a serum marker for prostate cancer that is truly proportional to the volume and grade of this ubiquitous cancer, and solid observations on who should and should not be treated which will surely require randomized trials once such a marker is available. Since there is no such marker for any other organ confined cancer, little is likely to change the current state of overdiagnosis (and over-treatment) of prostate cancer, a cancer we all get if we live long enough. [@Stamey_2004, 1301] 

[^2.33]: Carl Friedrich Gauss \index{Gauss, Carl Friedrich} and Adrien-Marie Legendre's work on linear regression at this time is well-known. The first independent use of linear regression was Gauss' prediction of the location of an 'occluded volume,' the position of the asteroid Ceres after it reappeared in its orbit behind the sun. [@Stigler_2002] -- TBA page ref


[^2.1]:  Later chapters will discuss various ways in which the vectoral dimensionality of data or its rendering as _vector space_ scales up and scales down in machine learning. In terms of multiplying matrices, dimensionality both constrains and enables many aspects of the prediction. Perhaps on the grounds of data dimensionality alone, we should attend to dimensionality practices in  `R`. A fuller discussion of dimensionality of data is the topic of chapter \ref{ch:pattern}. There I discuss how machine learning re-dimensions data in various ways, sometimes reducing dimensions and at other times, multiplying dimensions.

[^2.22]: By and large, I am not discussing networking and database infrastructures here. In other work, I have attempted to account for the multiplication of tables in databases but also in mundane practices of ordering. See [@Mackenzie_2011; @Mackenzie_2012]. My analysis there pivots mainly on a set-driven account of data derived from the work of Alain Badiou. \index{Badiou, Alain} Although I've already been using some set terminology here -- as in 'dataset' -- I'm persisting with a more geometrical-algebraic account of datasets in order to better deal with some of the spatial operations common in machine learning. \index{Badiou, Alain!set-based account of data}

[^2.41b]: We might add also approach the epistopic fault line in machine learning topologically \index{topology}. Over a decade ago, the cultural theorist Brian Massumi wrote that 'the space of experience is really, literally, physically a topological hyperspace of transformation' [@Massumi_2002, 184] \index{Massumi, Brian}. Much earlier, Gilles Deleuze had conceptualised Michel Foucault's philosophy as a topology, or 'thought of the outside' [@Deleuze_1988], as a set of movements that sought to map the diagrams that generated a 'kind of reality, a new model of truth' [@Deleuze_1988, 35]. More recently, this topological thinking has been extended and developed by Celia Lury amongst others. In 'The Becoming Topological of Culture,' Lury, Luciana Parisi and Tiziana Terranova suggests that 'a  new rationality is emerging: the moving ratio of a topological culture' [@Lury_2012, 4] \index{Lury, Celia}. \index{Parisi, Luciana} \index{Terranova, Tiziana} In this new rationality, practices of ordering, modelling, networking and mapping co-constitute culture, technology and science [@Lury_2012, 5].  At the core of this new rationality, however, lies a new ordering of continuity. The 'ordering of continuity,' Lury, Parisi and Terranova propose, takes shape 'in practices of sorting, naming, numbering, comparing, listing, and calculating' (4). The phrase 'ordering of continuity' is interesting, since we don't normally think of continuities as subject to ordering. In many ways, that which is continuous bears within it its own ordering, its own immanent seriation or lamination. But in the becoming topological of culture, movement itself undergoes a transformation according to these authors. Rather than movement as something moving from place to place relatively unchanged (as in geometrical translation), movement should be understood as more like an animation, a set of shape-changing operations. These transformations, I would suggest, should be legible in the way that machine learning, almost the epitome of the processes of modelling and calculation that Lury, Parisi and Terranova point to, itself moves through the data. And indeed, the juxtaposition of spam, biomedical data, gene expression data and handwritten digits already suggests that topological equivalences, and a 'radical expansion' of comparison might be occurring. Bringing epistopics and topologies together might, I suggest, help trace, map and importantly diagram some of the movements into the data occurring today.

[^2.101]: From the epistopic viewpoint, the most obvious result of fitting a linear model is the production not of a line on a diagram or in a graphic. As we have seen, such lines cannot be easily rendered visible. Instead, the model generates a new column-vector  of coefficients (see Table \ref{tab:prostate_linear_regression}) and some new numbers, _statistics_.  This table is not as extensive as the original data, the $\mathbf{X}$ and $Y$ vectors. But the names of the variables in the dataset appear as rows in the new table, a table that describes something of how a line has been fitted by the linear model to the data. The columns of the table now bear abbreviated and much more statistical names such as `estimate` (the estimated values of $\hat{\beta}$, the key parameters in any linear model), `Std. Error`, `t value`, and the all important _p_ values written as `Pr(|t|)`.  The numerical values ranging along the rows mostly range from -1 to 1, but the final column includes values that are incredibly small: `1.47e-06` is a few millionths. Other statistics range around the outside the table: the `F-statistic`, the `R-squared` statistic, and the `Residual standard error`. The numbers of the table \ref{tab:prostate} become epistopic here, since they now appear as a set of standard errors, estimates, t-statistics, and _p_ values, that together indicate how likely the estimated values of $\beta$ are, and therefore how well the diagonal line expresses the relations between different dimensions of the dataset in the vector space. \index{statistics!tests of significance!in linear regression}


[^2.300]: This is an important differentiation: it is not typical machine learning practice to construct one model, characterised by a single set of statistics (F scores, R^2 scores, _t_ values, etc.). In practice, most machine learning techniques construct many models, and the efficacy of some predictive techniques derives often from the multiplication or indeed proliferation of models. Techniques such as neural networks, cross-validation, bagging, shrinkage and subset selection, and random forests, to name a few, generate many statistics, and navigating the multiple or highly variable models that result becomes a major concern. An epistopic abundance will appear here -- bias, variance, precision, recall, training error, test error, expectation, Bayesian Information Criteria, etc. as well as graphisms such as ROC (Receiver-Operator-Characteristics) curves. Put simply, the proliferation of models start to drive the dimensional expansion of the vector space. At the same time, the multiplicity of models multiplied by the machine learners  becomes the topic of statistical analysis. 


[^2.104]:   I loosely borrow the term 'density' from statistics, where *probability density functions* are often used to describe the hardly ever uniform distribution of probabilities of different values of a variable. Sensing density as a form of variation matters greatly both in machine learning itself, where algorithms seek purchase on unevenly distributed data, and in any broader diagram of data. Probability densities are discussed in much more detail in Chapters \ref{ch:probability}. The collection _"Raw Data" is an Oxymoron_ [@Gitelman_2013] evinces some of these different densities and distributions of data.  

